TODO

1. The Order has no constitution except its members.

As it was said: \emph{Others he saved, himself he cannot save.}

2. It recommends nevertheless that its members shall make a formal act of union with it and of recognition of their own nature.

As it was said: \emph{Am I my brothers keeper?}

3. Its concern is the practice of the apprehension of the Co-inherence both as a natural and a supernatural principle.

As it was said: \emph{Let us make man in our image.}

4. It is therefore, \emmph{per necessitatem}, Christian.

As it was said: \emph{And who ever says that there was when this was not, let him be anathema.}

5.. It recommends therefor the study, on the contemplative side, of the Co-inherence of the Holy and Blessed Trinity, of the Two natures in the single person, of the Mother and Son, of the communicated Eucharist, and of the whole catholic Church.

As it was said: \emph{figlia et tuo figlio.}

And on the the active side, of methods of exchange, in the Sate, in all forms of love, and in all natural things, such as child-birth.

As it was said: \emph{Bear ye one another's burdens.}

6. It concludes in the Divine Substitution of Messias* all forms of exchange and substitution, and it invokes this Act as the root of all.

As it was said: \emph{He must become, as it were, a double man.}

7. The Order will associate itself primarily with four feasts: the Feast of the Annunciation, the Feast of the Blessed Trinity, the Feast of the Transfiguration, and the Commemoration of All Souls.

As it was said: \emph{Another will be in me and I in him.}





Appendix1
Blessing of anything
Blessing of Holy Water

Appendix 2

Personal prayers (1928 bcp)


Grace before Meat.

\drop{Bless, O Father, thy gifts to our use and us to thy service; for Christ’s sake. \R Amen.}

\drop{Give us grateful hearts, our Father, for all thy mercies, and make us mindful of the needs of others; through Jesus Christ our Lord. \R Amen.}
 

The Lord's Prayer 7+1 times a day.


Itinerary
before bathing (asperges, baptism?)
before washing (wash before praying, and eating/drinking)
Psalm 26: 6-12 in the Hebrew
\drop{I will wash my hands in innocency, O {\scshape Lord}; * and so will I go to thine altar.}

7 That I may shew the voice of thanksgiving, * and tell of all thy wondrous works.

8 {\scshape Lord}, I have loved the habitation of thy house, * and the place where thine honour dwelleth.

9 O shut not up my soul with the sinners, * nor my life with the blood-thirsty.

10 In whose hands is wickedness, * and their right hand is full of gifts.

11 But as for me, I will walk innocently: * O deliver me, and be merciful unto me.

12 My foot standeth right: * I will praise the {\scshape Lord} in the congregations.

before praying
Lucernaria/exorcism-blessing of flame

before mass
\drop{Come, Holy Ghost, our souls inspire,}\\
And lighten with celestial fire.\\
Thou the anointing Spirit art,\\
Who dost thy sevenfold gifts impart.

Thy blessed unction from above,\\
Is comfort, life, and fire of love.\\
Enable with perpetual light\\
The dulness of our blinded sight.

Anoint and cheer our soiled face\\
With the abundance of thy grace.\\
Keep far our foes, give peace at home;\\
Where thou art guide, no ill can come.

Teach us to know the Father, Son,\\
And thee, of both, to be but One;\\
That, through the ages all along,\\
This may be our endless song:

Praise to thy eternal merit,\\
Father, Son, and Holy Spirit.



130 for the dead.


Thanksgivings
after meals
after a journey
after mass

Appendix 3
Order of the coinherence

Fix Hierusalem -> Jerusalem; Iacob-Jacob. There being no textual evidence of the Formers I can find.


Ant. What ǐs this * that he sâith, ȧ lǐttlê while? allėlûya: wé cǎnnȯt têll whȧt hě sâith, állêluya.



Chapter: Big all-caps
Section: Regular all caps (Te deum, venite)
Subheading: Regular (Scriptural citation)
subsubsection (italic)


Chapter:   
The order for (caps, 1em?)
MORNING PRAYER (caps, 2em?)
daily throughout the year (caps, 1em?)

Section
;
a supplication
;
The First Sunday in Advent
;
The Nativity of Our Lord, or the 1
Birthday of Christ 1 
commonly called 2
Christmas day 1
[December 25] 2
;
The Sunday called  (it) 3
Septuagesima (all caps)1
or the third Sunday before Lent. 3
;
Day 9. Evening Prayer

subsection
VENITE< EXULTEMUS DOMINO (caps, 1em?) 1
Psalm 95. (.75em?) 2

;
Psalm 47 (ALL CAPS); Omnes gentes, plaudite. (it)


Anglican Breviary
The Song of Zacharias
_Benedictus Dominus, Deus Israel._
Luke I.68.

Diurnal
The Song of the Blessed Virgin (SC)
_Magnificat_

subsubsection
2 Corinthians 13. (it, 1em)
; 
The Collect
;
The Epistle. (it) Romans 13.8 (normal)

Baskerville:



_Te Deum Laduamus_
_Beedictus_ S. LUKE I. 68
_Jubilate DEO. PSAL. 100.

_the Second Collect, for Peace_


_The_ Nativity _of our Lord, or the_ Birth-Day _of_ CHRIST, _commonl called_ Christmas-day.



Psalms:
_The_ 1. _Day.     Psalms   |     Psalms  _The_ 2. _Day._
PSAL. 8. _Domine, Dominus noster

M O R N I N G  P R A Y E R
PSAL. 9. _Confitebor tibi.

Current:
Day 12 : Mn.   THE PSALMS   Ps. 64 | Ps. 65  THE PSALMS  Day 12 : Ev.

