
\chapter[The Order of Baptism]{\stylechapter{The Ministration of}{Holy Baptism}{To be Used in the Church }}

% 1549
\pilcrow{It appeareth by ancient writers that the Sacrament of Bapitism in the old time was not commonly ministered but at two times in the year, at Easter and Whitsuntide, at which times it was only ministered in presence of all the congregation: which custom now being grown out of use, although it cannot for many considerations be well restored again, yet it is thought good to follow the same, as near as conveniently may be: wherefore the people are to be admonished, that it is most convenient that Baptism should not be ministered but upon Sundays and other Holy Days, when the most number of people may come together, as well for that the Congregation there present may testify the receiving of them that be newly baptized into the number of Christ’s Church, as also because in the Baptism of Infants, every man present may be put in remembrance of his own profession made to God in his Baptism. Neverthelesse (if necessity so require) children may at all times be baptized at home.}

\smallskip

\pilcrow{Due notice, normally of at least a week, shall be given before a child is brought to the church to be baptized.}

\medskip


\pilcrow{At the time appointed the godfathers and godmothers and [the parents or guardians with] the \emph{child} must be ready at the church door, either immediately before the last Canticle at Mattins or else immediately before the last Canticle at Evensong, as the Curate by his discretion shall appoint.  And the Priest standing there shall proceed as follows.}

\drop{Hath this Child [\emph{or} Person] been already baptized or no?}

\smallskip

\centerline{\pilcrow{If they answer, \emph{No}: Then shall the Priest proceed as followeth.}}



\section{Admission to the Catechuminate}
\drop{Dearly beloved, forasmuch as all men are conceived and born in sin: and that our Saviour Christ saith, None can enter into the kingdom of God, except he be regenerate and born anew of Water and of the Holy Ghost: I beseech you to call upon God the Father, through our Lord Jesus Christ, that of his bounteous mercy he will grant to \emph{these Children} [\rubric{or} \emph{these persons}] that thing which by nature \emph{they} cannot have; that \emph{they} may be baptized with Water and the Holy Ghost, and received into Christ’s holy Church, and be made \emph{lively members} of the same.}

\centerline{\pilcrow{Then shall the Priest say,}}
\centerline{Let us pray.}
\drop{Almighty and everlasting God, who of thy great mercy didst save Noah and his family in the ark from perishing by water; and also didst safely lead the children of Israel thy people through the Red Sea, figuring thereby thy holy Baptism; and by the Baptism of thy well-beloved Son Jesus Christ, in the river Jordan, didst sanctify Water to the mystical washing away of sin: We beseech thee, for thine infinite mercies, that thou wilt mercifully look upon \emph{these Children} [\rubric{or} \emph{these} thy \emph{Servants and Handmaidens}]; wash \emph{them} and sanctify \emph{them} with the Holy Ghost; that \emph{they}, being delivered from thy wrath, may be received into the ark of Christ’s Church; and being stedfast in faith, joyful through hope, and rooted in charity, may so pass the waves of this troublesome world, that finally \emph{they} may come to the land of everlasting life, there to reign with thee world without end; through Jesus Christ our Lord. \R Amen.}

\medskip

%% <1549>
\pilcrow{Here shall the priest ask what shall be the name of the \emph{child}, [and when the Godfathers and Godmothers have told the name,] then shall he make a cross upon the \emph{child’s} forehead and breast, saying,}

\lettrine{\emph{N.}}{ receive} the sign of the holy Cross, both in thy \grealtcross\ forehead, and in thy \grealtcross\ breast, in token that thou shalt not be ashamed to confess thy faith in Christ crucified, and manfully to fight under his banner, against sin, the world, and the devil; and to continue Christ’s faithful soldier and servant unto thy life’s end. \R Amen.

{\footnotesize\rubric{And this he shall do and say to as many as be presented to be baptized, one after another.}\par}

\medskip


\centerline{Let us pray.}
%% </1549>
\drop{Almighty and immortal God, the aid of all that need, the helper of all that flee to thee for succour, the life of them that believe, and the resurrection of the dead: We call upon thee for \emph{these Infants} [\rubric{or} \emph{these Persons}], that \emph{they}, coming to thy holy Baptism, may receive remission of \emph{their} sins by spiritual regeneration. Receive \emph{them}, O Lord, as thou hast promised by thy well-beloved Son, saying, Ask, and ye shall have; seek, and ye shall find; knock, and it shall be opened unto you: So give now unto us that ask; let us that seek find; open the gate unto us that knock; that \emph{these Infants} [\rubric{or} \emph{these Persons}] may enjoy the everlasting benediction of thy heavenly washing, and may come to the eternal kingdom which thou hast promised by Christ our Lord. \R Amen.}


\medskip


%% <1549>
\pilcrow{Then let the priest, looking upon the \emph{Children}, say,} 

\drop{I command thee, unclean spirit, in the name of the Father, of the Son, and of the Holy Ghost, that thou come out, and depart from \emph{these Infants} [\rubric{or} \emph{these Persons}], whom our Lord Jesus Christ hath vouchsafed to call to his holy Baptism, to be made \emph{members} of his body, and of his holy congregation. Therefore thou cursed spirit, remember thy sentence, remember thy judgment, remember the day to be at hand wherein thou shalt burn in fire everlasting, prepared for thee and thy angels. And presume not hereafter to exercise any tyranny towards \emph{these Infants} [\rubric{or} \emph{these Persons}], whom Christ hath bought with his precious blood, and by this his holy Baptism, called to be of his flock, In the name of the same our Lord Jesus Christ, who shall come to judge the quick and the dead and the world by fire. \R Amen.}
%% </1549>


\medskip


\pilcrow{Then, if all candidates be children, the Priest shall say the Gospel and Exhortation as follow,}
\V The Lord be with you.  \R And with thy spirit.

\drop{Hear the words of the \grecross\ Gospel, written by Saint Mark, in the tenth chapter, at the thirteenth verse. \R Glory be to thee, O Lord.}

\smallskip

\drop{They brought young children to Christ, that he should touch them; and his disciples rebuked those that brought them. But when Jesus saw it, he was much displeased, and said unto them, Suffer the little children to come unto me, and forbid them not; for of such is the kingdom of God. Verily I say unto you, Whosoever shall not receive the kingdom of God as a little child, he shall not enter therein. And he took them up in his arms, put his hands upon them, and blessed them.}

\centerline{\R Praise be to thee, O Christ.}

\medskip
\pilcrow{After the Gospel is read, the Minister shall make this brief Exhortation upon the words of the Gospel.}

\drop{Beloved, ye hear in this Gospel the words of our Saviour Christ, that he commanded the children to be brought unto him; how he blamed those that would have kept them from him; how he exhorteth all men to follow their innocency. Ye perceive how by his outward gesture and deed he declared his good will toward them; for he embraced them in his arms, he laid his hands upon them, and blessed them. Doubt ye not therefore, but earnestly believe, that he will likewise favourably receive \emph{these} present \emph{Infants}; that he will embrace \emph{them} with the arms of his mercy; that he will give unto \emph{them} the blessing of eternal life, and make \emph{them} partaker\emph{s} of his everlasting kingdom.

Wherefore we being thus persuaded of the good will of our heavenly Father towards \emph{these Infants}, declared by his Son Jesus Christ; and nothing doubting but that he favourably alloweth this charitable work of ours in bringing \emph{these Infants} to his holy Baptism; let us faithfully and devoutly give thanks unto him, and say the prayer which the Lord himself taught. And in declaration of our faith, let us also recite the articles contained in our Creed.} %and say,}


\begin{leftbar}
\centerline{\pilcrow{Or, for those of Riper Years,}}
\V The Lord be with you.  \R And with thy spirit.

\drop{Hear the words of the \grecross\ Gospel, written by Saint John, in the third chapter, beginning at the first verse.  \R Glory be to thee, O Lord.}

\smallskip

\drop{There was a man of the Pharisees, named Nicodemus, a ruler of the Jews. The same came to Jesus by night, and said unto him, Rabbi, we know that thou art a teacher come from God; for no man can do these miracles that thou doest, except God be with him. Jesus answered and said unto him, Verily, verily I say unto thee, except a man be born again, he cannot see the kingdom of God. Nicodemus saith unto him, How can a man be born when he is old? Can he enter the second time into his mother’s womb, and be born? Jesus answered, Verily, verily I say unto thee, except a man be born of water and of the Spirit, he cannot enter into the kingdom of God. That which is born of the flesh is flesh; and that which is born of the Spirit is spirit. Marvel not that I said unto thee, Ye must be born again. The wind bloweth where it listeth, and thou hearest the sound thereof; but canst not tell whence it cometh, and whither it goeth: so is every one that is born of the Spirit.}

\centerline{\R Praise be to thee, O Christ.}

\medskip

\pilcrow{After which he shall say this Exhortation following.}
\drop{Beloved, ye hear in this Gospel the express words of our Saviour Christ, that except a man be born of water and of the Spirit, he cannot enter into the kingdom of God. Whereby ye may perceive the great necessity of this Sacrament, where it may be had. Likewise, immediately before his ascension into heaven, (as we read in the last chapter of Saint Mark’s Gospel,) he gave command to his disciples, saying, Go ye into all the world, and preach the Gospel to every creature. He that believeth and is baptized shall be saved; but he that believeth not shall be condemned. Which also sheweth unto us the great benefit we reap thereby. For which cause Saint Peter the Apostle, when upon his first preaching of the Gospel many were pricked at the heart, and said to him and the rest of the Apostles, Men and brethren, what shall we do? replied and said unto them, Repent, and be baptized every one of you for the remission of sins, and ye shall receive the gift of the Holy Ghost. For the promise is to you and your children, and to all that are afar off, even as many as the Lord our God shall call. \footnote{\rubric{These words may be omitted.}}[And with many other words exhorted he them, saying, Save yourselves from this untoward generation. For (as the same Apostle testifieth in another place) even Baptism doth also now save us, (not the putting away of the filth of the flesh, but the answer of a good conscience towards God,) by the resurrection of Jesus Christ.] Doubt ye not therefore, but earnestly believe, that he will favourably receive these present persons, truly repenting, and coming unto him by faith; that he will grant them remission of their sins, and bestow upon them the Holy Ghost; that he will give them the blessing of eternal life, and make them partakers of his everlasting kingdom.}

Wherefore we being thus persuaded of the good will of our heavenly Father towards \emph{these persons}, declared by his Son Jesus Christ; let us faithfully and devoutly give thanks to him, and say the prayer which the Lord himself taught. And in declaration of our faith, let us also recite the articles contained in our Creed.
\end{leftbar}


\pilcrow{Here the minister with the Godfathers, Godmothers, and people present, shall say,}

\drop{Our Father, which art in heaven, Hallowed be thy Name. Thy kingdom come. Thy will be done, in earth as it is in heaven. Give us this day our daily bread. And forgive us our trespasses, As we forgive them that trespass against us. And lead us not into temptation; But deliver us from evil. Amen.}
 

% And then shall say openly.
\medskip

\drop{I believe in God the Father Almighty, Maker of heaven and earth:}

And in Jesus Christ his only Son our Lord: Who was conceived by the Holy Ghost, Born of the Virgin Mary: Suffered under Pontius Pilate, Was crucified, dead, and buried: He descended into hell; The third day he rose again from the dead: He ascended into heaven, And sitteth on the right hand of God the Father Almighty: From thence he shall come to judge the quick and the dead.

I believe in the Holy Ghost: The holy Catholick Church; The Communion of Saints: The Forgiveness of sins: The Resurrection of the body, And the Life everlasting. Amen.

\smallskip

\centerline{\rubric{The priest shall add also this prayer.}}
\drop{Almighty and everlasting God, heavenly Father, we give thee humble thanks, for that thou hast vouchsafed to call us to the knowledge of thy grace, and faith in thee: Increase this knowledge, and confirm this faith in us evermore. Give thy Holy Spirit to \emph{these Infants} [\rubric{or} \emph{these Persons}], that \emph{they} may be born again, and be made \emph{an heir} of everlasting salvation; through our Lord Jesus Christ, who liveth and reigneth with thee and the Holy Spirit, now and for ever. \R Amen.}

\medskip

\pilcrow{Then let the priest take [one of] the \emph{children} by the right hand, [the others being brought after him]. And coming into the Church say,}

\drop{The Lord vouchsafe to receive \emph{you} into his holy household, and to keep and govern \emph{you} alway in the same, that \emph{ye may} have everlasting life. Amen.}




% It consisted of three acts. First the closing exorcism — the Bishop stretching his hands over them as they knelt facing eastwards, prayed for the last time for the ejection of the evil spirit from them; secondly, the exsufflation — he breathed in their faces; thirdly, the Effeta — he touched each candidate on the mouth, ears, \etc, with spittle or oil, after the example of our Lord’s action in; healing the deaf and dumb man.9

% sarum:
% Exorcism, Gospel, Effeta, LordsPrayer/creed. Signing on right hand + blessing, introduction to church.




\section{The Promises}

\pilcrow{Here may be sung a Hymn or an Anthem, \emph{or} the Invocations and Conclusion from the Litany.}

\medskip

\pilcrow{Then, standing at the Font, shall the Priest speak unto the Godfathers and Godmothers on this wise.}

\drop{Dearly beloved, ye have brought \emph{these Children} here to be baptized, ye have prayed that our Lord Jesus Christ would vouchsafe to receive \emph{them}, to release \emph{them} of \emph{their} sins, to sanctify \emph{them} with the Holy Ghost, to give \emph{them} the kingdom of heaven, and everlasting life. Ye have heard also that our Lord Jesus Christ hath promised in his Gospel to grant all these things that ye have prayed for: which promise he, for his part, will most surely keep and perform. 

Wherefore, after this promise made by Christ, \emph{these Infants} must also faithfully, for \emph{their} part, promise by you that are \emph{their} sureties, (until \emph{they} come of age to take it upon \emph{themselves},) that \emph{they} will renounce the devil and all his works, and constantly believe God’s holy Word, and obediently keep his commandments.}


\begin{leftbar}
\centerline{\rubric{Or, for those of Riper Years,}}

\pilcrow{Then the Priest shall speak to the persons to be baptized on this wise}

\drop{Well-beloved, who are come hither desiring to receive Holy Baptism, \emph{ye have} heard how the congregation hath prayed, that our Lord Jesus Christ would vouchsafe to receive \emph{you} and bless \emph{you}, to release \emph{you} of \emph{your} sins, to give \emph{you} the kingdom of heaven, and everlasting life. \emph{Ye have} heard also, that our Lord Jesus Christ hath promised in his holy Word to grant all those things that we have prayed for; which promise he, for his part, will most surely keep and perform.}

Wherefore, after this promise made by Christ, \emph{ye} must also faithfully, for \emph{your} part, promise in the presence of these \emph{your} Witnesses, and this whole congregation, that \emph{ye will} renounce the devil and all his works, and constantly believe God’s holy Word, and obediently keep his commandments.
\end{leftbar}

\medskip
\pilcrow{Then shall the Priest demand of each of the \emph{Children} to be baptized, severally, these questions following:}

\smallskip
\centerline{\rubric{Question.}}
\centerline{I demand therefore,}

\drop{Dost thou %, in the name of this Child, 
renounce the devil and all his works?\\
\qa{Answer.}  I renounce him.

\smallskip
\centerline{\rubric{Question.}}
Dost thou renounce the vain pomp and glory of the world, with all covetous desires of the same?

\qa{Answer.}  I renounce them.

\smallskip
\centerline{\rubric{Question.}}

Does thou renounce the sinful desires of the flesh, so that thou wilt not follow, nor be led by them?}

\qa{Answer.}  I renounce them all.

\medskip

% [sarum: Unction on breast and back)]


\centerline{\rubric{Question.}}
\drop{Dost thou believe in God the Father Almighty, Maker of heaven and earth?}

\qa{Answer.}  This I believe.

\smallskip
\centerline{\rubric{Question.}}
%And
Dost thou believe 
in Jesus Christ his only-begotten Son our Lord? And that he was conceived by the Holy Ghost; born of the Virgin Mary; that he suffered under Pontius Pilate, was crucified, dead, and buried; that he went down into hell, and also did rise again the third day; that he ascended into heaven, and sitteth at the right hand of God the Father Almighty; and from thence shall come again at the end of the world, to judge the quick and the dead?

\qa{Answer.}  This I believe.

\smallskip
\centerline{\rubric{Question.}}

And dost thou believe in the Holy Ghost; the holy Catholick Church; the Communion of Saints; the Remission of sins; the Resurrection of the flesh; and everlasting life after death?

\qa{Answer.} All this I stedfastly believe.

\medskip
\centerline{\rubric{Question.}}

\drop{Wilt thou be baptized in this faith?\\
\qa{Answer.} That is my desire.}

\smallskip
\centerline{\rubric{Question.}}

\drop{Wilt thou then obediently keep God’s holy will and commandments, and walk in the same all the days of thy life?}

\qa{Answer.} I will endeavour so to do, God being my helper. % 1928En

% \qa{Answer.} I will. 1662



% \begin{leftbar}

% \centerline{\rubric{Question.}}

% \drop{Dost thou renounce the devil and all his works, the vain pomp and glory of the world, with all covetous desires of the same, and the sinful desires of the flesh, so that thou wilt not follow, nor be led by them?}

% \qa{Answer.} I renounce them all.

% \centerline{\rubric{Question.}}

% Dost thou profess the Christian faith?

% \qa{Answer.} I do.

% \rubric{Then shall be said by the Candidates with the Priest and the Witnesses as followeth}

% \drop{I believe in God the Father Almighty, Maker of heaven and earth:}

% And in Jesus Christ his only Son our Lord, Who was conceived by the Holy Ghost, Born of the Virgin Mary, Suffered under Pontius Pilate, Was crucified, dead, and buried, He descended into hell; The third day he rose again from the dead, He ascended into heaven, And sitteth on the right hand of God the Father Almighty; From thence he shall come to judge the quick and the dead.

% I believe in the Holy Ghost; The holy Catholick Church; The Communion of Saints ; The Forgiveness of sins ; The Resurrection of the body; And the Life everlasting. Amen.

% \centerline{\rubric{Question.}}

% \drop{Wilt thou be baptized in this faith?\\
% \qa{Answer.} That is my desire.}

% \centerline{\rubric{Question.}}

% \drop{Wilt thou then obediently keep God’s holy will and commandments, and walk in the same all the days of thy life?}

% \qa{Answer.} I will endeavour so to do, God being my helper.
% \end{leftbar}








\section{The Blessing of the Water}


\centerline{\pilcrow{After which the Priest shall proceed, saying,}}
% The water in the fonte shalbe chaunged every moneth once at the lest, and afore any child be Baptized in the water so chaunged, the priest shall say at the font these prayers folowing.

\V The Lord be with you.  \R And with thy spirit.


\centerline{Let us pray.}
\drop{O most merciful God our Saviour Jesu Christ, who hast ordained the element of water for the regeneration of thy faithful people, Upon whom, being baptized in the river of Jordan, the Holy Ghost came down in likeness of a dove: Send down, we beseech thee, the same thy Holy Spirit to assist us, and to be present at this our invocation of thy Holy Name: Sanctify \grealtcross\ this fountain of baptism, thou that art the Sanctifier of all things, that by the power of thy Word all those that shall be baptized therein may be spiritually regenerated, and made the children of everlasting adoption. \R Amen.}

\smallskip
\centerline{\pilcrow{Then shall the Priest say,}}
\drop{O merciful God, grant that the old Adam in \emph{these Children} [\rubric{or} \emph{these Persons}] may be so buried, that the new man may be raised up in \emph{them}.  \R Amen.}

Grant that all %carnal
sinful %20th c. books
affections may die in \emph{them}, and that all things belonging to the Spirit may live and grow in \emph{them}. \R Amen.

Grant that \emph{they} may have power and strength to have victory, and to triumph, against the devil, the world, and the flesh. \R Amen.

%<1549>
Whosoever shall confess thee, O Lord, recognise him also in thy kingdom. \R Amen.

Grant that all sin and vice here may be so extinct, that they never have power to reign in thy servants. \R Amen.

Grant that whosoever here shall begin to be of thy flock, may evermore continue in the same. \R Amen.

Grant that all they which for thy sake in this life do deny and forsake themselves, may win and purchase thee, O Lord, which art everlasting treasure. \R Amen.
%</1549>

Grant that whosoever is here dedicated to thee by our office and ministry may also be endued with heavenly virtues, and everlastingly rewarded, through thy mercy, O blessed Lord God, who dost live, and govern all things, world without end. \R Amen.

\medskip
%1549
\V The Lord be with you.  \R And with thy spirit.

% 1928En, and other modern:
\V Lift up your hearts.  \R We lift them up unto the Lord.

\V Let us give thanks unto our Lord God. \R It is meet and right so to do.
\drop{It is very meet, right, and our bounden duty, that we should % [at all times, and in all places,] (all sources omit)
give thanks unto thee, O Lord, Holy Father, Almighty, Everlasting God, for that thy
% \drop{Almighty, everliving God, whose most * 
dearly beloved Son Jesus Christ, for the forgiveness of our sins, did shed out of his most precious side both water and blood; and gave commandment to his disciples, that they should go teach all nations, and baptize them In the Name of the Father, and of the Son, and of the Holy Ghost: Regard, we beseech thee, the supplications of thy congregation;}
%1662
Sanctify \grealtcross\ this Water to the mystical washing away of sin; and grant that \emph{these Children} [\rubric{or} \emph{these} thy \emph{Servants}], now to be baptized therein,
% Or. 1549
% and grant that all thy servants which shall be baptized in this water, prepared for the ministration of thy holy sacrament, 
may receive the fulness of thy grace, and ever remain in the number of thy faithful and elect children;
%  through Jesus Christ our Lord. \R Amen.}
% 1928Am:
through the same Jesus Christ our Lord, to whom, with thee, in the unity of the Holy Spirit, be all honour and glory, now and evermore. \R Amen.


\section{The Baptism}

\pilcrow{Then the Minister shall take the Child into his hands, and shall say to the Godfathers and Godmothers,}
\centerline{Name this Child.}

{\footnotesize\rubric{And then naming it after them (if they shall certify him that the Child may well endure it) % he shall dip it in the Water discreetly and warily, saying,
he shall dip it in the Water thrice. First dipping the right side, second the left side, the third time dipping the face towards the font, so it be discreetly and warily done, saying,}\par}

% And then naming it after them, he shall dip it in the water, or pour water upon it, saying,
% he shall dip it in the Water discreetly and warily, saying,

\lettrine{\emph{N.}}{ I} baptize thee in the Name of the Father, and of the Son, and of the Holy Ghost. Amen.

{\footnotesize\rubric{But if they certify that the Child is weak, it shall suffice to pour Water upon it, saying the foresaid words, N. \emph{I baptize thee \etc}}\par}

\smallskip
\begin{leftbar}
\pilcrow{But NOTE, That if the Person to be baptized be of Riper Years, the Minister shall take \emph{him} by the right hand, and placing \emph{him} conveniently by the Font, according to his discretion, shall ask the Godfathers and Godmothers the Name; and then shall dip \emph{him} in the Water, or pour Water upon \emph{him}, using the same form of words.}
\end{leftbar}

\medskip

\pilcrow{Then the Priest shall annoint the \emph{Infant} upon the head with chrism, saying,}
    
\drop{Almighty God, the Father of our Lord Jesus Christ, who hath regenerated thee by water and the Holy Ghost, and hath given unto thee remission of all thy sins, may he vouchsafe to anoint thee with the unction \grealtcross\ of his Holy Spirit, and bring thee to the inheritance of everlasting life. \R Amen.}


% \centerline{\rubric{Then the Priest shall say,}}
% \drop{We receive this \emph{Child} into the congregation of Christ’s flock, \grealtcross\ and do sign \emph{him} with the sign of the Cross, in token that hereafter \emph{he} shall not be ashamed to confess the faith of Christ crucified, and manfully to fight under his banner, against sin, the world, and the devil; and to continue Christ’s faithful soldier and servant unto \emph{his} life’s end. \R Amen.}
% * here the Priest shall make a Cross upon the Child's forehead.

\medskip

%1549
\pilcrow{Then the Godfathers and Godmothers shall take and lay their hands upon the \emph{Child}, and the Priest shall put upon \emph{him his} white vesture, commonly called the chrysom; and say}

\drop{Take this white vesture as a token of the innocency which, by God’s grace, in this holy sacrament of Baptism, is given unto thee; and for a sign whereby thou art admonished, so long as thou livest, to give thyself to innocency of living: that after this transitory life thou mayest be partaker of the life everlasting.}
     
\medskip

%india supplement.
\pilcrow{Then the Minister may deliver to each candidate a burning light, saying,}
% \drop{Receive this burning light, and walk in the light ft by faith in Jesus Christ. Amen.}
% or, South Africa % and Alcuin
\drop{Receive the light of Christ, that when the Bridegroom cometh thou mayest go forth with all the saints to meet him; and see that thou keep the grace of thy baptism.}

\medskip

\pilcrow{When there are many to be baptized, this order of baptizing, annointing, putting on the chrysom, and delivering the light, shall be used severally with every \emph{Child}. Those that be first baptized departing from the font, and remainyng in some convenient place within the Church until all be baptized.}


\section{The Thanksgiving}
\centerline{\pilcrow{Then shall the Priest say,}}
\drop{Seeing now, dearly beloved brethren, that \emph{these Children} [\rubric{or} \emph{these Persons}] \emph{are} regenerate, and grafted into the body of Christ’s Church, let us give thanks unto Almighty God for these benefits; and with one accord make our prayers unto him, that \emph{these Children} [\rubric{or} \emph{these Persons}] may lead the rest of \emph{their} life according to this beginning.}


% Then shall be said, all kneeling;
% \drop{Our Father}
\smallskip

\centerline{\rubric{Then shall the Priest say,}}
\drop{We yield thee hearty thanks, most merciful Father, that it hath pleased thee to regenerate \emph{these Infants} [\rubric{or} \emph{these Persons}] with thy Holy Spirit, to receive \emph{them} for thine own \emph{Children} by adoption, and to incorporate \emph{them} into thy holy Church. And humbly we beseech thee to grant, that \emph{they}, being dead unto sin, and living unto righteousness, and being buried with Christ in his death, may crucify the old man, and utterly abolish the whole body of sin; and that, as \emph{they are} made partaker of the death of thy Son, \emph{they} may also be partaker of his resurrection; so that finally, with the residue of thy holy Church, \emph{they} may be \emph{inheritors} of thine everlasting kingdom; through Christ our Lord. \R Amen.}


\bigskip

\section{The Duties of the Godfathers and Godmothers}


\pilcrow{Then the Priest shall say to the Godfathers and Godmothers and Parents this Exhortation following.}
\drop{Forasmuch as \emph{these Children have} promised by you \emph{their} sureties to renounce the devil and all his works, to believe in God, and to serve him: ye must remember, that it is your parts and duties to see that \emph{these Infants} be taught, so soon as \emph{they} shall be able to learn, what a solemn vow, promise, and profession, \emph{they have} here made by you. And that \emph{they} may know these things the better, ye shall call upon \emph{them} to hear Sermons; and chiefly ye shall provide, that \emph{them} may learn the Creed, the Lord’s Prayer, and the Ten Commandments, in the vulgar tongue, and all other things which a Christian ought to know and believe to his soul’s health; and that \emph{these Children} may be virtuously brought up to lead a godly and a Christian life.}


Will you help \emph{them} to learn and to do all these things?

\qa{Answer.}  I will, the Lord being my helper.

Remember always that Baptism doth represent unto us our profession; which is, to follow the example of our Saviour Christ, and to be made like unto him; that, as he died, and rose again for us, so should we, who are baptized, die from sin, and rise again unto righteousness; continually mortifying all our evil and corrupt affections and daily proceeding in all virtue and godliness of living.

\medskip
% the following is a repetition of the previous exhortation, and the following rubric; and doesn't survive into the modern books.
% \centerline{\rubric{Then shall he add and say,}}
% \drop{Ye are to take care that this Child be brought to the Bishop to be confirmed by him, so soon as he can say the Creed, the Lord's Prayer, and the Ten Commandments, in the vulgar tongue, and be further instructed in the Church-Catechism set forth for that purpose.}
% It is certain by God's Word, that children which are baptized, dying before they commit actual sin, are undoubtedly saved.


\pilcrow{The Minister shall command that the chrysoms be brought to the church, and delivered to the Priests after the accustomed manner, at the purification of the mother of every child: and that the children be brought to the Bishop to be confirmed of him, so soon as they can say, in their vulgar tongue, the Articles of the Faith, the Lord’s Prayer, and the Ten Commandments, and further be instructed in the Catechism set forth for that purpose accordingly as it is there expressed.}

% And so let the congregation depart in the name of the Lord.
\begin{leftbar}
\section{The Duties of the Witnesses and of the New Baptized}
\pilcrow{Then, all standing up, the Priest shall use this exhortation following; speaking to the Godfathers and Godmothers first.}
\drop{Forasmuch as \emph{these persons have} promised in your presence to renounce the devil and all his works, to believe in God, and to serve him: ye must remember, that it is your part and duty to put \emph{them} in mind, what a solemn vow, promise, and profession \emph{they have} now made before this congregation, and especially before you \emph{their} chosen witnesses. And ye are also to call upon \emph{them} to use all diligence to be rightly instructed in God’s holy Word; that so \emph{they} may grow in grace, and in the knowledge of our Lord Jesus Christ, and live godly, righteously, and soberly in this present world.}


{\footnotesize\rubric{(And then, speaking to the new baptized \emph{persons}, he shall proceed, and say,)}\par}


\drop{And as for \emph{you}, who \emph{have} now by Baptism put on Christ, it is \emph{your} part and duty also, being made \emph{the children} of God and of the light by faith in Jesus Christ, to walk answerably to \emph{your} Christian calling, and as becometh the children of light; remembering always that Baptism representeth unto us our profession; which is, to follow the example of our Saviour Christ, and to be made like unto him; that as he died, and rose again for us; so should we, who are baptized, die from sin, and rise again unto righteousness; continually mortifying all our evil and corrupt affections, and daily proceeding in all virtue and godliness of living.}

\medskip
\pilcrow{It is expedient that every person, thus baptized, should be confirmed by the Bishop so soon after his Baptism as conveniently may be.}
\end{leftbar}

\fleuron


\section{Private Baptism}
\pilcrow{When, in consideration of extreme sickness, necessity may require, then the following form shall suffice:}

\pilcrow{The Child (or Person) being named by some one who is present, the Minister shall pour Water upon him, saying these words:}

\lettrine{\emph{N.}}{ I} baptize thee In the Name of the Father, and of the Son, and of the Holy Ghost. Amen.

\pilcrow{After which shall be said the Lord’s Prayer, and the Thanksgiving from the Office, beginning, \emph{We yield thee hearty thanks, \etc}}

\begin{leftbar}
    \pilcrow{But NOTE, That in the case of an Adult, the Minister shall first ask the questions provided in this Office for the Baptism of Adults.}
\end{leftbar}
\pilcrow{In cases of extreme sickness, or any imminent peril, if a Minister cannot be procured, then any person present may administer holy Baptism, using the foregoing form. Such Baptism shall be promptly reported to the authorities.}

\pilcrow{And let them not doubt, but that the \emph{Child} so baptized is lawfully and sufficiently baptized, and ought not to be baptized again.}

\section{The Receiving of one Privately Baptized}
\pilcrow{It is expedient that a Child or Person so baptized be afterward brought to the Church, at which time these parts of the foregoing service shall be used:}

{\footnotesize\rubric{The Gospel, the Questions (omitting the question \emph{Wilt thou be baptized in this Faith?} and the answer thereto), the Annointing, and the remainder of the Office.}\par}



%Am1928
\section{Conditional Baptism}
\pilcrow{If there be reasonable doubt whether any Person were baptized with Water, \emph{In the Name of the Father, and of the Son, and of the Holy Ghost} (which are essential parts of Baptism), such Person may be baptized in the manner herein appointed; saving that, at the immersion or the pouring of water, the Minister shall use this Form of words.}

\drop{If thou art not already baptized, \emph{N.}, I baptize thee In the Name of the Father, and of the Son, and of the Holy Ghost. Amen.}



\section{General Rubrics}

\pilcrow{For every child to be baptized there shall be not fewer than three godparents, of whom at least two shall be of the same sex as the child and of whom at least one shall be of the opposite sex; save that, when three cannot be conveniently had, one godparent shall suffice. Parents may be godparents for their own children provided that the child shall have at least one other godparent. The godparents shall be persons who have been baptized and confirmed and will faithfully fulfil their responsibilities both by their care for the child committed to their charge and by the example of their own godly living. Nevertheless the Minister shall have power to dispense with the requirement of confirmation in any case in which in his judgement need so requires.}

\pilcrow{The Minister shall instruct the parents or guardians of an infant to be admitted to Holy Baptism that the same responsibilities rest on them as are in the service of Holy Baptism required of the godparents.}

\pilcrow{No Minister shall refuse or, save for the purpose of preparing or instructing the parents or guardians or godparents, delay to baptize any infant within his cure that is brought to the church to be baptized, provided that due notice has been given and the provisions relating to godparents are observed. If the Minister shall refuse or unduly delay to baptize any such infant, the parents or guardians may apply to the Bishop of the diocese who shall, after consultation with the Minister, give such directions as he thinks fit.}

\pilcrow{The Minister, before proceeding to the Baptism, shall have satisfied himself that the child presented to him has not already been baptized.}
\begin{leftbar}    
\pilcrow{When any such persons, as are of riper years, are to be baptized, timely notice shall be given to the Bishop, or whom he shall appoint for that purpose, a week before at the least, by the Minister of the Parish, the parents, or some other discreet persons; that so due care may be taken for their examination, whether they be sufficiently instructed in the principles of the Christian Religion; and that they may be exhorted to prepare themselves with prayers and fasting for the receiving of this holy Sacrament.}


\pilcrow{And if they shall be found fit, they shall each choose three, or at the least one, to be their Witnesses, who shall be ready to present them at the Font, immediately after the Second Lesson, either at Morning or Evening Prayer, or (if need so require) at such other time as the Minister in his discretion shall think fit.}

\smallskip

\pilcrow{It is convenient that the Admission to the Catechuminate should precede the Baptism by a week or more.}
\end{leftbar}



\fleuron
