\documentclass[foolscapvopaper,11pt,twoside,openany,extrafontsizes,final]{memoir}
% \documentclass[ebook,12pt,twoside,openany,extrafontsizes,final]{memoir}
\usepackage{fontspec}
\usepackage{lettrine}
\usepackage[autocompile]{gregoriotex}
\usepackage{multicol}
\usepackage{wrapfig}
% \usepackage[rflt]{floatflt}
\usepackage{longtable,tabu}
% \usepackage{lua-visual-debug}
% a Font that supplies Astrological glyphs
\newfontfamily\versiculus{DejaVu Sans}

\directlua{dofile("lua-functions.lua")}
\newcommand{\drop}[1]{
    \directlua{drop("\luaescapestring{\unexpanded{#1}}")}
}


\setsecnumdepth{part} % Only number the parts.

% Chapters
\setlength{\beforechapskip}{0\baselineskip}
\setlength{\afterchapskip}{.5\baselineskip}
\renewcommand*{\chaptitlefont}{\sffamily}
\renewcommand*{\printchaptertitle}[1]{\centering\LARGE{\chaptitlefont #1}}

% Sections: Major chapter divisions; Days in the Psalter
\setbeforesecskip{1em plus 2em minus 1ex}
\setsecheadstyle{\scshape\centering}
\setaftersecskip{1pt}

%Subsection: Collects, Canticles, Psalms
%{\schshape Title}{\itshape incipit}{\normal scriptural reference.}
\setbeforesubsecskip{1em plus 0ex minus 1ex}
\setsubsecheadstyle{\centering}
\setaftersubsecskip{1pt}

\newcommand{\stylesubsec}[3]{\small{\mbox{\scshape#1}} {\mbox{\itshape\red #2}} {\mbox{#3}}}
\newcommand{\stylesec}[3]{\section[#3]{#1\\ {\normalfont\small#2\\ } #3}}
\newcommand{\scripture}[1]{\hspace*{\fill}{\mbox{\small\itshape\red #1}}}
%Subsubsection
%Descriptive names of prayers, etc.
\setbeforesubsubsecskip{1em plus 0ex minus 1ex}
\setsubsubsecheadstyle{\itshape\footnotesize\red\centering}
\setaftersubsubsecskip{1pt}

\emergencystretch 1.5em


\newcommand \fleuron{{\centering\Large\red❦\par}}

\nouppercaseheads
\makepagestyle{mystyle}
\makeevenhead{mystyle}{}{\scshape\leftmark}{}
\makeoddhead{mystyle}{}{\scshape\rightmark}{}
\makeevenfoot{mystyle}{}{\thepage}{}
\makeoddfoot{mystyle}{}{\thepage}{}
\makepsmarks{mystyle}{%
  \createmark{chapter}{left}{nonumber}{}{}}

\pagestyle{mystyle}



% For the wrapped rubric on the embertide intercession
\setlength{\intextsep}{2pt}
% Get rid of extra space around longtable
\setlength{\LTpre}{.5em}
\setlength{\LTpost}{.5em}


\setmainfont[  
    Numbers={OldStyle},
    Ligatures={Rare,Historic},
    % ItalicFeatures={Colour=990000}
]{EBGaramond}
\setsansfont{UnifrakturMaguntia}[StylisticSet={1},CharacterVariant={11,12,4:1}]


\newcommand \red{\addfontfeature{Color=990000}}
\newcommand \R{{\red℟.\ }}
\newcommand \V{{\red℣.\ }}
\newcommand \ant{{\scshape\small\red Ant.\ }}
\newcommand \etc{\emph{\red\&c.}}
\newcommand \minorheading[2]{{\unexpanded\expandafter{\scshape #1}}\par{\unexpanded\expandafter{\itshape #2}}}
\newcommand \subseccaption[2]{\subsection{#1 {\itshape\small\red #2}}}

\newcommand \rubric[1]{\emph{\footnotesize\red #1}}
\newcommand \pilcrow[1]{\par\footnotesize\noindent\makebox[1em][l]{¶ }\hangindent1em\rubric{#1}\par\normalsize}

\newcommand \cross {\grecross\ }

\medievalpage
\raggedbottom
\checkandfixthelayout


\begin{document}
\frontmatter
\title{Book of Common Prayer}
\author{Michael Shirk}
\date{MMXX}
% \maketitle

\settocdepth{chapter}
\tableofcontents*


Preface


\mainmatter

\include{kalendar}
\mainmatter
% \part{Service Music}
\include{mattins}
\include{evensong}
\include{quicunque}
\include{litany}
\include{collects}


% \include{mass}

The Order of Baptism
The Catachism
The Order of Confirmation
The Form of Solemnization of Matrimony
The Order for the Visitation of the Sick, and the Communion of the Sick
The Order for the Burial of the Dead.
The Thanksgiving of Women after Child-birth

The Psalter

Appendix?
Blessing of anything
Blessing of Holy Water
Personal prayers (1928 bcp)

\end{document}

Chapter: Big all-caps
Section: Regular all caps (Te deum, venite)
Subheading: Regular (Scriptural citation)
subsubsection (italic)


Chapter:   
The order for (caps, 1em?)
MORNING PRAYER (caps, 2em?)
daily throughout the year (caps, 1em?)

Section
;
a supplication
;
The First Sunday in Advent
;
The Nativity of Our Lord, or the 1
Birthday of Christ 1 
commonly called 2
Christmas day 1
[December 25] 2
;
The Sunday called  (it) 3
Septuagesima (all caps)1
or the third Sunday before Lent. 3
;
Day 9. Evening Prayer

subsection
VENITE< EXULTEMUS DOMINO (caps, 1em?) 1
Psalm 95. (.75em?) 2

;
Psalm 47 (ALL CAPS); Omnes gentes, plaudite. (it)


Anglican Breviary
The Song of Zacharias
_Benedictus Dominus, Deus Israel._
Luke I.68.

Diurnal
The Song of the Blessed Virgin (SC)
_Magnificat_

subsubsection
2 Corinthians 13. (it, 1em)
; 
The Collect
;
The Epistle. (it) Romans 13.8 (normal)

Baskerville:



_Te Deum Laduamus_
_Beedictus_ S. LUKE I. 68
_Jubilate DEO. PSAL. 100.

_the Second Collect, for Peace_


_The_ Nativity _of our Lord, or the_ Birth-Day _of_ CHRIST, _commonl called_ Christmas-day.



Psalms:
_The_ 1. _Day.     Psalms   |     Psalms  _The_ 2. _Day._
PSAL. 8. _Domine, Dominus noster

M O R N I N G  P R A Y E R
PSAL. 9. _Confitebor tibi.

Current:
Day 12 : Mn.   THE PSALMS   Ps. 64 | Ps. 65  THE PSALMS  Day 12 : Ev.