\chapter{Blessings}

\section{The order for the Conjuring of Water}

\smallskip
\pilcrow{The Priest prepares the salt as follows,}

\drop{I exorcize thee, creature of salt, by the living \grealtcross\ God, by the holy \grealtcross\ God, by the omnipotent \grealtcross\  God, that thou mayest be purified from all evil influence, in the Name of Him who is Lord of Angels and of men, who filleth the whole earth with his majesty and glory.  \R Amen.}

\drop{We pray thee, O God, in thy boundless lovingkindness to stretch forth the right hand of thy power upon this creature of salt which we \grealtcross\ bless and \grealtcross\ hallow in thy holy Name.  Grant that this salt may make for health of mind and body to all who partake thereof, and that there may be banished from the place where it is used every power of adversity and every illusion or artifice of evil; through Christ our Lord.  \R Amen.}

\smallskip

\pilcrow{The Priest prepares the water as follows,}

\drop{I exorcize thee, creature of water, by the living \grealtcross\ God, by the holy \grealtcross\ God, by the omnipotent \grealtcross\  God, that thou mayest be purified from all evil influence, in the Name of Him who is Lord of Angels and of men, who filleth the whole earth with his majesty and glory.  \R Amen.}

\drop{O God, who for the helping and safeguarding of men dost hallow the water set apart for the service of thy holy Church, send forth thy light and thy power upon this element of water which we \grealtcross\ bless and \grealtcross\ hallow in thy holy Name.  Grant that whosoever uses this water in faithfulness of spirit may be strengthened in all goodness, and that everything sprinkled with it may be made holy and pure and guarded from all assaults of evil; through Christ our Lord.  \R Amen.}

\smallskip

\pilcrow{The Priest casts the salt thrice into the water crosswise, as he says the following,}
\drop{Let salt and water mingle together in the Name of the \grealtcross\ Father, and of the \grealtcross\  Son, and of the Holy \grealtcross\ Ghost.  \R Amen.}


\smallskip

\V The Lord be with you.  \R And with thy spirit.

\drop{O God, the giver of invincible strength and King of irresistible power, whose splendour shines throughout the whole of creation: We pray thee to look upon this thy creature of salt and water, to pour down upon it the radiance of thy \grealtcross\ blessing and to \grealtcross\ hallow it with the dew of thy lovingkindness, that wherever it shall be sprinkled and thy holy Name shall be invoked in prayer, every noble aspiration may be strengthened, every good resolve made firm, and the fellowship of the Holy Spirit vouchsafed to us who place our trust in thee; thou who with the Son livest and reignest in the unity of the same Holy Spirit, God throughout all ages of ages.  \R Amen.}

\pilcrow{The Altar, clergy, and people are then sprinkled, while the following is sung.}

Anthem. Thou shalt purge me, \star\ O Lord, with hyssop, and I shall be clean: thou shalt wash me, and I shall be whiter than snow.

\centerline{\rubric{But from Easter until Whitsunday,}}

Anthem.
\begin{leftbar}
    
\V O Lord, shew thy mercy upon us.  \R And grant us thy salvation. (Alleluya.)


\V Glory be to the Father, and to the Son, and to the Holy Ghost; As it was in the beginning, is now, and ever shall be, world without end. Amen.

\star\ Thou shalt wash me, and I shall be whiter than snow.

\pilcrow{This Anthem is said at the sprinkling of holy water on all Sundays throughout the year, except from Easter to the Feast of the Holy Trinity.  It shall be said even on Passion Sunday and Palm Sunday with \emph{Glory be to the father \etc}}

\pilcrow{From Easter to the Feast of the Holy Trinity the following Anthem should be said at the sprinkling of holy water, the precentor commencing the Anthem.}
\drop{I saw water issuing out of the temple on the right-hand side, alleluya.  And all to whom that water came were made whole, and shall say, Alleluya, alleluya.}

\rubric{Ps.} O give thanks unto the {\scshape Lord}, for he is grácious; because his mercy endureth for ever.

\rubric{Ant.} I saw water, \etc

\V Glory be to the Father, and to the Son, and to the Holy Ghost;As it was in the beginning, is now, and ever shall be, world without end. Amen.

\end{leftbar}


\centerline{Let us pray.}

\drop{Graciously hear us, O Lord, Holy Father, Almighty, everlasting God; and vouchsafe to send thy holy Angel from Heaven to keep, cherish, protect, visit, and defend all who are assembled in this holy habitation.  Through Christ our Saviour.  \R Amen.}

\fleuron