\chapter[The Burial of the Dead]{\stylechapter{The Order for}{The Burial of the Dead}{}}

\pilcrow{Here is to be noted, that the Office ensuing is not to be used for any that die unbaptized, or excommunicate, or have laid violent hands upon themselves.}

% Canada:1918
\pilcrow{Note also, That when this Office is not to be used, the Priest may at the grave read the Sentences beginning \emph{Man that is born,} followed by the Lesser Litany, the Lord’s Prayer, one or more Collects from this Book at his discretion, and \emph{The grace of the Lord \etc}}

\section{The Procession}

\pilcrow{The Minister and Clerks, meeting the Corpse at the entrance of the Church-yard, and going before it, either into the Church, or towards the Grave, shall say, or sing [one or more of the following Sentences; together with one or more of the Penitential Psalms \emph{(6, 32, 38, 51, 102, 130, 143)} if need so require.],}
\drop{I am the resurrection and the life, saith the Lord: he that believeth in me, though he were dead, yet shall he live: and whosoever liveth and believeth in me shall never die.\scripture{St.~John xi.~25, 26.}}

\drop{I know that my Redeemer liveth, and that he shalt stand at the latter day upon the earth. And though after my skin 
worms destroy this body, %KJV
% hath been thus destroyed, %adapted from rsv
% earth: and though this body be destroyed, yet shall I see God: %am1928
yet in my flesh shall I see God: whom I shall see for myself, and mine eyes shall behold, and not another. \scripture{Job xix.~25, 26, 27.}}


\drop{We brought nothing into this world, and it is certain we can carry nothing out. The {\scshape Lord} gave, and the {\scshape Lord} hath taken away; blessed be the Name of the {\scshape Lord}.\scripture{1 Timothy.~vi.~7. Job i.~21.}}

\section{The Service in church}
\pilcrow{After they are come into the Church, shall be sung or said one or more of these Psalms following. Note, that at the end of each of the Psalms the \emph{Gloria Patri} shall be omitted.} %1954


%1928 proposed
% \pilcrow{Before and after any psalm or group of psalms may be said or sung the Anthem following,}

% \drop{O Saviour of the world, who by thy Cross and precious Blood hast redeemed us, Save us and help us, we humbly beseech thee, O Lord.}
{\red\scshape Anthem.} O Saviour of the world, who by thy Cross and precious Blood hast redeemed us, \star\ Save us, and help us, we humbly beseech thee, O Lord.

%1929 
% After they are come into the church, shall be sung or said one or more of these Psalms following.


% Anthem. O Saviour of the world, who by thy Cross and precious Blood hast redeemed us : save us, and help us, we humbly beseech thee, O Lord.

%(1918)(1926)(1928e)(1929)(1954)
\subsection[{Psalm 23}]{\stylesubsec{Psalm 23.}{Dominus regit me.}{}}
\drop{The {\scshape Lord} is my shepherd; \star\ therefore can I lack nothing.}

2\enspace He shall feed me in a green pasture, \star\ and lead me forth beside the waters of comfort.

3\enspace He shall convert my soul, \star\ and bring me forth in the paths of righteousness, for his Name’s sake.

4\enspace Yea, though I walk through the valley of the shadow of death, I will fear no evil; \star\ for thou art with me; thy rod and thy staff comfort me.

5\enspace Thou shalt prepare a table before me against them that trouble me; \star\ thou hast anointed my head with oil, and my cup shall be full.

6\enspace But thy loving-kindness and mercy shall follow me all the days of my life; \star\ and I will dwell in the house of the Lord for ever.

Rest eternal grant unto them, O Lord \star\ and let light perpetual shine upon them.


% Psalm 39.  (1918)(1662)(1926)(1928a)(1929)
\subsection[{Psalm 39}]{\stylesubsec{Psalm 39.}{Dixi, custodiam.}{}}
\drop{I said, I will take heed to my ways, \star\ that I offend not in my tongue.}

2\enspace I will keep my mouth as it were with a bridle, \star\ while the ungodly is in my sight.

3\enspace I held my tongue, and spake nothing: \star\ I kept silence, yea, even from good words; but it was pain and grief to me.

4\enspace My heart was hot within me: and while I was thus musing the fire kindled, \star\ and at the last I spake with my tongue:

5\enspace {\scshape Lord}, let me know mine end, and the number of my days; \star\ that I may be certified how long I have to live.

6\enspace Behold, thou hast made my days as it were a span long, \star\ and mine age is even as nothing in respect of thee; and verily every man living is altogether vanity.

7\enspace For man walketh in a vain shadow, and disquieteth himself in vain; \star\ he heapeth up riches, and cannot tell who shall gather them.

8\enspace And now, Lord, what is my hope? \star\ truly my hope is even in thee.

9\enspace Deliver me from all mine offences; \star\ and make me not a rebuke unto the foolish.

10\enspace I became dumb, and opened not my mouth; \star\ for it was thy doing.

11\enspace Take thy plague away from me: \star\ I am even consumed by the means of thy heavy hand.

12\enspace When thou with rebukes dost chasten man for sin, thou makest his beauty to consume away, like as it were a moth fretting a garment: \star\ every man therefore is but vanity.

13\enspace Hear my prayer, O {\scshape Lord}, and with thine ears consider my calling; \star\ hold not thy peace at my tears;

14\enspace For I am a stranger with thee, and a sojourner, \star\ as all my fathers were.

15\enspace O spare me a little, that I may recover my strength, \star\ before I go hence, and be no more seen.

Rest eternal grant unto them, O Lord \star\ and let light perpetual shine upon them.

% Psalm 90.  (1918)(1662)(1926)(1928a)(1929)(1954)
\subsection[{Psalm 90}]{\stylesubsec{Psalm 90.}{Domine, refugium.}{}}
\drop{Lord, thou hast been our refuge, \star\ from one generation to another.}

2\enspace Before the mountains were brought forth, or ever the earth and the world were made, \star\ thou art God from everlasting, and world without end.

3\enspace Thou turnest man to destruction; \star\ again thou sayest, Come again, ye children of men.

4\enspace For a thousand years in thy sight are but as yesterday, \star\ seeing that is past as a watch in the night.
5\enspace As soon as thou scatterest them they are even as a sleep; \star\ and fade away suddenly like the grass.

6\enspace In the morning it is green, and groweth up; \star\ but in the evening it is cut down, dried up, and withered.

7\enspace For we consume away in thy displeasure, \star\ and are afraid at thy wrathful indignation.

8\enspace Thou hast set our misdeeds before thee; \star\ and our secret sins in the light of thy countenance.

9\enspace For when thou art angry all our days are gone: \star\ we bring our years to an end, as it were a tale that is told.

10\enspace The days of our age are threescore years and ten; and though men be so strong that they come to fourscore years, \star\ yet is their strength then but labour and sorrow; so soon passeth it away, and we are gone.

11\enspace But who regardeth the power of thy wrath? \star\ for even thereafter as a man feareth, so is thy displeasure.

12\enspace So teach us to number our days, \star\ that we may apply our hearts unto wisdom.

13\enspace Turn thee again, O {\scshape Lord}, at the last, \star\ and be gracious unto thy servants.

14\enspace O satisfy us with thy mercy, and that soon: \star\ so shall we rejoice and be glad all the days of our life.

15\enspace Comfort us again now after the time that thou hast plagued us; \star\ and for the years wherein we have suffered adversity.

16\enspace Shew thy servants thy work, \star\ and their children thy glory.

17\enspace And the glorious majesty of the {\scshape Lord} our God be upon us: \star\ prosper thou the work of our hands upon us, O prosper thou our handywork.

Rest eternal grant unto them, O Lord \star\ and let light perpetual shine upon them.


% Psalm 130. (1928a)(1928e)(1929)(1954)
\subsection[{Psalm 130}]{\stylesubsec{Psalm 130.}{De profundis.}{}}

\drop{Out of the deep have I called unto thee, O {\scshape Lord}; \star\ Lord, hear my voice.}

2\enspace O let thine ears consider well \star\ the voice of my complaint.

3\enspace If thou, {\scshape Lord}, wilt be extreme to mark what is done amiss, \star\ O Lord, who may abide it?

4\enspace For there is mercy with thee; \star\ therefore shalt thou be feared.

5\enspace I look for the {\scshape Lord}; my soul doth wait for him; \star\ in his word is my trust.

6\enspace My soul fleeth unto the Lord \star\ before the morning watch, I say, before the morning watch.

7\enspace O Israel, trust in the {\scshape Lord}, for with the {\scshape Lord} there is mercy, \star\ and with him is plenteous redemption.

8\enspace And he shall redeem Ísrael \star\ from all his sins.

Rest eternal grant unto them, O Lord \star\ and let light perpetual shine upon them.


\rubric{Anthem.} O Saviour of the world, who by thy Cross and precious Blood hast redeemed us, \star\ Save us, and help us, we humbly beseech thee, O Lord.


\medskip
\pilcrow{Then shall follow the Lesson, taken out of the fifteenth Chapter of the former Epistle of Saint Paul to the Corinthians.}
\centerline{1 Corinthians 15.~20.}
\drop{Now is Christ risen from the dead, and become the first-fruits of them that slept. For since by man came death, by man came also the resurrection of the dead. For as in Adam all die, even so in Christ shall all be made alive. But every man in his own order: Christ the firstfruits; afterward they that are Christ’s, at his coming. Then cometh the end, when he shall have delivered up the kingdom to God, even the Father; when he shall have put down all rule, and all authority, and power. For he must reign, till he hath put all enemies under his feet. The last enemy that shall be destroyed is death.
\begin{leftbar}
For he hath put all things under his feet. But when he saith, all things are put under him, it is manifest that he is excepted, which did put all things under him. And when all things shall be subdued unto him, then shall the Son also himself be subject unto him that put all things under him, that God may be all in all. Else what shall they do which are baptized for the dead, if the dead rise not at all? Why are they then baptized for the dead? and why stand we in jeopardy every hour? I protest by your rejoicing, which I have in Christ Jesus our Lord, I die daily. If after the manner of men I have fought with beasts at Ephesus, what advantageth it me, if the dead rise not? Let us eat and drink, for to-morrow we die. Be not deceived: evil communications corrupt good manners. Awake to righteousness, and sin not: for some have not the knowledge of God. I speak this to your shame.
\end{leftbar}

But some man will say, How are the dead raised up? and with what body do they come? Thou fool, that which thou sowest is not quickened, except it die. And that which thou sowest, thou sowest not that body that shall be, but bare grain, it may chance of wheat, or of some other grain: But God giveth it a body, as it hath pleased him, and to every seed his own body. All flesh is not the same flesh; but there is one kind of flesh of men, another flesh of beasts, another of fishes, and another of birds. There are also celestial bodies, and bodies terrestrial; but the glory of the celestial is one, and the glory of the terrestrial is another. There is one glory of the sun, and another glory of the moon, and another glory of the stars; for one star differeth from another star in glory. So also is the resurrection of the dead: It is sown in corruption; it is raised in incorruption: It is sown in dishonour; it is raised in glory: It is sown in weakness; it is raised in power: It is sown a natural body; it is raised a spiritual body. There is a natural body, and there is a spiritual body. And so it is written, The first man Adam was made a living soul; the last Adam was made a quickening spirit. Howbeit, that was not first which is spiritual, but that which is natural; and afterward that which is spiritual. The first man is of the earth, earthy: the second man is the Lord from heaven. As is the earthy, such are they that are earthy: and as is the heavenly, such are they also that are heavenly. And as we have borne the image of the earthy, we shall also bear the image of the heavenly. 

Now this I say, brethren, that flesh and blood cannot inherit the kingdom of God; neither doth corruption inherit incorruption. Behold, I shew you a mystery: We shall not all sleep, but we shall all be changed, in a moment, in the twinkling of an eye, at the last trump, (for the trumpet shall sound,) and the dead shall be raised incorruptible, and we shall be changed. For this corruptible must put on incorruption, and this mortal must put on immortality. So when this corruptible shall have put on incorruption, and this mortal shall have put on immortality; then shall be brought to pass the saying that is written, Death is swallowed up in victory. O death, where is thy sting? O grave, where is thy victory? The sting of death is sin, and the strength of sin is the law. But thanks be to God, which giveth us the victory through our Lord Jesus Christ. Therefore, my beloved brethren, be ye stedfast, unmoveable, always abounding in the work of the Lord, forasmuch as ye know that your labour is not in vain in the Lord.}

% \begin{leftbar}
\centerline{\rubric{Or one of the following Lessons:}}
\centerline{2 Corinthians 4.~16-end. and 5.~1-10.}
\drop{For which cause we faint not; but though our outward man perish, yet the inward man is renewed day by day. For our light affliction, which is but for a moment, worketh for us a far more exceeding and eternal weight of glory; While we look not at the things which are seen, but at the things which are not seen: for the things which are seen are temporal; but the things which are not seen are eternal. For we know that if our earthly house of this tabernacle were dissolved, we have a building of God, an house not made with hands, eternal in the heavens. For in this we groan, earnestly desiring to be clothed upon with our house which is from heaven: If so be that being clothed we shall not be found naked. For we that are in this tabernacle do groan, being burdened: not for that we would be unclothed, but clothed upon, that mortality might be swallowed up of life. Now he that hath wrought us for the selfsame thing is God, who also hath given unto us the earnest of the Spirit. Therefore we are always confident, knowing that, whilst we are at home in the body, we are absent from the Lord: (For we walk by faith, not by sight:) We are confident, I say, and willing rather to be absent from the body, and to be present with the Lord. Wherefore we labour, that, whether present or absent, we may be accepted of him. For we must all appear before the judgment seat of Christ; that every one may receive the things done in his body, according to that he hath done, whether it be good or bad.}


\medskip
\centerline{Revelation 7.~9.}%-17  Proofed 12-17-21
\drop{After this I beheld, and, lo, a great multitude, which no man could number, of all nations, and kindreds, and people, and tongues, stood before the throne, and before the Lamb, clothed with white robes, and palms in their hands; and cried with a loud voice, saying, Salvation to our God which sitteth upon the throne, and unto the Lamb. And all the angels stood round about the throne, and about the elders, and the four beasts, and fell before the throne on their faces, and worshipped God, saying, Amen; Blessing, and glory, and wisdom, and thanksgiving, and honour, and power, and might, be unto our God for ever and ever. Amen. And one of the elders answered, saying unto me, What are these which are arrayed in white robes? and whence came they? And I said unto him, Sir, thou knowest. And he said to me, These are they which came out of great tribulation, and have washed their robes, and made them white in the blood of the Lamb. Therefore are they before the throne of God, and serve him day and night in his temple: and he that sitteth on the throne shall dwell among them. They shall hunger no more, neither thirst any more; neither shall the sun light on them, nor any heat. For the Lamb which is in the midst of the throne shall feed them, and shall lead them unto living fountains of waters: and God shall wipe away all tears from their eyes.}


\medskip
\centerline{Revelation 21.~1-7.}
\drop{And I saw a new heaven and a new earth: for the first heaven and the first earth were passed away; and there was no more sea. And I John saw the holy city, new Jerusalem, coming down from God out of heaven, prepared as a bride adorned for her husband. And I heard a great voice out of heaven saying, Behold, the tabernacle of God is with men, and he will dwell with them, and they shall be his people, and God himself shall be with them, and be their God. And God shall wipe away all tears from their eyes; and there shall be no more death, neither sorrow, nor crying, neither shall there be any more pain: for the former things are passed away. And he that sat upon the throne said, Behold, I make all things new. And he said unto me, Write: for these words are true and faithful. And he said unto me, It is done. I am Alpha and Omega, the beginning and the end. I will give unto him that is athirst of the fountain of the water of life freely. He that overcometh shall inherit all things; and I will be his God, and he shall be my son.}
% \end{leftbar}


% Scott has 3 others  more like gospels.


\medskip

\centerline{\rubric{Then the Minister shall say,}}
\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\medskip
\ourFather

\smallskip

\V Enter not into judgement with thy servant, O Lord; \R For in thy sight shall no man living be justified.

\V Grant unto \emph{him} eternal rest; \R And let perpetual light shine upon \emph{him}.

\V We believe verily to see the goodness of the Lord: \R In the land of the living.

\V O Lord, hear our prayer; \R And let our cry come unto thee.

{\centering\pilcrow{THen shall be said one or more of the following Prayers, the Minister first saying,}}

\centerline{Let us pray.}
\drop{Almighty God, with whom do live the spirits of them that depart hence in the Lord, and with whom the souls of the faithful, after they are delivered from the burden of the flesh, are in joy and felicity:
% We give thee hearty thanks, for that it hath pleased thee to deliver this our brother out of the miseries of this sinful world; beseeching thee, that it may please thee, of thy gracious goodness, shortly to accomplish the number of thine elect, and to hasten thy kingdom; that we, with all those that are departed in the true faith of thy holy Name, may have our perfect consummation and bliss, both in body and soul, in thy eternal and everlasting glory; through Jesus Christ our Lord. \R Amen.}
%1549
% \drop{O Lord, with whom do live the spirits of them that be dead; and in whom the souls of them that be elected, after they be delivered from the burden of the flesh, be in joy and felicity: 
Grant unto this thy servant, that the sins which \emph{he} committed in this world be not imputed unto \emph{him}, but that \emph{he}, escaping the gates of hell and the pains of eternal darkness, may ever dwell in the region of light, with Abraham, Isaac, and Jacob, in the place where is no weeping, sorrow, nor heaviness; and when that dreadful day of the general resurrection shall come, make \emph{him} to rise also with the just and righteous, and receive this body again to glory, then made pure and incorruptible: set \emph{him} on the right hand of thy Son Jesus Christ, among thy holy and elect, that then \emph{he} may hear with them these most sweet and comfortable words: Come, ye blessed children of my Father, receive the kingdom prepared for you from the beginning of the world: Grant this, we beseech thee, O merciful Father, through Jesus Christ, our Mediator and Redeemer. \R Amen.}

%This collect from the old mass, moved back with its original ending.
% The Collect
% \drop{O merciful God, the Father of our Lord Jesus Christ, who is the resurrection and the life; ... Resurrection in the last day,}

% we may be found acceptable in thy sight; and receive that blessing, which thy well-beloved Son shall then pronounce to all that love and fear thee, saying, 

% This from the original prayer above:
% Come, ye blessed children of my Father, receive the kingdom prepared for you from the beginning of the world: Grant this, we beseech thee, O merciful Father, through Jesus Christ, our Mediator and Redeemer. \R Amen.

% \begin{leftbar}
% 1923:
% BLESSED Lord, who art the Father of mercies and the God of all consolation: We beseech thee of thy great goodness, to comfort those who by the death of this our brother are sorely bereaved; and teach us so to number our days that, while we live here, we may seriously apply our hearts to that holy and heavenly wisdom, which may in the end bring us to life everlasting, through the merits of Jesus Christ, thine only Son our Lord. Amen.

\smallskip
% 1928
\subsection{\stylesubsec{}{The Collects.}{}}
\drop{O Father of all, we pray to thee for those whom we love, but see no longer. Grant them thy peace; let light perpetual shine upon them; and in thy loving wisdom and almighty power work in them the good purpose of thy perfect will; through Jesus Christ our Lord. \R Amen.}

\smallskip

\drop{Almighty God, Father of all mercies and giver of all comfort: Deal graciously, we pray thee, with those who mourn, that casting every care on thee, they may know the consolation of thy love; through Jesus Christ our Lord. \R Amen.}

\smallskip

\drop{O heavenly Father, who in thy Son Jesus Christ, hast given us a true faith, and a sure hope: Help us, we pray thee, to live as those who believe and trust in the Communion of Saints, the forgiveness of sins, and the resurrection to life everlasting, and strengthen this faith and hope in us all the days of our life: through the love of thy Son, Jesus Christ our Saviour. \R Amen.}

\smallskip

{\centering\footnotesize\rubric{Here may follow the Collect of \emph{All Saints’ Day}, or that of the \emph{Twelfth Sunday after Trinity}, or others from the \emph{Prayers upon Several Occasions.}}\par}
% \end{leftbar}

\medskip
\centerline{\rubric{And then shall be said,}}
\smallskip

\medskip
\theGrace

\smallskip
%Scottish
\drop{May the souls of the faithful departed through the mercy of God rest in peace. \R Amen.}


\section{The Communion}
\pilcrow{When there is a special celebration of the Holy Communion on the day of the Burial, the Priest shall use the Collect appointed in this Order, or the Collect of Easter Even, and for the Epistle}

\subsection{\stylesubsec{}{The Collect.}{}} %Moved back from above.
\drop{O merciful God, the Father of our Lord Jesu Christ, who is the resurrection and the life; in whom whosoever believeth shall live, though he die; and whosoever liveth, and believeth in him, shall not die eternally; who also hath taught us by his holy Apostle Paul, not to be sorry, as men without hope, for them that sleep in him: We meekly beseech thee, O Father, to raise us from the death of sin unto the life of righteousness; that, when we shall depart this life, we may sleep in him, as our hope is this our \emph{brother} doth: and at the general resurrection in the last day both we, and this our \emph{brother} departed, receiving again our bodies, and rising again in thy most gracious favour, may with all thine elect Saints obtain eternal joy.
Grant this, O Lord God, by the means of our Advocate Jesus Christ: which with thee and the Holy Ghost, liveth and reigneth one God for ever. \R Amen.}

\subsection{\stylesubsec{}{The Epistle.}{1 Thessalonians 4.~13}}
\drop{I would not have you to be ignorant, brethren, concerning them which are asleep, that ye sorrow not, even as others which have no hope. For if we believe that Jesus died and rose again, even so them also which sleep in Jesus will God bring with him. For this we say unto you by the word of the Lord, that we which are alive and remain unto the coming of the Lord shall not precede them which are asleep. For the Lord himself shall descend from heaven with a shout, with the voice of the archangel, and with the trump of God: and the dead in Christ shall rise first: then we which are alive and remain shall be caught up together with them in the clouds, to meet the Lord in the air: and so shall we ever be with the Lord. Wherefore comfort one another with these words.}

\begin{leftbar}
\centerline{\rubric{Or this.}}
\centerline{2 Corinthians 4.~16, \emph{and part of chapter 5.}}
‘Though our outward man. . . swallowed up of life’
\end{leftbar}


\subsection{\stylesubsec{}{The Gospel.}{John 6.~37.}}
\drop{Jesus said, All that the Father giveth me shall come to me; and him that cometh to me I will in no wise cast out. For I came down from heaven, not to do mine own will, but the will of him that sent me. And this is the Father's will which hath sent me, that of all which he hath given me I should lose nothing, but should raise it up again at the last day. And this is the will of him that sent me, that every one which seeth the Son, and believeth on him, may have everlasting life: and I will raise him up at the last day.}

\begin{leftbar}
\centerline{\rubric{Or this.}}
\centerline{St.~John 5.~24}
‘Jesus said, Verily, verily, . . . the resurrection of judgement’.
\end{leftbar}

\medskip

%scottish 1929
{\centering\footnotesize\rubric{At Holy Communion in connexion with burials or at memorials of the departed, if the Agnus Dei be sung or said for ‘\emph{have mercy upon us}’, and ‘\emph{grant us thy peace}’, substitute ‘\emph{grant them rest}’, and ‘\emph{grant them rest eternal}’.}\par}




\section{The Burial}
\begin{leftbar}
%1928 proposed
\pilcrow{If the ground be not consecrated, the Priest on coming to the grave may say the prayer following.}

\drop{O God, the Father of our Lord Jesus Christ, vouchsafe, we beseech thee, to \grealtcross\ bless this grave to be the peaceful resting-place of the body of thy servant; through the same thy blessed Son, who is the resurrection and the life, and who liveth and reigneth with thee and the Holy Ghost; one God, world without end. \R Amen.}

Or

scottish, 1929
BENEDICTION OF A GRAVE IN UNCONSECRATED GROUND (also south african)

\pilcrow{When the Priest and people shall have come to the place the Priest shall say,}

\centerline{Let us pray.}
\drop{O Lord Jesu Christ, who wast laid in the new tomb of Joseph, and didst thereby sanctify the grave to be a bed of hope to thy people: Vouchsafe, we beseech thee, to \grealtcross\ bless, \grealtcross\ hallow, and \grealtcross\ consecrate this grave, that it may be a resting-place, peaceful and secure, for the body of thy servant which we are about to commit to thy gracious keeping; who art the resurrection and the life, and who livest and reignest with the Father and the Holy Ghost, one God, world without end. \R Amen.}
\end{leftbar}


\pilcrow{When they come to the Grave, while the Corpse is made ready to be laid into the earth, the Priest shall say, or the Priest and Clerks shall sing:}
\drop{Man that is born of a woman hath but a short time to live, and is full of misery. He cometh up, and is cut down, like a flower; he fleeth as it were a shadow, and never continueth in one stay.}

\drop{In the midst of life we are in death: of whom may we seek for succour, but of thee, O Lord, who for our sins art justly displeased?}

Yet, O Lord God most holy, O Lord most mighty, O holy and most merciful Saviour, deliver us not into the bitter pains of eternal death.

Thou knowest, Lord, the secrets of our hearts; shut not thy merciful ears to our prayer; but spare us, Lord most holy, O God most mighty, O holy and merciful Saviour, thou most worthy judge eternal, suffer us not, at our last hour, for any pains of death, to fall from thee.

\medskip

\pilcrow{Then, while the earth shall be cast upon the Body by some standing by, the Priest shall say,}

% 1549
\drop{I commend thy soul to God the Father Almighty, and thy
% 1662
% \drop{Forasmuch as it hath pleased Almighty God of his great mercy to take unto himself the soul of our dear brother here departed, we therefore commit his}
body to the ground; earth to earth, ashes to ashes, dust to dust; in sure and certain hope of the Resurrection to eternal life, through our Lord Jesus Christ; who shall change our vile body, that it may be like unto his glorious body, according to the mighty working, whereby he is able to subdue all things to himself.}

\pilcrow{When this Order is used at the cremation of the body, in place of the words ‘\emph{commit his body to the ground, earth to earth, ashes to ashes, dust to dust}’ shall be said the words ‘\emph{commit his body to be consumed by fire}’: and in this case it shall suffice to say one or more of the \emph{\scshape Prayers} at the burial of the ashes.}

\rubric{When this Order is used at the burial of the body after cremation, in place of the words ‘\emph{commit his body to the ground, earth to earth, ashes to ashes, dust to dust}’ shall be said the words ‘\emph{commit his ashes to the ground, earth to earth, dust to dust}’, or ‘\emph{commit his ashes to their resting-place}’.}

\centerline{\pilcrow{Then shall be said or sung,}}
\drop{I heard a voice from heaven, saying unto me, Write, From henceforth blessed are the dead which die in the Lord: even so saith the Spirit: for they rest from their labours.}


\centerline{Let us pray.}
\drop{We commend into thy hands of mercy, most merciful Father, the soul of this, our \emph{brother} departed, \emph{N.} And \emph{his} body we commit to the earth, beseeching thine infinite goodness to give us grace to live in thy fear and love, and to die in thy favour: that when the judgment shall come, which thou hast committed to thy well-beloved Son, both this our \emph{brother}, and we, may be found acceptable in thy sight, and receive that blessing, which thy well-beloved Son shall then pronounce to all that love and fear thee, saying, Come, ye blessed children of my Father, receive the kingdom prepared for you from the beginning of the world: Grant this, O merciful Father, for the honour of Jesu Christ, our only Savior, Mediator, and Advocate. \R Amen.}

\centerline{\rubric{This Prayer shall also be added.}}
\drop{Almighty God, we give thee hearty thanks for this thy \emph{servant}, whom thou hast delivered from the miseries of this wretched world, from the body of death and all temptation, and, as we trust, hast brought \emph{his} soul, which \emph{he} committed into thy holy hands, into sure consolation and rest: Grant, we beseech thee, that at the day of judgment \emph{his} soul, and all the souls of thy elect, departed out of this life, may with us, and we with them, fully receive thy promises, and be made perfect altogether, through the glorious resurrection of thy Son Jesus Christ our Lord. \R Amen.}

\medskip
1923/1928
\nowUntoTheKing

%scottish
\begin{leftbar}    
    \theGrace
\end{leftbar}

\fleuron

% \centerline{\rule{0.5\textwidth}{0.5pt}}

\chapter[The Burial of a Child]{\stylechapter{An Order which may be used for}{The Burial of a Child}{}}


% \section{An Order which may be Used for the Burial of A [Baptized] Child}
\pilcrow{The Minister and Clerks meeting the body at the entrance of the church-yard, and going before it either into the church or towards the grave, shall say or sing,}

\section{The Procession}
\drop{I am the resurrection and the life, saith the Lord: he that believeth in me, though he were dead, yet shall he live: and whosoever liveth and believeth in me shall never die.\scripture{St.~John xi.~25, 26.}}

\drop{I know that my Redeemer liveth, and that he shalt stand at the latter day upon the earth. 
% And though after my skin worms destroy this body, %KJV
% hath been thus destroyed, %adapted from rsv
% earth: and though this body be destroyed, yet shall I see God: %am1928
% yet in my flesh shall I see God: 
Whom I shall see for myself, and mine eyes shall behold, and not another. \scripture{Job 19.~25, 27.}}

\drop{We brought nothing into this world, and it is certain we can carry nothing out. The {\scshape Lord} gave, and the {\scshape Lord} hath taken away; blessed be the Name of the {\scshape Lord}.\scripture{1 Timothy.~6.~7. Job 1.~21.}}

\drop{He shall feed his flock like a shepherd: he shall gather the lambs with his arm, and carry them in his bosom. \scripture{Isaiah 40.~11}}


\section{The Service in Church}

\centerline{\pilcrow{After they are come into the church shall be read this Psalm,}}

\subsection[{Psalm 23}]{\stylesubsec{Psalm 23.}{Dominus regit me.}{}}
\drop{The {\scshape Lord} is my shepherd; \star\ therefore can I lack nothing.}

2\enspace He shall feed me in a green pasture, \star\ and lead me forth beside the waters of comfort.

3\enspace He shall convert my soul, \star\ and bring me forth in the paths of righteousness, for his Name’s sake.

4\enspace Yea, though I walk through the valley of the shadow of death, I will fear no evil; \star\ for thou art with me; thy rod and thy staff comfort me.

5\enspace Thou shalt prepare a table before me against them that trouble me; \star\ thou hast anointed my head with oil, and my cup shall be full.

6\enspace But thy loving-kindness and mercy shall follow me all the days of my life; \star\ and I will dwell in the house of the Lord for ever.

Glory be to the Father, and to the Son,\ \star\ and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be,\ \star\ world without end. Amen.

\medskip
\centerline{\pilcrow{Then shall follow this Lesson.}}

\centerline{St.~Mark 10.~13.}
\drop{They brought young children to him, that he should touch them: and his disciples rebuked those that brought them. But when Jesus saw it, he was much displeased, and said unto them, Suffer the little children to come unto me, and forbid them not; for of such is the kingdom of God. Verily I say unto you, Whosoever shall not receive the kingdom of God as a little child, he shall not enter therein. And he took them up in his arms, put his hands upon them, and blessed them.}
 
\medskip

\centerline{\pilcrow{Then the Minister shall say,}}
\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\medskip
\ourFather

\smallskip
\centerline{\pilcrow{The following Versicles and Responses may then be said:}}

\V Grant unto \emph{him} eternal rest; \R And let perpetual light shine upon \emph{him}.

\V We believe verily to see the goodness of the Lord; \R In the land of the living.

\V O Lord, hear our prayer; \R And let our cry come unto thee.

\pilcrow{Then shall be said one or both of the following prayers, the Minister first saying,}

\centerline{Let us pray.}
\drop{O Lord Jesu Christ, who didst take little children into thine arms and bless them:

[1928]Open thou our eyes, we beseech thee, to perceive that it is of thy goodness that thou hast taken this thy

[SA]Grant that in perfect confidence we may commit this 

child into the everlasting arms of thine infinite love; who livest and reignest with the Father and the Holy Spirit, ever one God, world without end. \R Amen.}

\smallskip
\drop{O God, whose ways are hidden and thy works most wonderful, who makest nothing in vain and lovest all that thou hast made: Comfort thou thy servants, whose hearts are sore smitten and oppressed; and grant that they may so love and serve thee in this life, that together with this thy child they may obtain the fulness of thy promises in the world to come; through Jesus Christ our Lord. \R Amen.}

\medskip
\theGrace
 
\section{The Burial}
\pilcrow{When they come to the grave, while the body is made ready to be laid into the earth, the Minister shall say, or the Minister and Clerks shall sing:}

\drop{Man that is born of a woman hath but a short time to live. He cometh up, and is cut down, like a flower; he fleeth as it were a shadow, and never continueth in one stay.}

% (these from SA)
\begin{leftbar}
While the child was yet alive I fasted and wept: for I said, Who can tell whether God will be gracious to me, that the child may live? But now he is dead, wherefore should I fast? Can I bring him back again? I shall go to him, but he shall not return to me.

A voice was heard in Ramah, lamentation, and bitter weeping; Rachel weeping for her children refused to be comforted for her children, because they were not.  Thus saith the Lord; Refrain thy voice from weeping, and thine eyes from tears: for thy work shall be rewarded, saith the Lord; and they shall come again from the land of the enemy.  And there is hope in thine end, saith the Lord, that thy children shall come again to their own border
\end{leftbar}


\medskip

\pilcrow{Then, while the earth shall be cast upon the body by some standing by, the Minister shall say,}
\drop{Forasmuch as it hath pleased Almighty God of his great mercy to take unto himself the soul of this child here departed, we therefore commit \emph{his} body to the ground; earth to earth, ashes to ashes, dust to dust; in sure and certain hope of the resurrection to eternal life, through our Lord Jesus Christ; who shall change the body of our low estate, that it may be like unto his glorious body, according to the mighty working, whereby he is able to subdue all things to himself.}

\centerline{\rubric{Or this.}}
\drop{We commend unto thy hands of mercy, most merciful Father, the soul of this thy child; and we commit \emph{his} body to the ground; earth to earth, ashes to ashes, dust to dust.  And we beseech thine infinite goodness to give us grace to live in thy fear and love, and to die in thy favour, that when the judgement shall come which thou hast comitted to thy well-beloved Son, both this child and we may be found acceptable in thy sight.  Grant this, O merciful Father, for the sake of Jesus Christ, our only Saviour, Mediator, and Advocate. \R Amen.}


\smallskip
\centerline{\pilcrow{Then shall be said or sung,}}

\drop{They shall hunger no more, neither thirst any more; neither shall the sun light on them, nor any heat. For the Lamb which is in the midst of the throne shall feed them, and shall lead them unto living fountains of waters: and God shall wipe away all tears from their eyes.}

\centerline{\pilcrow{Here shall be added by the Minister,}}

\nowUntoTheKing



\centerline{\rule{0.5\textwidth}{0.5pt}}
American 1928:
\drop{It is to be noted that this Office is appropriate to be used only for the faithful departed in Christ, provided that in any other case the Minister may, at his discretion, use such part of this Office, or such devotions taken from other parts of this Book, as may be fitting.}

 
\section{An Order which may be Used when the Prayer Book Service may not Be Used}
\pilcrow{The Priest, meeting the Corpse at the entrance to the Church-yard, and going before it towards the Grave, shall say,}

\pilcrow{The Minister shall await the body at the grave. When the body has been brought to the graveside, he shall say:}
 (Indian)
\subsection[{Psalm 130}]{\stylesubsec{Psalm 130.}{De profundis.}{}}

\drop{Out of the deep have I called unto thee, O {\scshape Lord}; \star\ Lord, hear my voice.}

2\enspace O let thine ears consider well \star\ the voice of my complaint.

3\enspace If thou, {\scshape Lord}, wilt be extreme to mark what is done amiss, \star\ O Lord, who may abide it?

4\enspace For there is mercy with thee; \star\ therefore shalt thou be feared.

5\enspace I look for the {\scshape Lord}; my soul doth wait for him; \star\ in his word is my trust.

6\enspace My soul fleeth unto the Lord \star\ before the morning watch, I say, before the morning watch.

7\enspace O Israel, trust in the {\scshape Lord}, for with the {\scshape Lord} there is mercy, \star\ and with him is plenteous redemption.

8\enspace And he shall redeem Israel \star\ from all his sins.

\medskip
\centerline{\pilcrow{When they come to the Grave shall be said,}}

% Domine, refugium. Psalm 90. 1-12.

% LORD, thou hast been our refuge : from one generation to another.
%     Before the mountains were brought forth, or ever the earth and the world were made : thou art God from everlasting, and world without end.
%     Thou turnest man to destruction : again thou sayest, Come again, ye children of men.
%     For a thousand years in thy sight are but as yesterday: seeing that is past as a watch in the night.
%     As soon as thou scatterest them, they are even as a sleep: and fade away suddenly like the grass.
%     In the morning it is green, and groweth up : but in the evening it is cut down, dried up, and withered.
%     For we consume away in thy displeasure: and are afraid at thy wrathful indignation.
%     Thou hast set our misdeeds before thee : and our secret sins in the light of thy countenance.
%     For when thou art angry all our days are gone : we bring our years to an end, as it were a tale that is told.
%     The days of our age are threescore years and ten; and though men be so strong, that they come to fourscore years : yet is their strength then but labour and sorrow; so soon passeth it away, and we are gone.
%     But who regardeth the power of thy wrath : for even thereafter as a man feareth, so is thy displeasure.
%     So teach us to number our days : that we may apply our hearts unto wisdom.

%     SA and Indian skip here.

\centerline{\pilcrow{Then may be read,}}
\centerline{St.~John 5.~25.}
\drop{Jesus said, Verily, verily, I say unto you, The hour is coming, and now is, when the dead shall hear the voice of the Son of God: and they that hear shall live. For as the Father hath life in himself; so hath he given to the Son to have life in himself; and hath given him authority to execute judgement also, because he is the Son of man. Marvel not at this: for the hour is coming, in the which all that are in the graves shall hear his voice, and shall come forth; they that hath done good, unto the resurrection of life; and they that hath done evil, unto the resurrection of judgement.}

% SA, Indian: Anthems here.
\drop{Man that is born of a woman hath but a short time to live, and is full of misery. He cometh up, and is cut down, like a flower; he fleeth as it were a shadow, and never continueth in one stay.}

\drop{In the midst of life we are in death: of whom may we seek for succour, but of thee, O Lord, who for our sins art justly displeased?}

Yet, O Lord God most holy, O Lord most mighty, O holy and most merciful Saviour, deliver us not into the bitter pains of eternal death.

Thou knowest, Lord, the secrets of our hearts; shut not thy merciful ears to our prayer; but spare us, Lord most holy, O God most mighty, O holy and merciful Saviour, thou most worthy judge eternal, suffer us not, at our last hour, for any pains of death, to fall from thee.

\medskip
\pilcrow{When the Corpse has been laid in the Grave the Priest shall say,}
SA, : 
\drop{We commit the body of our dear \emph{brother} to the ground; earth to earth, ashes to ashes, dust to dust; and we commend \emph{his} soul to the just and merciful judgement of him who alone hath perfect understanding, even Jesus Christ our Lord. \R Amen.}



\centerline{\rubric{Then the Minister shall say,}}
\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\medskip
\ourFather


Ant. Remember not, Lord, our offenses.. (from litany)

\V O Lord, deal not with us after our sins. \R Neither reward us after our iniquities.

\drop{We humbly beseech thee, O Father, mercifully to look upon our infirmities; and for the glory of thy Name turn from us all those evils that we most righteously have deserved; and grant, that in all our troubles we may put our whole trust and confidence in thy mercy, and evermore serve thee in holiness and pureness of living, to thy honour and glory; through our only Media tor and Advocate, Jesus Christ our Lord. Amen.}


Collect: Almighty God, give us grace that we may

\drop{Almighty God, the fountain of all wisdom, who knowest our necessities before we ask, and our ignorance in asking: We beseech thee to have compassion upon our infirmities; and those things, which for our unworthiness we dare not, and for our blindness we cannot ask, vouchsafe to give us, for the worthiness of thy Son Jesus Christ our Lord. \R Amen.}

SA: 
\drop{Almighty God, Father of all mercies and giver of all comfort: Deal graciously, we pray thee, with those who mourn, that casting every care on thee, they may know the consolation of thy love; through Jesus Christ our Lord. \R Amen.}

but indian: 
\drop{O God, whose ways are hidden and thy works past understanding, who makest nothing in vain and lovest all that thou hast made: Deal graciously, we pray thee, with those who mourn because of this bereavement, that casting every care on thee, they may know the consolation of thy love; through Jesus Christ our Saviour. \R Amen.}

Indian: Fountain of all wisdom
Sa: \drop{O Saviour of the world, who by thy Cross and precious Blood hast redeemed us, Save us and help us, we humbly beseech thee, O Lord.}

\medskip
\centerline{\rubric{And then shall be said,}}
\theGrace

 

 
\centerline{\rule{0.5\textwidth}{0.5pt}}

 
\section{An Order for the Burial of an Unbaptized Child}
\pilcrow{On the way to the Grave the following sentences may be said,}

\drop{God made not death: neither delighteth he when the living perish. He created man for incorruption: and made him an image of his own proper being.\scripture{Wisdom of Solomon 1.~13, 2.~23}}

\drop{Despair not then, seeing that thou art far from being able to love his creature more than he. For as his majesty is, so also is his mercy. \scripture{2 Esdras 8.~47; Ecclesiasticus 2.~18}}

\drop{He shall feed his flock like a shepherd, he shall gather the lambs with his arm, and carry them in his bosom. \scripture{Isaiah 40.~11}}

\drop{The Lord gave, and the Lord hath taken away; blessed be the name of the Lord.\scripture{Job 1.~21}}
(SA omits first)
(Indian is only the 4th and 3rd)

\medskip
\centerline{\pilcrow{When they come to the Grave shall be said,}}
% SA: Psalm 121


\subsection[{Psalm 23}]{\stylesubsec{Psalm 23.}{Dominus regit me.}{}}
\drop{The {\scshape Lord} is my shepherd; \star\ therefore can I lack nothing.}

2\enspace He shall feed me in a green pasture, \star\ and lead me forth beside the waters of comfort.

3\enspace He shall convert my soul, \star\ and bring me forth in the paths of righteousness, for his Name’s sake.

4\enspace Yea, though I walk through the valley of the shadow of death, I will fear no evil; \star\ for thou art with me; thy rod and thy staff comfort me.

5\enspace Thou shalt prepare a table before me against them that trouble me; \star\ thou hast anointed my head with oil, and my cup shall be full.

6\enspace But thy loving-kindness and mercy shall follow me all the days of my life; \star\ and I will dwell in the house of the Lord for ever.

(SA: Psalm 121)

\medskip
\centerline{\pilcrow{Then may be read one of the following,}}

Indian: Jeremiah 31:15
Thus saith the Lord: A voice is heard in Ramah...again to their own border.
Or baruch.


\centerline{St.~Matthew 18.~10.}
\drop{Take heed that ye despise not one of these little ones; for I say unto you, That in heaven their angels do always behold the face of my Father which is in heaven. For the Son of man is come to save that which was lost. How think ye? if a man have an hundred sheep, and one of them be gone astray, doth he not leave the ninety and nine, and goeth into the mountains, and seeketh that which is gone astray? And if so be that he find it, verily I say unto you, he rejoiceth more of that sheep, than of the ninety and nine which went not astray. Even so it is not the will of your Father which is in heaven, that one of these little ones should perish.}


\centerline{Baruch 4.~19.}
\drop{Go your way, my children, go your way: for I am left desolate. I have put off the clothing of peace, and put upon me the sack-cloth of my prayer: I will cry unto the Everlasting in my days. Be of good cheer, my children, cry unto the Lord, and he shall deliver you from the power and hand of the enemies. For my hope is in the Everlasting, that he will save you; and joy is come unto me from the Holy One, because of the mercy which shall soon come unto you from the Everlasting our Saviour. For I sent you out with mourning and weeping: but God will give you to me again with joy and gladness for ever.}

\medskip
\centerline{\pilcrow{As the body is being laid in the Grave shall be said,}}
\drop{Unto God’s loving mercy we commit this child, that he may grant \emph{him} a share in the unsearchable riches of the redemption wrought by his Son, our Lord and Saviour Jesus Christ. \R Amen.}


\medskip

\centerline{\rubric{Then the Minister shall say,}}
\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\medskip
\ourFather

\medskip

\centerline{Let us pray.}
\drop{O God, whose ways are hidden and thy works most wonderful, who makest nothing in vain and lovest all that thou hast made: Comfort thou thy servants, whose hearts are sore smitten and oppressed; and grant that they may so love and serve thee in this life, that together with this thy child they may obtain the fulness of thy promises in the world to come; through Jesus Christ our Lord. \R Amen.}

Indian: Numerous collects.
\medskip
\centerline{\rubric{And then shall be said,}}
\theGrace
