\chapter[The Churching of Women]{\stylechapter{The Thanksgiving of Women after Child-Birth\\ {\small commonly called}}{The Churching of Women}{}}

\pilcrow{The woman, at the usual time after her delivery, shall come into the Church decently apparelled, and there shall kneel down in some convenient place, as hath been accustomed, or as the Ordinary shall direct: And then the Priest shall say unto her,}

\drop{Forasmuch as it hath pleased Almighty God of his goodness to give you safe deliverance, and hath preserved you in the great danger of child-birth; you shall therefore give hearty thanks unto God, and say,}


\bigskip
% 1549: Levavi oculos. Psalm cxxi., 1559

% 1928 
\pilcrow{Then shall be said by both of them the following Psalm, the woman still kneeling.}

%The following psalm has some verses changed from the psalter.
%(Then shall the Priest say the 116th Psalm.)

\subsection[{Psalm 116}]{\stylesubsec{Psalm 116.}{Dilexi, quoniam.}{}}
\drop{I am wéll pléased \star\  that the {\scshape Lord} hath heard the  vóice of mý prayer;}

2\enspace That he hath inclined his éar  unto mé; \star\  therefore will I call upon him as  lóng as Í live.

3\enspace The snares of death compassed me  róund abóut, \star\  and the pains of hell gat  hóld upón me.

4\enspace I \emph{found} trouble and heaviness, and I \emph{called} upon the Náme  of the {\scshape Lórd}; \star\  O {\scshape Lord}, I beseech thee, delíver mý soul.

5\enspace Gracious is the  {\scshape Lórd}, and ríghteous; \star\ yea, our Gód is mérciful.

6\enspace The {\scshape Lord} presérveth the símple: \star\ I was in misery, ánd he hélped me.

7\enspace Turn again then unto thy rést,  O my sóul; \star\ for the {\scshape Lórd} hath rewárded thee.

8\enspace And why? thou hast delivered my  sóul from déath, \star\ mine eyes from tears, and my féet from fálling.

9\enspace I will wálk before the {\scshape Lórd} \star\  in the lánd  of the líving.

10\enspace I believed, and therefore will I speak; but Í  was sore tróubled: \star\  I said in my haste, All mén are líars.

11\enspace What reward shall I gíve unto the {\scshape Lórd} \star\  for all the benefits that hé hath done únto me?

12\enspace I will receive the cúp of salvátion, \star\  and call upon the Náme of thé {\scshape Lord}.

13\enspace I will pay my vows now in the presence of áll his péople: \star\   in the courts of the {\scshape Lord}’s house, even in the midst of thee, O Jerúsalem. Práise the {\scshape Lord}.

Glory be to the Father, and to the Son, \star\  and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be, \star\  world without end. Amen.

\medskip
\centerline{\rubric{Or,}}
\subsection[{Psalm 127}]{\stylesubsec{Psalm 127.}{Nisi Dominus.}{}}
\drop{Except the {\scshape Lord} búild the hóuse,  \star\  their labour is but lóst that búild it.}

2\enspace Except the {\scshape Lord} kéep the cíty,  \star\  the watchman wáketh bút in vain.

3\enspace It is but lost labour that ye haste to rise up early, and so late take rest, and eat the bréad of cárefulness;  \star\  for so he giveth hís belóved sleep.

4\enspace Lo, children and the frúit of the wómb,  \star\  are an heritage and gift that cómeth óf the {\scshape Lord}.

5\enspace Like as the arrows in the hánd of the gíant, * even so are the yöung chíldren.

6\enspace Happy is the man that hath his quíver fúll of them; * they shall not be ashamed when they speak with their én·emies ín the gate.

Glory be to the Father, and to the Son, \star\  and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be, \star\  world without end. Amen.

\medskip
\centerline{\pilcrow{Then the Priest shall say,}}


\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\smallskip
\ourFather

\V O Lord, save this woman thy servant; \R Who putteth her trust in thee.

\V Be thou to her a strong tower;  \R From the face of her enemy.

\V Lord, hear our prayer.  \R And let our cry come unto thee.

\centerline{Let us pray.}
\drop{O Almighty God, we give thee humble thanks for that thou hast vouchsafed to deliver this woman thy servant from the great pain and peril of child-birth; Grant, we beseech thee, most merciful Father, that she, through thy help, may both faithfully live, and walk according to thy will, in this life present; and also may be partaker of everlasting glory in the life to come; through Jesus Christ our Lord. \R Amen.}

% 1923 adds only the following; 1928 adds the following two.
\begin{leftbar}
\bigskip
\pilcrow{Then, if there be no Communion at the time of the Churching, shall the Priest say to the Woman,}
\drop{Unto God’s gracious mercy and protection we commit thee. The {\scshape Lord} bless thee, and keep thee. The {\scshape Lord} make his face to shine upon thee, and be gracious unto thee. The {\scshape Lord} lift up his countenance upon thee, and give thee peace, both now and evermore. \R Amen.}




\bigskip
\pilcrow{Prayers which may be used at the discretion of the Priest before the Blessing.}
\drop{O God, our heavenly Father, we thank thee and praise thy glorious name, that thou hast been pleased to bless this woman thy servant, and to bestow upon her the gift of a child: Grant, we beseech thee, most merciful Father, that she [with her husband] may diligently lead her child in the way of righteousness, to their own great blessing and the glory of thy name; through Jesus Christ our Lord. \R Amen.}

\medskip
\drop{O God, whose ways are hidden and thy works most wonderful, who makest nothing in vain, and lovest all that thou hast made: Comfort this thy servant whose heart is sore sitten and oppressed; and grant that she may so love and serve thee in this life, that she may obtain the fulness of thy promises in the world to come; through Jesus Christ our Lord. \R Amen.}
\end{leftbar}

\pilcrow{The woman, that cometh to give her thanks, must offer accustomed offerings; and, if there be a Communion, it is convenient that she receive the Holy Communion.}


\fleuron
