\chapter[A Commination]{\stylechapter{}{A Commination}{or Denouncing of God’s Anger and Judgements against Sinners\\ }}
{\centering\footnotesize\emph{With certain Prayers, to be used on the first Day of Lent, and at other times}\par}
%, as the Ordinary shall appoint

% "General Sentence" - warning of excommunication, 4 times a year.


\medskip
\pilcrow{After Morning Prayer, the Litany ended according to the accustomed manner, the Priest shall, in the Reading-Pew or Pulpit, say,}

\drop{Brethren, in the Primitive Church there was a godly discipline, that, at the beginning of Lent, such persons as stood convicted of notorious sin were put to open penance, and punished in this world, that their souls might be saved in the day of the Lord; and that others, admonished by their example, might be the more afraid to offend.}

Wherefore, lest by disuse of the said discipline God’s judgement upon sin be lightly regarded, it is thought good, that at this time (in the presence of you all) it should be declared that God will surely judge them that transgress his holy Commandments; and that ye, imploring his mercy, should answer \emph{Amen} in token that ye assent and submit to his righteous condemnation: To the intent that, being admonished of the great indignation of God against sinners, ye may the rather be moved to earnest and true repentance; and may walk more warily in these dangerous days; fleeing from such vices, for which ye affirm with your own mouths the judgement of God to be due.

% \drop{Cursed is the man that maketh any carved or molten image, to worship it.}

% \centerline{\rubric{And the people shall answer and say,}  Amen.}

% Cursed is he that curseth his father or mother. \R Amen.

% Cursed is he that removeth his neighbour’s landmark. \R Amen.

% Cursed is he that maketh the blind to go out of his way. \R Amen.

% Cursed is he that perverteth the judgement of the stranger, the fatherless, and widow. \R Amen.

% Cursed is he that smiteth his neighbour secretly. \R Amen.

% Cursed is he that lieth with his neighbour’s wife. \R Amen.

% Cursed is he that taketh reward to slay the innocent. \R Amen.

% Cursed is he that putteth his trust in man, and taketh man for his defence, and in his heart goeth from the Lord. \R Amen.

% Cursed are the unmerciful, fornicators, and adulterers, covetous persons, idolaters, slanderers, drunkards, and extortioners. \R Amen.

% \medskip
% \centerline{\rubric{Or,}}
% \medskip

\drop{The Lord our God is one Lord: them that serve other gods, God shall judge.}

\centerline{\pilcrow{And the people shall answer and say,}}
\centerline{\R Amen.  Lord have mercy upon us.}

II. Idolaters and all them that worship God’s creatures, God shall judge;

\centerline{\R Amen.  Lord have mercy upon us.}

III. Blasphemers and all them that take God’s name in vain, God shall judge; 

\centerline{\R Amen.  Lord have mercy upon us.}

IV. The Lord’s day is holy; them that profane it, God shall judge;

\centerline{\R Amen.  Lord have mercy upon us.}

V. Him that honoureth not his father or his mother, and them that are lawless or seditious, God shall judge;

\centerline{\R Amen.  Lord have mercy upon us.}

VI. Murderers and all them that are malicious or cruel, God shall judge;

\centerline{\R Amen.  Lord have mercy upon us.}

VII. Adulterers and fornicators and all unclean persons, God shall judge;

\centerline{\R Amen.  Lord have mercy upon us.}

VIII. Robbers and thieves and them that defraud, God shall judge;

\centerline{\R Amen.  Lord have mercy upon us.}

IX. False witnesses and all evil speakers, liars and slanderers, God shall judge;

\centerline{\R Amen.  Lord have mercy upon us.}

X. Covetous persons and extortioners and them that grind the faces of the poor, God shall judge;

\R Amen. Lord, have mercy upon us, and lay not these sins to our charge.


\centerline{\rubric{Minister.}}

% \drop{Now seeing that all they are accursed (as the prophet David beareth witness) who do err and go astray from the commandments of God; let us (remembering the dreadful judgement hanging over our heads, and always ready to fall upon us) return unto our Lord God, with all contrition and meekness of heart; bewailing and lamenting our sinful life, acknowledging and confessing our offences, and seeking to bring forth worthy fruits of penance. For now is the axe put unto the root of the trees, so that every tree that bringeth not forth good fruit is hewn down, and cast into the fire. It is a fearful thing to fall into the hands of the living God: he shall pour down rain upon the sinners, snares, fire and brimstone, storm and tempest; this shall be their portion to drink. For lo, the Lord is come out of his place to visit the wickedness of such as dwell upon the earth. But who may abide the day of his coming? Who shall be able to endure when he appeareth? His fan is in his hand, and he will purge his floor, and gather his wheat into the bam; but he will burn the chaff with unquenchable fire. The day of the Lord cometh as a thief in the night: and when men shall say, Peace, and all things are safe, then shall sudden destruction come upon them, as sorrow cometh upon a woman travailing with child, and they shall not escape. Then shall appear the wrath of God in the day of vengeance, which obstinate sinners, through the stubbornness of their heart, have heaped unto them, selves; which despised the goodness, patience, and long, sufferance of God, when he calleth them continually to repentance. Then shall they call upon me, (saith the Lord,) but I will not hear; they shall seek me early, but they shall not find me; and that, because they hated knowledge, and received not the fear of the Lord, but abhorred my counsel, and despised my correction. Then shall it be too late to knock when the door shall be shut; and too late to cry for mercy when it is the time of justice. O terrible voice of most just judgement, which shall be pronounced upon them, when it shall be said unto them, Go, ye cursed, into the fire everlasting, which is prepared for the devil and his angels. Therefore, brethren, take we heed betime, while the day of salvation lasteth; for the night cometh, when none can work. But let us, while we have the light, believe in the light, and walk as children of the light; that we be not cast into utter darkness, where is weeping and gnashing of teeth. Let us not abuse the goodness of God, who calleth us mercifully to amendment, and of his endless pity promiseth us forgiveness of that which is past, if with a perfect and true heart we return unto him. For though our sins be as red as scarlet, they shall be made white as snow; and though they be like purple, yet they shall be made white as wool. Turn ye (saith the Lord) from all your wickedness, and your sin shall not be your destruction: Cast away from you all your ungodliness that ye have done: Make you new hearts, and a new spirit: Wherefore will ye die, O ye house of Israel, seeing that I have no pleasure in the death of him that dieth, saith the Lord God? Tom ye then, and ye shall live. Although we have sinned, yet have we an Advocate with the Father, Jesus Christ the righteous; and he is the propitiation for our sins. For he was wounded for our offences, and smitten for our wickedness. Let us therefore return unto him, who is the merciful receiver of all true penitent sinners; assuring ourselves that he is ready to receive us, and most willing to pardon us, if we come unto him with faithful repentance; if we submit ourselves unto him, and from henceforth walk in his ways; if we will take his easy yoke, and light burden upon us, to follow him in lowliness, patience, and charity, and be ordered by the governance of his Holy Spirit; seeking always his glory, and serving him duly in our vocation with thanksgiving: This if we do, Christ will deliver us from the curse of the law, and from the extreme malediction which shall light upon them that shall be set on the left hand; and he will set us on his right hand, and give us the gracious benediction of his Father, commanding us to take possession of his glorious kingdom: Unto which he vouchsafe to bring us all, for his infinite mercy. Amen.}

% \centerline{Here shall follow The Prayers}
%from the proposed 1928:
% \section{An Alternative Commination}
% 


\drop{Now seeing that all they are condemned who do err and go astray from the commandments of God; let us (remembering the dreadful judgement hanging over our heads, and always ready to fall upon us) return unto our Lord God, with all contrition and meekness of heart; bewailing and lamenting our sinful life, acknowledging and confessing our offences, and seeking to bring forth worthy fruits of penance.  For it is a fearful thing to fall into the hands of the living God and to hear the terrible voice of his most just judgement which shall be pronounced upon obstinate sinners when it shall be said unto them, Go, ye cursed, into the fire everlasting, which is prepared for the devil and his angels.}
Therefore, brethren, take we heed betime, while the day of salvation lasteth.
% \begin{leftbar}
% For now is the axe put unto the root of the trees, so that every tree that bringeth not forth good fruit is hewn down, and cast into the fire. 
% It is a fearful thing to fall into the hands of the living God: he shall pour down rain upon the sinners, snares, fire and brimstone, storm and tempest; this shall be their portion to drink. For lo, the Lord is come out of his place to visit the wickedness of such as dwell upon the earth. But who may abide the day of his coming? Who shall be able to endure when he appeareth? His fan is in his hand, and he will purge his floor, and gather his wheat into the barn; but he will burn the chaff with unquenchable fire. The day of the Lord cometh as a thief in the night: and when men shall say, Peace, and all things are safe, then shall sudden destruction come upon them, as sorrow cometh upon a woman travailing with child, and they shall not escape. Then shall appear the wrath of God in the day of vengeance, which obstinate sinners, through the stubbornness of their heart, have heaped unto themselves; which despised the goodness, patience, and long-sufferance of God, when he calleth them continually to repentance. Then shall they call upon me, (saith the Lord,) but I will not hear; they shall seek me early, but they shall not find me; and that, because they hated knowledge, and received not the fear of the Lord, but abhorred my counsel, and despised my correction. Then shall it be too late to knock when the door shall be shut; and too late to cry for mercy when it is the time of justice. O terrible voice of most just judgement, which shall be pronounced upon them, when it shall be said unto them, Go, ye cursed, into the fire everlasting, which is prepared for the devil and his angels.
% Therefore, brethren, take we heed betime, while the day of salvation lasteth; for the night cometh, when none can work. But let us, while we have the light, believe in the light, and walk as children of the light; that we be not cast into utter darkness, where is weeping and gnashing of teeth. Let us not abuse the goodness of God, who calleth us mercifully to amendment, and of his endless pity promiseth us forgiveness of that which is past, if with a perfect and true heart we return unto him. For though our sins be as red as scarlet, they shall be made white as snow; and though they be like purple, yet they shall be made white as wool. Turn ye (saith the Lord) from all your wickedness, and your sin shall not be your destruction: Cast away from you all your ungodliness that ye have done: Make you new hearts, and a new spirit: Wherefore will ye die, O ye house of Israel, seeing that I have no pleasure in the death of him that dieth, saith the Lord God? Turn ye then, and ye shall live. 
% \end{leftbar}
Although we have sinned, yet have we an Advocate with the Father, Jesus Christ the righteous; and he is the propitiation for our sins. For he was wounded for our offences, and smitten for our wickedness. Let us therefore return unto him, who is the merciful receiver of all true penitent sinners; assuring ourselves that he is ready to receive us, and most willing to pardon us, if we come unto him with faithful repentance; if we submit ourselves unto him, and from henceforth walk in his ways; if we will take his easy yoke, and light burden upon us, to follow him in lowliness, patience, and charity, and be ordered by the governance of his Holy Spirit; seeking always his glory, and serving him duly in our vocation with thanksgiving. This if we do, Christ will deliver us from the extreme malediction which shall light upon them that shall be set on the left hand; and he will set us on his right hand, and give us the gracious benediction of his Father, commanding us to take possession of his glorious kingdom: Unto which he vouchsafe to bring us all, for his infinite mercy. Amen.


\medskip

% The Supplement
% to the Indian Book of Common Prayer
\begin{leftbar}
\stylesec{The Form for the Blessing of Ashes}{on}{Ash Wednesday}
% \section{The Blessing of Ashes}
% \pilcrow{Before the Lord’s Supper ashes prepared from the palms blessed the previous Palm Sunday, or other suitable ashes, may be blessed as follows:}

\pilcrow{Ashes prepared from the palms blessed the previous Palm Sunday, or other suitable ashes, shall be placed in a vessel near the holy Table; and the Priest, standing at the Epistle side, shall say,}
\smallskip
\V The Lord be with you.  \R And with thy spirit.

\centerline{Let us pray.}
\drop{O God, our faithful Creator, who wouldest not the death of a sinner, but rather that he should turn from his wickedness, and live: Look with mercy upon the frailty of our mortal nature; and of thy goodness vouchsafe to \grealtcross\ bless these ashes which are now to be set upon our heads as a token of humility and of sorrow for our sins. We acknowledge that we are but dust and ashes, and that, by reason of our offences, unto dust we shall return; yet we beseech thy mercy to grant the forgiveness of all our sins and the pardon which thou hast promised to all who truly repent and believe in thy Son; who with thee and the Holy Spirit, liveth and reigneth, one God, world without end. \R Amen.}

\centerline{\rubric{Here may the ashes be sprinkled and censed.}}

\bigskip
\pilcrow{Then shall the Priest put ashes on his own forehead, or if there be another Priest present, he shall put the ashes on the officiant’s forehead; after which the people shall kneel at the Communion rail and the Priest shall put the ashes on their foreheads. During the imposition \emph{Psalm 25} may be said or sung, or some suitable Lenten hymn may be sung.}

\medskip

\centerline{\rubric{The Priest shall say to each person, as the ashes are imposed:}}

Remember, O man, that dust thou art, and unto dust shalt thou return.

\centerline{\rubric{Or}}

Remember that thou art a sinner, and repent.
\end{leftbar}

\medskip

\section{The Prayers}
\pilcrow{Then shall they all kneel upon their knees, and the Priest and Clerks kneeling (in the place where they are accustomed to say the Litany) shall say this Psalm.}

\subsection{\stylesubsec{Psalm 51.}{Miserere mei, Deus.}{}}
\drop{Have mercy upon me, O God, after thy great goodness;\ \star\ according to the multitude of thy mercies do away mine offences.}

2\enspace Wash me throughly from my wickedness,\ \star\ and cleanse me from my sin.

3\enspace For I knowledge my faults,\ \star\ and my sin is ever before me.

4\enspace Against thee only have I sinned, and done this evil in thy sight;\ \star\ that thou mightest be justified in thy saying, and clear when thou art judged.

5\enspace Behold, I was shapen in wickedness,\ \star\ and in sin hath my mother conceived me.

6\enspace But lo, thou requirest truth in the inward parts,\ \star\ and shalt make me to understand wisdom secretly.

7\enspace Thou shalt purge me with hyssop, and i shall be clean;\ \star\ thou shalt wash me, and I shall be whiter than snow.

8\enspace Thou shalt make me hear of joy and gladness,\ \star\ that the bones which thou hast broken may rejoice.

9\enspace Turn thy face from my sins,\ \star\ and put out all my misdeeds.

10\enspace Make me a clean heart, O God,\ \star\ and renew a right spirit within me.

11\enspace Cast me not away from thy presence,\ \star\ and take not thy holy Spirit from me.

12\enspace O give me the comfort of thy help again,\ \star\ and stablish me with thý free Spirit.

13\enspace Then shall I teach thy ways únto the wicked,\ \star\ and sinners shall be converted únto thee.

14\enspace Deliver me from blood-guiltiness, O God, thou that art the God of my health;\ \star\ and my tongue shall sing of thy righteousness.

15\enspace Thou shalt open my lips, O Lord,\ \star\ and my mouth shall shew thy praise.

16\enspace For thou desirest no sacrifice, else would I give it thee;\ \star\ but thou delightest not in burnt-offerings.

17\enspace The sacrifice of God is a troubled spirit:\ \star\ a broken and contrite heart, O God, shalt thou not despise.

18\enspace O be favourable and gracious únto Sion;\ \star\ build thou the walls of Hierúsalem.

19\enspace Then shalt thou be pleased with the sacrifice of righteousness, with the burnt-offerings and oblations;\ \star\ then shall they offer young bullocks upon thine altar.

Glory be to the Father, and to the Son,\ \star\ and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be,\ \star\ world without end. Amen.

\medskip

\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\smallskip
\ourFather


\V O Lord, save thy servants; \R That put their trust in thee.

\V Send unto them help from above; \R And evermore mightily defend them.

\V Help us, O God our Saviour. \R And for the glory of thy Name deliver us; be merciful to us sinners, for thy name’s sake.

\V O Lord, hear our prayer; \R And let our cry come unto thee.

\centerline{Let us pray.}
\drop{O Lord, we beseech thee, mercifully hear our prayers, and spare all those who confess their sins unto thee; that they, whose consciences by sin are accused, by thy merciful pardon may be \grecross\ absolved; through Christ our Lord. \R Amen.}

\smallskip

\drop{O most mighty God, and merciful Father, who hast compassion upon all men, and hatest nothing that thou hast made; who wouldest not the death of a sinner, but that he should rather turn from his sin, and be saved: Mercifully forgive us our trespasses; receive and comfort us, who are grieved and wearied with the burden of our sins. Thy property is always to have mercy; to thee only it appertaineth to forgive sins. Spare us therefore, good Lord, spare thy people, whom thou hast redeemed; enter not into judgement with thy servants, who are vile earth, and miserable sinners; but so turn thine anger from us, who meekly acknowledge our vileness, and truly repent us of our faults, and so make haste to help us in this world, that we may ever live with thee in the world to come; through Jesus Christ our Lord. \R Amen.}


\medskip

\pilcrow{Then shall the people say this Anthem that followeth, after the Minister.}
\drop{Turn thou us, O good Lord, and so shall we be turned. Be favourable, O Lord, \Be favourable to thy people, \ Who turn to thee in weeping, fasting, and praying. For thou art a merciful God, \ Full of compassion, long-suffering, and of great pity. Thou sparest when we deserve punishment, \ And in thy wrath thinkest upon mercy. Spare thy people, good Lord, spare them, \ And let not thine heritage be brought to confusion. Hear us, O Lord, for thy mercy is great, \ And after the multitude of thy mercies look upon us; \ Through the merits and mediation of thy blessed Son, Jesus Christ our Lord. Amen.}

\medskip

\centerline{\pilcrow{Then the Minister alone shall say,}}
\drop{The {\scshape Lord} \cross bless us, and keep us: the {\scshape Lord} make his face to shine upon us, and be gracious unto us: the {\scshape Lord} lift up the light of his countenance upon us, and give us peace, now and for evermore. \R Amen.}

\fleuron