
\chapter[Confirmation]{\stylechapter{}{The Order of Confirmation}{Or Laying on of Hands upon Those That Are Baptized and Come to Years of Discretion\\ }}



\pilcrow{Upon the day appointed, all that are to be then confirmed, being placed, and standing in order, before the Bishop; he (or some other Minister appointed by him) shall read this Preface following.}

% 1549 rubric, then become a preface.
% \drop{To the end that Confirmation may be ministered to the more edifying of such as shall receive it, the Church hath thought good to order, That none hereafter shall be Confirmed, but such as can say the Creed, the Lord's Prayer, and the Ten Commandments; and can also answer to such other Questions, as in the short Catechism are contained; which order is very convenient to be observed; to the end, that children, being now come to the years of discretion, and having learned what their Godfathers and Godmothers promised for them in Baptism, they may themselves, with their own mouth and consent, openly before the Church, ratify and confirm the same; and also promise, that by the grace of God they will evermore endeavour themselves faithfully to observe such things, as they, by their own confession, have assented unto.}
%En28, Sc29

\drop{Dearly beloved in the Lord, in ministering Confirmation the Church doth follow the example of the Apostles of Christ. For in the eighth chapter of the Acts of the Apostles we thus read:—}

They therefore that were scattered abroad went about preaching the word. And Philip went down to the city of Samaria, and proclaimed unto them the Christ. When they believed Philip preaching good tidings concerning the kingdom of God and the name of Jesus Christ, they were baptized, both men and women. Now when the Apostles which were at Jerusalem heard that Samaria had received the word of God, they sent unto them Peter and John; who, when they were come down, prayed for them, that they might receive the Holy Ghost: for as yet he was fallen upon none of them; only they had been baptized into the Name of the Lord Jesus. Then laid they their hands on them, and they received the Holy Ghost.

The Scripture here teacheth us that a special gift of the Holy Spirit is bestowed through laying on of hands with prayer. And forasmuch as this gift cometh from God alone, let us that are here present pray to Almighty God, that he will strengthen with his Holy Spirit in Confirmation those who in Baptism were made his children.

You, then, who are to be confirmed must now declare before this congregation that you are stedfastly purposed, with the help of this gift, to lead your life in the faith of Christ and in obedience to God’s will and commandments; and must openly acknowledge yourselves bound to fulfil the Christian duties to which your Baptism hath pledged you.

\section{The Renewal of Baptismal Vows}

\centerline{\pilcrow{Then shall the Bishop say,}}

\lettrine{D}{\emph{o}} \emph{ye} here, in the presence of God, and of this congregation, renew the solemn promise and vow that was made in \emph{your} name at \emph{your} Baptism; ratifying and confirming the same in \emph{your} own persons, and acknowledging \emph{yourselves} bound to believe, and to do, all those things, which \emph{your} Godfathers and Godmothers then undertook for \emph{you}?

\centerline{\rubric{And every one shall audibly answer,}}
\centerline{I do.}

\medskip

\centerline{\rubric{Or else the Bishop shall say,}}
%En1928
% \drop{Do ye here, in the presence of God, and of this congregation, renounce the devil and all his works, the pomps and vanity of this wicked world, and all the sinful lusts of the flesh, so that ye will not follow nor be led by them?}
% Scottish / 28Proposed
\lettrine{D}{\emph{o}} \emph{ye} here, in the presence of God, and of this congregation, renounce the devil and all his works, the vain pomp and glory of the world, with all covetous desires of the same, and the sinful desires of the flesh, so that \emph{ye will} not follow, nor be led by them?

\R I do.

\lettrine{D}{\emph{o}} \emph{ye} believe all the Articles of the Christian Faith as contained in the Apostles’ Creed?

\R I do.

\lettrine{D}{\emph{o}} \emph{ye} promise that \emph{ye will} endeavour to keep God’s holy will and commandments, and to walk in the same all the days of \emph{your} life?

% \drop{Will ye endeavour to keep God’s holy will and commandments, and to walk in the same all the days of your life?}

\R I do.

\medskip

\section{The Confirmation}

\centerline{\pilcrow{The Bishop.}}

\drop{Our \grecross\ help is in the Name of the Lord;  \R Who hath made heaven and earth.}

\V Blessed be the Name of the Lord;  \R Henceforth, world without end.

\V Lord, hear our prayers.  \R And let our cry come unto thee. %1662

\V The Lord be with you.  \R And with thy spirit. %1549

\centerline{Let us pray.}
\drop{Almighty and everliving God, who hast vouchsafed to regenerate these thy servants by Water and the Holy Ghost, and hast given unto them forgiveness of all their sins: Strengthen them, we beseech thee, O Lord, with the Holy Ghost the Comforter, and daily increase in them thy manifold gifts of grace;

The Spirit of wisdom and understanding;

The Spirit of counsel and ghostly strength;

The Spirit of knowledge and true godliness;

and fill them, O Lord, with the Spirit of thy holy fear, now and for ever. \R Amen.}

\smallskip

% 1549 & 1929
\drop{Sign them, O Lord, and mark them to be thine for ever by the virtue of the Holy Cross; mercifully confirm them with the inward unction of the Holy Ghost, that they may attain unto everlasting life.  \R Amen.}

\medskip

\pilcrow{Then all of them in order kneeling before the Bishop, he shall lay his hand upon the head of every one severally, saying}

%1912, 1549
\lettrine{\emph{N.}}{ I} sign thee with the sign of the \grealtcross\ Cross\footnote{\rubric{Here the Bishop shall sign the person with the sign of the Cross on the forehead with the holy Chrism.}} and I lay my hands [\rubric{or} hand] upon thee, in the Name of the Father, and of the Son, and of the Holy Ghost.

% Sarum
% and I confirm thee with the chrism of salvation.  In the name...
\smallskip

% The following from 1552, etc.
\drop{Defend, O Lord, this thy Child [\rubric{or} this thy Servant \rubric{or} Handmaiden] with thy heavenly grace, that \emph{he} may continue thine for ever; and daily increase in thy Holy Spirit more and more, until \emph{he} come unto thy everlasting kingdom. \R Amen.}

\medskip
\centerline{\pilcrow{Then shall the Bishop say,}}

\V Peace be with you.  \R And with thy spirit. %1549; 1552 on it was "The Lord be with you.

\centerline{\rubric{And (all kneeling down) the Bishop shall add,}}

\centerline{Let us pray.}

\begin{leftbar} % added in 1662
    \ourFather

\centerline{\rubric{And this Collect.}}
\end{leftbar}


\drop{Almighty and everliving God, who makest us both to will and to do those things that be good and acceptable unto thy divine Majesty; We make our humble supplications unto thee for these thy servants, upon whom (after the example of thy holy Apostles) we have now laid our hands, to certify them (by this sign) of thy favour and gracious goodness towards them. Let thy fatherly hand, we beseech thee, ever be over them; let thy Holy Spirit ever be with them; and so lead them in the knowledge and obedience of thy Word, that in the end they may obtain everlasting life; through our Lord Jesus Christ, who with thee and the Holy Ghost liveth and reigneth, ever one God, world without end. \R Amen.}

\section{The Conclusion}
\drop{O almighty Lord, and everlasting God, vouchsafe, we beseech thee, to direct, sanctify, and govern, both our hearts and bodies, in the ways of thy laws, and in the works of thy commandments; that, through thy most mighty protection both here and ever, we may be preserved in body and soul; through our Lord and Saviour Jesus Christ. \R Amen.}

\medskip
\centerline{\pilcrow{Then the Bishop shall bless them, saying thus,}}
% En1928 only
\begin{leftbar}
    
\drop{Go forth into the world in peace; be of good courage; hold fast that which is good; render to no man evil for evil; strengthen the fainthearted; support the weak; help the afflicted; honour all men; love and serve the Lord, rejoicing in the power of the Holy Spirit.}
\end{leftbar}

And the Blessing of God Almighty, the \grealtcross\ Father, the \grealtcross\ Son, and the Holy \grealtcross\ Ghost, be upon you, and remain with you for ever. \R Amen.

\medskip

\pilcrow{And there shall none be admitted to the holy Communion, until such time as he be confirmed, or be ready and desirous to be confirmed.}

\fleuron