\chapter{The Order for Evensong}
\subsubsection{Daily Throughout the Year.}

\pilcrow{Here all standing up, the Priest shall say,}

\drop{O God, \cross make speed to save \emph{me}.  \R O Lord, make haste to help \emph{me}.}

\V Glory be to the Father, and to the Son, and to the Holy Ghost;  \R As it was in the beginning, is now, and ever shall be, world without end. Amen.

\V Praise ye the Lord.  \R The Lord's Name be praised.

\medskip

\pilcrow{Then shall be said or sung the Psalms in order as they be appointed. Then a Lesson of the Old Testament, as is appointed. And after that \emph{Magnificat} (or the Song of the blessed Virgin Mary) in English, as followeth.}

\subsection{\stylesubsec{The Song of the Blessed Virgin Mary}{Magnificat.}{St.~Luke i.~46.}}

\drop{My \cross soul doth magnify the Lord : and my spirit hath rejoiced in God my Saviour.}

2 For he hath regarded : the lowliness of his handmaiden.

3 For behold, from henceforth : all generations shall call me blessed.

4 For he that is mighty hath magnified me : and holy is his Name.

5 And his mercy is on them that fear him : throughout all generations.

6 He hath shewed strength with his arm : he hath scattered the proud in the imagination of their hearts.

7 He hath put down the mighty from their seat : and hath exalted the humble and meek.

8 He hath filled the hungry with good things : and the rich he hath sent empty away.

9 He remembering his mercy hath holpen his servant Israel : as he promised to our forefathers, Abraham and his seed, for ever.

Glory be to the Father, and to the Son : and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be : world without end. Amen.

\medskip

\pilcrow{Then a Lesson of the New Testament, as it is appointed. And after that \emph{Nunc dimittis} (or the Song of Symeon) in English, as followeth.}
\subsection{\stylesubsec{The Song of Symeon}{Nunc dimittis.}{St.~Luke ij.~29.}}

\drop{Lord, \cross now lettest thou thy servant depart in peace : according to thy word.}

2 For mine eyes have seen : thy salvation,

3 Which thou hast prepared : before the face of all people;

4 To be a light to lighten the Gentiles : and to be the glory of thy people Israel.

Glory be to the Father, and to the Son : and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be : world without end. Amen.

\medskip

\pilcrow{Then shall be sung or said the Apostles' Creed, by the Minister and the people standing.}
\drop{I believe in God the Father Almighty, Maker of heaven and earth :}

And in Jesus Christ his only Son our Lord: Who was conceived by the Holy Ghost, Born of the Virgin Mary: Suffered under Pontius Pilate, Was crucified, dead, and buried: He descended into hell; The third day he rose again from the dead: He ascended into heaven, And sitteth on the right hand of God the Father Almighty: From thence he shall come to judge the quick and the dead.

I believe in the Holy Ghost: The holy Catholick Church; The Communion of Saints: The Forgiveness of sins: The Resurrection of the body, And the Life everlasting. Amen.

\medskip

\pilcrow{And after that, these Prayers following, all devoutly kneeling,}

\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\pilcrow{Then the Minister, Clerks, and people shall say the Lord's Prayer with a loud voice.}
\drop{Our Father, which art in heaven, Hallowed be thy Name. Thy kingdom come. Thy will be done in earth, As it is in heaven. Give us this day our daily bread. And forgive us our trespasses, As we forgive them that trespass against us. And lead us not into temptation, But deliver us from evil. Amen.}

\pilcrow{Then the Priest standing up shall say,}

\V O Lord, shew thy mercy upon us.  \R And grant us thy salvation.

\V O Lord, save the \emph{State}.  \R And mercifully hear us when we call upon thee.

\V Endue thy Ministers with righteousness.  \R And make thy chosen people joyful.

\V  O Lord, save thy people.  \R And bless thine inheritance.

\V Give peace in our time, O Lord.  \R Because there is none other that fighteth for us, but only thou, O God.

\V O God, make clean our hearts within us.  \R And take not thy Holy Spirit from us.

\V The Lord be with you.  \R And with thy spirit.

\centerline{Let us pray.}


\pilcrow{Then shall follow three Collects ; the first of the day; The second for Peace ; The third for Aid against all Perils, as hereafter followeth : which two last Collects shall be daily said at Evening Prayer without alteration.}

\subsection{\stylesubsec{The second Collect, }{for Peace.}{}}
\drop{O God, from whom all holy desires, all good counsels, and all just works do proceed; Give unto thy servants that peace which the world cannot give; that both our hearts may be set to obey thy commandments, and also that by thee, we, being defended from the fear of our enemies, may pass our time in rest and quietness; through the merits of Jesus Christ our Saviour. \R Amen.}

\subsection{\stylesubsec{The third Collect, }{for  Aid against all Perils.}{}}
\drop{Lighten our darkness, we beseech thee, O Lord; and by thy great mercy defend us from all perils and dangers of this night; for the love of thy only Son, our Saviour, Jesus Christ. \R Amen.}

\V The Lord be with you.  \R And with thy spirit.

\V Let us bless the Lord.  \R Thanks be to God.

\medskip

\pilcrow{In Quires and Places where they sing here followeth the Anthem.}

\pilcrow{Here may follow any of the Occasional Prayers and Thanksgivings, as need may require, ending with one of the Conclusions.}

\subsubsection{Here endeth the Order of Evening Prayer throughout the Year.}

\fleuron