\chapter{The Hymnal}

\def\hymn{\obeylines\hymna}
\def\hymnverses{\obeylines\hymna}
\newcommand{\hymna}[1]{
    \directlua{printhymn([[#1]])}
}



\section{Sunday, during Summer}
¶ At Mattins\hfill Nocte surgentes

\hymnverses{
Now, from the slumbers of the night arising,
Chaunt we the holy psalmody of David,
Hymns to our Master, with our best endeavour,
Sweetly intoning.

2. So may our Monarch pitifully hear us,
That we may merit with his Saints to enter
Mansions eternal, therewithal possessing
Joy beatific.

3. This he vouchsafe us, God for ever blessed,
Father eternal, Son, and Holy Spirit,
Whose is the glory, which through all creation
Ever resoundeth. Amen.
}

\rubric{Sundays} \V The Lord is high above all péople.  \R And his glory above the héavens.

\rubric{Weekdays} \V Let thy merciful kindness, O Lord, be upon us.  \R As we do put our trust in thee.

\medskip
¶ At Lauds\hfill Ecce jam noctis

\hymnverses{
Lo! the dim shadows of the night are waning;
Lightsome and blushing, dawn of day returneth;
Fervent in spirit, to the mighty Father
Pray we devoutly.

2. So shall our Maker, of his great compassion,
Banish all sickness, kindly health bestowing;
And may he grant us, of a Father’s goodness,
Mansions in heaven.


3. This he vouchsafe us, God for ever blessed,
Father eternal, Son, and holy Spirit,
Whose is the glory, which through all creation
Ever resoundeth. Amen.
}

\rubric{Sundays} \V The Lord is Kíng.  \R He hath put on glorious apparel, allelúya.

\rubric{Weekdays} \V Have I not thought upon thee when I was waking?  \R Because thou hast been my helper.

¶ At Evensong\hfill Lucis Creator optime

\hymnverses{
O blest Creator of the light,  
Who mak’st the day with radiance bright,  
And o’er the forming world didst call  
The light from chaos first of all:

2. Whose wisdom join’d in meet array
The morn and eve, and named them Day:
Night comes with all its darkling fears,
Regard thy people’s prayers and tears.

3. Lest sunk in sin, and ’whelm’d with strife,
They lose the gift of endless life:
While thinking but the thoughts of time,
They weave new chains of woe and crime.

4. But grant them grace that they may strain
The heav’nly gate and prize to gain:
Each harmful lure aside to cast.
And purge away each error past.

5. O Father, that we ask be done,
Through Jesus Christ, thine only Son:
Who, with the Holy Ghost and thee,
Shall live and reign eternally. Amen.
}

\V Let my prayer be set forth, O Lord.  \R In thy sight as the incense.


\rubric{Monday.} My soul doth magnify \star\ the Lord.

\rubric{Tuesday.} My spirit hath rejoiced \star\ in God my Saviour.

\rubric{Wednesday.} O Lord my God, \star\ thou hast regarded my lowliness.

\rubric{Thursday.} He hath put down \star\ the mighty from their seat; and hath exalted the humble and meek that confess his Christ.

\rubric{Friday.} God hath holpen \star\ his servant Israel, as he promised Abraham and his seed; and he hath exalted the humble for ever.


¶ At Compline.\hfill Te lucis ante terminum

\hymnverses{
To thee, before the close of day,  
Creator of the world, we pray,  
That, with thy wonted favour, thou  
Wouldst be our guard and keeper now.

2. From all ill dreams defend our eyes,
From nightly fears and fantasies:
Tread under foot our ghostly foe,
That no pollution we may know.

3. O Father, that we ask be done.
Through Jesus Christ, thine only Son:
Who, with the holy Ghost and thee,
Shall live and reign eternally. Amen.
}

\V Keep us, O Lórd.  \R As the apple of an eye, hide us under the shadow of thy wíngs.

\bigskip
\section{A Virgin}
¶ At First Evensong \& Mattins.\hfill Virginis Proles

\hymnverses{
Child of a Virgin,   Maker of thy Mother,
Born of a Maiden,  as of Maid conceived,
While we a Virgin’s   triumphs are rehearsing,
Hear our petition.

*2. She, thine own maiden, double blessing winneth,
Striving to vanquish all her nature’s weakness.
E’en by that weakness o’er a world of bloodshed
Victory gaining.

*3. Death and its terrors undismay’d beholding,
Death’s cruel handmaid, torture, she despiseth;
Shedding her life-blood, meet is she to enter
Holiest heaven.

4. God ever-loving, as for us she pleadeth,
Pity our failings, all our sins forgiving:
Thus shall re-echo pure and heart-felt praises
Unto thine honour.

5. Praise to the Father, to the Sole-begotten,
And the blest Spirit, with the twain co-equal,
One only Godhead, who throughout the ages
Reigneth for ever. Amen.
}

¶ Note that if the Virgin be not a Martyr, verses 2 and 3 are ommitted.
¶ For Many Virgins, the hymn and \V is {Ihesu, corona virginum}, as at Lauds.

\rubric{Evensong.} \V Full of grace are thy líps. \R Therefore God hath blessed thee for éver. .

\rubric{Mattins.} \V God shall give her the help of his cóuntenance.  \R God is in the midst of her, therefore shall she not be remóved. .

\bigskip


¶ At Lauds \& Second Evensong.\hfill Ihesu corona virginum

\hymnverses{
Jesu, the Virgins’ Crown, do thou
Accept us as in prayer we bow:
Born of that Virgin, whom alone
The Mother and the Maid we own.

2. Among the lilies thou dost feed,
By Virgin quires accompanied—
With glory deck’d, the spotless brides
Whose bridal gifts thy love provides.

3. They, wheresoe’er thy footsteps bend,
With hymns and praises still attend:
In blessed troops they follow thee,
With dance, and song, and melody.

4. We pray thee therefore to bestow
Upon our senses here below
Thy grace, that so we may endure
From taint of all corruption pure.

Ordinary Doxology
5. All laud to God the Father be,
All praise, eternal Son, to thee:
All glory, as is ever meet,
To God the holy Paraclete. Amen.
}

\V The virgins that be her fellows shall bear her company.  \R And shall be brought unto thee.

¶ For Many Virgins, the hymn and \V is {Rex gloriose martyrum}, as at Lauds for Many Martyrs.




\section{The First Sunday in Advent}

Office. Ad te levavi.
\lettrine{U}{nto} thee, O Lord, lift I up my soul; O my God, in thee have I trusted, let me not • be confóunded; * neither let mine enemies triumph over me; for all they that look for thee shall • not be ashamed. \rubric{Ps.} Shew me • thý ways, Ó Lord * and • teach me thý paths.

\rubric{Grail.} For all they • that lóok for thee * shall not be a•shámed, Ó Lord.  \V Make known to me • thy wáys, O Lord * and • téach me thy paths.  \rubric{Alleluya.} \V Shew us thy mercy, • O Lord * and grant us • thy salvátion.   \rubric{Offertory.} Unto thee, O Lord, lift I up my soul; O my God, in thee have I trusted, let me • not be confóunded * neither let mine enemies triumph over me; for all they that look for thee shall • not be ashamed.  \rubric{Communion.} The Lord shall shew • lóving-kíndness * and our land • shall give her increase.

\section{The Second Sunday of Advent}
Office. Populus Sion.
\lettrine{O}{ people} of Sion, behold, the Lord is nigh at hand to re•déem the nátions: * and in the gladness of your heart the Lord shall cause his glori•ous voice to be heard.  \rubric{Ps.} Hear, O thou • Shépherd of Ísrael: * thou that leadest • Jóseph like a sheep.

\rubric{Grail.} Out of Sion hath God • appéared: * in • pérfect béauty.  \V Gather my saints toge•ther únto me: * those that have made a covenant with • mé with sácrifice.  \rubric{Alleluya.} \V For the powers of heaven shall be • shaken: * and then shall they see the Son of Man coming in a cloud with power • and great glóry. \rubric{Offertory.} Wilt not thou turn again, O God, and quicken us; that thy peo•ple may rejóice in thee: * shew us thy mercy, O Lord; and grant • us thy salvation.  \rubric{Communion.} Jerusalem, haste • thee, and stánd on high: * and behold the joy and gladness, which cometh unto thee • from God thy Saviour.


\section{The Third Sunday of Advent}
Office. Gaudete.
\lettrine{R}{ejoice} ye in the Lord alway, and again I say, rejoice ye;   Let your moderation be known unto all men; the • Lord is át hand. * Be careful for nothing, nor troubled; but in all things, by prayer and supplication, with thanksgiving, let your requests be • made known unto God.  \rubric{Ps.} And the peace of God, which passeth all • únderstánding * shall • keep your héarts and minds.

\rubric{Grail.} Shew thyself, O Lord, thou that sittest upon • the Chérubim: * stir • úp thy stréngth and come.  \V Hear, O thou Shepherd • of Ísrael: * thou that leadest • Jóseph líke a sheep.  \rubric{Alleluya.} \V Stir up thy strength, • O Lord: * and • come and hélp us. \rubric{Offertory.} Lord, thou art become gracious unto thy land; thou hast turned away the capti•vity of Jácob; * thou hast forgiven the of•fence of thy people.  \rubric{Communion.} Say to them that are • of a féarful heart: * Be strong, fear not; behold, your God • will come and save you.


\section{The Fourth Sunday of Advent}
¶ Should the Fourth Sunday fall on {Christmas Eve}, the {Office, Grail, Alleluya, Offertory,} and {Communion} will be those of {Christmas Eve}, below.
Office. Memento nostri.

\lettrine{R}{emember} us, O Lord, with the favour that thou bearest unto thy people; O visit us with • thy salvátion: * that we, beholding the felicity of thy chosen, may rejoice in the gladness of thy people, and may glory with • thine inheritance. \rubric{Ps.} We have sinned • wíth our fáthers: * we have done amiss, • and dealt wíckedly.

\rubric{Grail.} The Lord is nigh unto all them that call • upón him; * yea, all such as call up•ón him fáithfully. \V My mouth shall speak the praise • of thé Lord; * and let all flesh give thanks un•tó his hóly name. \rubric{Alleluia.} \V Come, O Lord, and tar•ry not: * forgive the misdeeds • of thy péople. \rubric{Offertory.} Be strong, fear not; behold, our God will • come with a récompense: * – he • will come and save us. \rubric{Communion.} Behold, a Virgin shall con•ceive and béar a Son, * and his name shall be call•ed Emmanuel.


\section{Christmas Eve}
Office. Hodie scietis.
\lettrine{T}{o-day} shall ye know that the Lord will come to deliver you * and at sunrise shall ye behold his glory. \rubric{Ps.} The earth is the Lord’s, and all that is therein * the compass of the world, and they that dwell therein.

\rubric{Grail.} To-day shall ye know that the Lord will come to deliver you; * and at sunrise shall ye behold his glory. \V Hear, O thou Shepherd of Israel, thou that leadest Joseph like a sheep: * show thyself also, thou that sittest upon the Cherubim, before Ephraim, Benjamin, and Manasses. \rubric{Alleluya.} \V On the morrow the iniquity of the earth shall be blotted out: * and the Saviour of the world shall reign over us. \rubric{Offertory.} Lift up your heads, O ye gates, and be ye lift up, ye everlasting doors: * and the King of Glory shall come in. \rubric{Communion.} The glory of the Lord shall be revealed: * and all flesh shall see the salvation of our God.

\section{The Fifteenth Sunday after Trinity}
\centerline{Office. \emph{Inclina, Domine.}}
\drop{Bów dówn, O Lórd, thine éar to • me, and héar me: \star\ O my Gód, sáve thy sérvant, that trústeth in thée; have mércy upón me, O Lórd, for I have cálled • dáily upón thee. \rubric{Ps.} Cómfort the • sóul of thy sérvant: \star\ for únto thée, O Lórd, do I • líft up my sóul.}

\rubric{Grail.} It is bétter to • trüst in the Lórd: \star\ than to pút any • cónfidence in mán. \V It is bétter to • trüst in the Lórd: \star\ than to pút any cónfi•dence in prínces. \rubric{Allelúya.} \V My héart is réady, O Gód, my héart is • réady: \star\ I will síng, yéa, I will práise thee, with the bést • mémber thát I have. \rubric{Offertory.} I wáited pátiently for the Lórd, and he • inclíned únto me: \star\ hé heard my cálling, and hath pút a néw sóng in my móuth, éven a thanksgív•ing únto our Gód. \rubric{Communion.} Whóso éateth my Flésh, and • drínketh my Blóod: \star\ dwélleth in mé, and I • in hím, sáith the Lórd.
