Touching for the King's Evil.
At the Healing.
\drop{Prevent us, O Lord, in all our doings with thy most gracious favour, and further us with thy continual help, that in all our works begun, continued, and ended in thee, we may glorify thy holy name, and finally by thy mercy obtain everlasting life: through Jesus Christ our Lord. \R Amen.}

The Gospel (for Ascension-day)   S. Mark xvi. 14-20.

Let us pray.
Lord have mercy upon us.
    Christ have mercy upon us.
Lord have mercy upon us.

\drop{Our Father, which art in heaven, Hallowed be thy Name. Thy kingdom come. Thy will be done in earth, As it is in heaven. Give us this day our daily bread. And forgive us our trespasses, As we forgive those who trespass against us. And lead us not into temptation, But deliver us from evil. Amen.}

¶ Then shall the infirm persons, one by one, be presented to the Queen upon their knees; and as every one is presented and while the Queen is laying her hands upon them, and putting the gold about their necks, the Chaplain that officiates, turning himself to her Majesty, shall say these words following:
God give a blessing to this work; and grant that these sick persons on whom the Queen lays her hands may recover, through Jesus Christ our Lord.

¶ After all have been presented, the Chaplain shall say,
O Lord, save thy servants;
Answer. That put their trust in thee.
Minister. Send unto them help from above.
Answer. And evermore mightily defend them.
Minister. Help us, O God our Saviour.
Answer. And for the glory of thy Name deliver us; be merciful to us sinners, for thy Name's sake.
Minister. O Lord, hear our prayer.
Answer. And let our cry come unto thee.

Let us pray.
O almighty God, who art the Giver of all health, and the aid of them that seek to thee for succour, we call upon thee for thy help and goodness mercifully to be showed upon these thy servants, that they being healed of their infirmities may give thanks unto thee in thy holy Church, through Jesus Christ our Lord. Amen.

¶ Then the Chaplain, standing with his face towards them that come to be healed, shall say,
The almighty Lord, who is a most strong tower to all them that put their trust in him, to whom all things in heaven, in earth, and under the earth, do bow and obey, be now and evermore thy defence; and make thee know and feel, that there is none other Name under heaven given to man, in whom, and through whom, thou mayest receive health and salvation, but only the Name of our Lord Jesus Christ. Amen.

The grace + of our Lord Jesus Christ, and the love of God, and the fellowship of the Holy Ghost, be with us all evermore. Amen.

