
\chapter{A Devotion}

{\centering\rubric{which may be said by the Priest and people immediately before the celebration of the Holy Communion.}\par}


\pilcrow{The Priest, standing at God’s Board, shall say with the Ministers and the people, all kneeling, as follows.}

\drop{In the name of the Father, \grecross\ and of the Son, and of the Holy Ghost. Amen.}

\rubric{Anthem.} I will go unto the altar of God, \star\ even unto the God of my joy and gladness.

\subsection{\stylesubsec{Psalm 43.}{Judica me, Deus.}{}}

\drop{Give sentence with me, O God, and defend my cause against the ungódly péople; \star\  O deliver me from the decéitful and wícked man.}

2\enspace For thou art the God of my strength, why hast thou pút me fróm thee? \star\  and why go I so heavily, while the enemý opprésseth me?

3\enspace O send out thy light and thy truth, that théy may léad me, \star\  and bring me unto thy holy hill, and tó thy dwélling.

4\enspace And that I may go unto the altar of God, even unto the God of my jóy and gládness; \star\  and upon the harp will I give thanks unto thée, O Gód, my God.

5\enspace Why art thou so héavy, O my sóul? \star\  and why art thou so disquietéd withín me?

6\enspace O pút thy trust in Gód; \star\  for I will yet give him thanks, which is the help of my cóuntenance, ánd my God.

Glory be to the Father, and to the Son, \star\  and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be, \star\  world without end. Amen.

\rubric{Anthem.} I will go unto the altar of God, even unto the God of my joy and gladness.

\medskip

\V Our help standeth in the name of the Lord;  \R Who hath made heaven and earth.

\medskip

\rubric{If desired, the Confession and Absolution may be said here, and ommitted in the Order of Communion.}

\medskip

\V Wilt thou not turn again and quicken us;  \R That thy people may rejoice in thee?

\V O Lord, show thy mercy upon us;  \R And grant us thy salvation.

\V O Lord, hear our prayer;  \V And let our cry come unto thee.

\V The Lord be with you; \V And with thy spirit.

\centerline{Let us pray.}

\pilcrow{Then shall the Priest proceed with the celebration of the Holy Communion}

\chapter[Holy Communion]{\stylechapter{The Order for the Administration of the Lord's Supper, or}{Holy Communion}{Commonly Called the Mass.}}


\pilcrow{It is an ancient and laudable custom of the Church to receive this Holy Sacrament fasting. Yet for the avoidance of all scruple it is hereby declared that such preparation may be used or not used, according to every man’s conscience in the sight of God.}

\medskip
% \pilcrow{So many as intend to be partakers of the holy Communion shall signify their names to the Curate at least some time the day before.}

% And if any of those be an open and notorious evil liver, or have done any wrong to his neighbours by word or deed, so that the Congregation be thereby offended; the Curate having knowledge thereof, shall call him and advertise him, that in any wise he presume not to come to the Lord's Table, until he have openly declared himself to have truly repented and amended his former naughty life, that the Congregation may thereby be satisfied, which before were offended; and that he have recompensed the parties, to whom he hath done wrong; or at least declare himself to be in full purpose so to do, as soon as he conveniently may.

% The same order shall the Curate use with those betwixt whom he perceiveth malice and hatred to reign; not suffering them to be partakers of the Lord's Table. until he know them to be reconciled. And if one of the parties so at variance be content to forgive from the bottom of his heart all that the other hath trespassed against him, and to make amends for that he himself hath offended; and the other party will not be persuaded to a godly unity, but remain still in his frowardness and malice: the Minister in that case ought to admit the penitent person to the holy Communion, and not him that is obstinate.

% And when any person is warned as in the two precedent paragraphs not to come to the Lord's Table, the Minister shall inform him that the case shall be laid before the Bishop of the Diocese without delay, and that pending the judgement of the Bishop he cannot be admitted to the Holy Communion.

% And on every such occasion as is set forth in the three precedent paragraphs, the Minister shall immediately give an account of the case to the Bishop and shall await his directions. And if occasion require, the Ordinary shall proceed against the offending person according to the Canon.

% ¶ The Priest shall say the Service following in a distinct and audible voice.

\pilcrow{The Holy Table, having at the Communion time a fair white linen cloth upon it, with other decent furniture meet for the high Mysteries there to be celebrated, shall stand at the uppermost part of the Chancel or Church. And the Priest, standing at the Holy Table, shall say the Lord’s Prayer, with the collect following for due preparation, the people kneeling.} % 1912 Scott (this only) 


% 1549: Upon the date and at the tyme appoincted for the ministracion of the holy Communion, the Priest that shal execute the holy ministery, shall put upon hym the vesture appoincted for that ministracion, that is to saye: a white Albe plain, with a vestement or Cope. And where there be many Priestes, or Decons, there so many shalbe ready to helpe the Priest, in the ministracion, as shalbee requisite: And shall have upon them lykewise the vestures appointed for their ministery, that is to saye, Albes with tunacles. Then shall the Clerkes syng in Englishe for the office, or Introite, (as they call it,) a Psalme appointed for that daie.

% 1549: The Priest standing humbly afore the middes of the Altar, shall saie the Lordes praier, with this Collect.

\section{The Introduction}

\ourFather

\subsection{\stylesubsec{}{The Collect.}{}}
\drop{Almighty God, unto whom all hearts be open, all desires known, and from whom no secrets are hid; Cleanse the thoughts of our hearts by the inspiration of thy Holy Spirit, that we may perfectly love thee, and worthily magnify thy holy Name; through Christ our Lord. \R Amen.}

\bigskip
%location from green book; text from am28

\pilcrow{Here may be sung a Hymn or an Anthem.}
% he saie a Psalme appointed for the introite: whiche Psalme ended the Priest shall saye, or els the Clerkes shal syng,}

\bigskip

% iii. Lorde have mercie upon us.
% iii. Christ have mercie upon us.
% iii. Lorde have mercie upon us.

% Then the Prieste standyng at Goddes borde shall begin,

% Glory be to God on high.


\pilcrow{Then shall the Priest, turning to the people, rehearse distinctly all the TEN COMMANDMENTS; and the people, still kneeling, shall after every Commandment ask God mercy for their transgression of every duty therein (either according to the letter or according to the spiritual import thereof) for the time past, and grace to keep the same for the time to come, as followeth.} %1923, includes 1912 Scott

\smallskip

\centerline{God spake these words, and said,}
\drop{I am the Lord thy God: Thou shalt have none other gods but me.}

\R Lord, have mercy upon us, and incline our hearts to keep this law.

II. Thou shalt not make to thyself any graven image, nor the likeness of any thing that is in heaven above, or in the earth beneath, or in the water under the earth. Thou shalt not bow down to them, nor worship them.%: for I the Lord thy God am a jealous God, and visit the sins of the fathers upon the children unto the third and fourth generation of them that hate me, and shew mercy unto thousands in them that love me, and keep my commandments.

\R Lord, have mercy upon us, and incline our hearts to keep this law.

III. Thou shalt not take the Name of the Lord thy God in vain%: for the Lord will not hold him guiltless, that taketh his Name in vain.
.

\R Lord, have mercy upon us, and incline our hearts to keep this law.

IV. Remember that thou keep holy the Sabbath-day. Six days shalt thou labour, and do all that thou hast to do; but the seventh day is the Sabbath of the Lord thy God.% [In it thou shalt do no manner of work, thou, and thy son, and thy daughter, thy man-servant, and thy maid-servant, thy cattle, and the stranger that is within thy gates. For in six days the Lord made heaven and earth, the sea, and ail that in them is, and rested the seventh day: wherefore the Lord blessed the seventh day, and hallowed it.]

\R Lord, have mercy upon us, and incline our hearts to keep this law.

V. Honour thy father and thy mother.%; that thy days may be long in the land which the Lord thy God giveth thee.


\R Lord, have mercy upon us, and incline our hearts to keep this law.

VI. Thou shalt do no murder.

\R Lord, have mercy upon us, and incline our hearts to keep this law.

VII. Thou shalt not commit adultery.
    
\R Lord, have mercy upon us, and incline our hearts to keep this law.

VIII. Thou shalt not steal.

\R Lord, have mercy upon us, and incline our hearts to keep this law.

IX. Thou shalt not bear false witness% against thy neighbour.
.

\R Lord, have mercy upon us, and incline our hearts to keep this law.

X. Thou shalt not covet% thy neighbour's house, thou shalt not covet thy neighbour's wife, nor his servant, nor his maid, nor his ox, nor his ass, nor any thing that is his.
.

\R Lord, have mercy upon us, and write all these thy laws in our hearts, we beseech thee.

\medskip

%scottish 1912
\subsubsection{Or the Priest may rehearse, instead of the Ten Commandments, the Summary of the Law as followeth:}

%Irish, Canadian, American
\centerline{Hear what our Lord Jesus Christ saith.}
 
% \drop{Hear, O Israel, the Lord our God is one Lord: and thou shalt love the Lord thy God with all thy heart, and with all thy soul, and with all thy mind, with all thy strength: This is the first commandment; And the second is like, namely this, Thou shalt love thy neighbour as thyself: there is none other commandment greater than these.}\scripture{St.~Mark xij.~29.}
% On these two commandments hang all the Law and the Prophets.

\drop{Thou shalt love the Lord thy God with all thy heart, and with all thy soul, and with all thy mind.  This is the first and great commandment.  And the second is like unto it, Thou shalt love thy neighbour as thyself.  On these two commandments hang all the law and the prophets.}\scripture{St.~Matthew xxij.~37.}

\R Lord, have mercy upon us, and write these thy laws in our hearts, we beseech thee.


%\rubric{Or else, instead of the Ten Commandments or the Summary of the Law, may be sung or said on week-days, not being Great Festivals, as followeth:}

\medskip

\pilcrow{Here, if the Decalogue hath been omitted, shall be said or sung,}

\smallskip

\centerline{Lord, have mercy upon us. \rubric{iij.}}
\centerline{Christ, have mercy upon us. \rubric{iij.}}
\centerline{Lord, have mercy upon us. \rubric{iij.}}
\subsubsection{or}

\centerline{Kyrie eleison. \rubric{iij.}}
\centerline{Christe, eleison. \rubric{iij.}}
\centerline{Kyrie eleison. \rubric{iij.}}

\bigskip

\pilcrow{Then, on Sundays and on Feast days (except in Advent and from Septuagesima to Palm Sunday inclusive), shall be said or sung as follows:}

\drop{Glory be to God on high, and in earth peace, good will towards men. We praise thee, we bless thee, \footnote{Bow} we worship thee, we glorify thee, we give thanks to thee for thy great glory, O Lord God, heavenly King, God the Father Almighty.}
%\emph{and to thee, O God, the only begotten Son Jesu Christ; and to thee, O God, the Holy Ghost. - Scottish}

O Lord, the only-begotten Son Jesu Christ; O Lord God, Lamb of God, Son of the Father, that takest away the sins of the world, have mercy upon us. 
%\emph{Thou that takest away the sins of the world, have mercy upon us. - English} 
Thou that takest away the sins of the world, * receive our prayer. Thou that sittest at the right hand of God the Father, have mercy upon us.

For thou only art holy; thou only art the Lord; thou only, * O {Jesu} Christ, with the Holy Ghost, art most high \grecross\ in the glory of God the Father. Amen.

% Rubric from "The Orange Book" - 1923
{\footnotesize\rubric{This hymn may be omitted here, and sung instead at the end of this Order after the Thanksgiving after Communion.}\par}

\bigskip
%1549, 1912 Scottish:
\pilcrow {Then the priest shall turn him to the people and say,}
\V The Lord be with you. \R And with thy spirit.

% \pilcrow{Then shall follow one of these two Collects for the Queen, the Priest standing as before, and saying,}
\centerline{Let us pray.}
% \drop{Almighty God, whose kingdom is everlasting, and power infinite: Have mercy upon the whole Church; and so rule the heart of thy chosen servant ELIZABETH, our Queen and Governor, that she (knowing whose minister she is) may above all things seek thy honour and glory; and that we, and all her subjects (duly considering whose authority she hath) may faithfully serve, honour, and humbly obey her, in thee, and for thee, according to thy blessed Word and ordinance; through Jesus Christ our Lord, who with thee and the Holy Ghost liveth and reigneth, ever one God, world without end. \R Amen.}

% Or,

% \drop{Almighty and everlasting God, we are taught by thy Holy Word, that the hearts of Kings are in thy rule and governance, and that thou dost dispose and turn them as it seemeth best to thy godly wisdom: We humbly beseech thee so to dispose and govern the heart of ELIZABETH thy Servant, our Queen and Governor, that, in all her thoughts, words, and works, she may ever seek thy honour and glory, and study to preserve thy people committed to her charge, in wealth, peace, and godliness: Grant this, O merciful Father, for thy dear Son's sake, Jesus Christ our Lord. \R Amen.}

% \pilcrow{Then the Priest, turning to the Holy Table, shall say the \emph{Collect} or Collects.}
%En28
\medskip
\pilcrow{And turning to the Holy Table he shall say the Collect of the Day. Other collects contained in this Book or authorized by the Bishop may follow.}

\medskip

\section{The Ministry of the Word}

\pilcrow{Immediately thereafter he that readeth the Epistle shall say,}
The Epistle [\rubric{or,} The portion of Scripture appointed for the Epistle] is written in the --- chapter of --- beginning at the --- verse. 
\rubric{And the Epistle ended, he shall say,} Here endeth the Epistle.

\medskip
% From the Am28
\pilcrow{Here may be sung a Hymn or an Anthem.}
\medskip

\pilcrow{Then the Deacon or Priest that readeth the Gospel (the people all standing up) shall say,}
\V The Lord be with you. \R And with thy Spirit. 

\drop{The \grealtcross\ Holy Gospel is written in the — chapter of — beginning at the — verse. \R Glory \grecross\ be to thee, O Lord.}
% \pilcrow{He that readeth the Epistle or the Gospel shall so stand and turn himself as he may best be heard of the people.}
% \rubric{And after the Gospel the people may in like manner say or sing,}
% Thanks be to thee, O Lord, for this thy glorious Gospel. %1912: 

\centerline{\rubric{The Gospel ended, there may be said,}}
\centerline{Praise be to thee, O Christ.}
%Am28, Green, En28
% ¶ And after the Gospel may be said,
% Praise be to thee, O Christ.

\bigskip

\pilcrow{Then shall be sung or said the Creed following, the people still standing as before: except that at the discretion of the Minister it may be omitted on any day not being a Sunday or a Holy-day.} %Eng28

\drop{I believe in one God the Father Almighty, Maker of heaven and earth, And of all things visible and invisible:}

And in one Lord Jesus Christ, the only-begotten son of God, Begotten of his Father before all worlds, God of God, Light of Light, Very God of very God, Begotten, not made, Being of one substance with the Father, By whom all things were made: Who for us men, and for our salvation came down from heaven, * And was incarnate by the Holy Ghost of the Virgin Mary, * And was made man, * And was crucified also for us under Pontius Pilate. He suffered and was buried, And the third day he rose again according to the Scriptures, And ascended into heaven, And sitteth on the right hand of the Father. And he shall come again with glory to judge both the quick and the dead: Whose kingdom shall have no end.

And I believe in the Holy Ghost, The Lord and giver of life, Who proceedeth from the Father and the Son, Who with the Father and the Son together is worshipped and glorified, Who spake by the Prophets. And I believe One {Holy} Catholick and Apostolick Church. I acknowledge one Baptism for the remission of sins. And I look for the Resurrection of the dead, * And the life of the world to come. Amen.

\bigskip
\pilcrow{Then the Curate shall declare unto the people what Holy-days, or Fasting-days, are in the week following to be observed. And then also (if occasion be) shall notice be given of the Holy Communion, or of other services; Banns of matrimony may be published, and Briefs, Citations, and Excommunications shall be read, and Bidding of Prayers may be made. And nothing shall be proclaimed or published in the Church during the time of Divine Service, but by the Minister: nor by him any thing, but what is prescribed in the rules of this Book, or enjoined by the %Queen, or by the 
Ordinary of the place.}% adapted toward En1928


\smallskip


\pilcrow{Then may follow the Sermon, or one of the Homilies already set forth, or hereafter to be set forth, by authority.}


\smallskip


%scottish 1912
\pilcrow{When the Priest giveth warning of the Holy Communion he may, at his discretion, use the first or the second of the Exhortations appended to this Liturgy.}


\smallskip


% \pilcrow{The third Exhortation appended to this Liturgy may be used at the discretion of the Presbyter before the Offertory, the people standing.}
\medskip

\pilcrow{At the time of the celebration of the Holy Communion, the communicants being conveniently placed for the receiving of the Holy Sacrament, the Priest may say this Exhortation.}

\drop{Dearly beloved in the Lord, ye that mind to come to the holy Communion of the Body and Blood of our Saviour Christ, must consider how Saint Paul exhorteth all persons diligently to try and examine themselves, before they presume to eat of that Bread, and drink of that Cup. For as the benefit is great, if with a true penitent heart and lively faith we receive that holy Sacrament; \emph{(for then we spiritually eat the flesh of Christ, and drink his blood; then we dwell in Christ, and Christ in us; we are one with Christ, and Christ with us;)} so is the danger great, if we receive the same unworthily.}
\rubric{For then we are guilty of the Body and Blood of Christ our Saviour; we eat and drink our own damnation, not considering the Lord's Body; we kindle God's wrath against us; we provoke him to plague us with divers diseases, and sundry kinds of death.}

Judge therefore yourselves, brethren, that ye be not judged of the Lord; repent you truly for your sins past; have a lively and stedfast faith in Christ our Saviour; amend your lives, and be in perfect charity with all men; so shall ye be meet partakers of those holy mysteries. And above all things ye must give most humble and hearty thanks to God, the Father, the Son, and the Holy Ghost, for the redemption of the world by the death and passion of our Saviour Christ, both God and man; who did humble himself, even to the death upon the Cross, for us, miserable sinners, who lay in darkness and the shadow of death; that he might make us the children of God, and exalt us to everlasting life.

And to the end that we should alway remember the exceeding great love of our Master, and only Saviour, Jesus Christ, thus dying for us, and the innumerable benefits which by his precious blood-shedding he hath obtained to us; he hath instituted and ordained holy mysteries, as pledges of his love, and for a continual remembrance of his death, to our great and endless comfort. To him therefore, with the Father and the Holy Ghost, let us give (as we are most bounden) continual thanks; submitting ourselves wholly to his holy will and pleasure, and studying to serve him in true holiness and righteousness all the days of our life. Amen.


\bigskip

\section{The Offertory}
\label{offertory}
% From the "Green Book"
\pilcrow{Then shall the Priest return to the Lord's Table, and begin the \emph{Offertory}. The Priest shall say, or the Clerks shall sing, one of these Sentences following, or some other convenient sentence taken out of Holy Scripture. A Hymn may follow.}
% \pilcrow{1549 : ¶ Then shall folowe for the Offertory, one or mo, of these Sentences of holy scripture, to bee song whiles the People doo offer, or els one of them to bee saied by the minister, immediatly afore the offeryng.}

% \pilcrow{Then shall the Priest return to the Lord's Table, and begin the Offertory, saying one or more of these Sentences following, as he thinketh most convenient in his discretion.}

\drop{Let your light so shine before men, that they may see your good works, and glorify your Father which is in heaven.}\scripture{St.~Matthew v.~16.}

Lay not up for yourselves treasure upon the earth; where the rust and moth doth corrupt, and where thieves break through and steal: but lay up for yourselves treasures in heaven; where neither rust nor moth doth corrupt, and where thieves do not break through and steal.\scripture{St.~Matthew vj.~19.}

Whatsoever ye would that men should do unto you, even so do unto them; for this is the Law and the Prophets.\scripture{St.~Matthew vij.~12.}

Not every one that saith unto me, Lord, Lord, shall enter into the kingdom of heaven; but he that doeth the will of my Father which is in heaven.\scripture{St.~Matthew vij.~21.}

Remember the words of the Lord Jesus, how he said, It is more blessed to give than to receive.\scripture{Acts xx.~35.}

Godliness is great riches, if a man be content with that he hath: for we brought nothing into the world, neither may we carry any thing out.\scripture{1 Timothy vj.~6.}

Be merciful after thy power. If thou hast much, give plenteously: if thou hast little, do thy diligence gladly to give of that little: for so gatherest thou thyself a good reward in the day of necessity.\scripture{Tobit iv.~8.}

%\emph{Zacchæus stood forth, and said unto the Lord, Behold, Lord, the half of my goods I give to the poor; and if I have done any wrong to any man, I restore four-fold.\scripture{St.~Luke xix.}} Omitted in En28

%\emph{Who goeth a warfare at any time of his own cost? Who planteth a vineyard, and eateth not of the fruit thereof? Or who feedeth a flock, and eateth not of the milk of the flock?\scripture{1 Corithians ix}} Omitted in En28

If we have sown unto you spiritual things, is it a great matter if we shall reap your worldly things?\scripture{1 Corithians ix.~11.}

Do ye not know, that they who minister about holy things live of the sacrifice; and they who wait at the altar are partakers with the altar? Even so hath the Lord also ordained, that they who preach the Gospel should live of the Gospel.\scripture{1 Corithians ix.~13.}

He that soweth little shall reap little; and he that soweth plenteously shall reap plenteously. Let every man do according as he is disposed in his heart, not grudging, or of necessity; for God loveth a cheerful giver.\scripture{2 Corithians ix.~6.}

Let him that is taught in the Word minister unto him that teacheth, in all good things. Be not deceived, God is not mocked: for whatsoever a man soweth that shall he reap.\scripture{Galatians vj.~6.}

While we have time, let us do good unto all men; and specially unto them that are of the household of faith.\scripture{Galatians vj.~10.}

God is not unrighteous, that he will forget your works, and labour that proceedeth of love; which love ye have shewed for his Name’s sake, who have ministered unto the saints, and yet do minister.\scripture{Hebrews vj.~10.}

Lift up your eyes and look upon the fields; for they are white already to harvest.\scripture{St.~John iv.~35.}

Charge them who are rich in this world, that they be ready to give, and glad to distribute; laying up in store for themselves a good foundation against the time to come, that they may attain eternal life.\scripture{1 Timothy vj.~17.}

Whoso hath this world’s good, and seeth his brother have need, and shutteth up his compassion from him, how dwelleth the love of God in him?\scripture{1 St.~John iij.~17}

Blessed be the man that provideth for the sick and needy: the Lord shall deliver him in the time of trouble.\scripture{Psalm xli.~1.}

To do good, and to distribute, forget not; for with such sacrifices God is pleased.\scripture{Hebrews xiij.~16.}

Offer unto God thanksgiving, and pay thy vows unto the most Highest.\scripture{Psalm l.~14.}

I will offer in his dwelling an oblation with great gladness: I will sing and speak praises unto the Lord.\scripture{Psalm xxvij.~6.}

Melchizedek king of Salem brought forth bread and wine; and he was the priest of the most high God.\scripture{Genesis xiv.~18.}

%\emph{Give alms of thy goods, and never turn thy face from any poor man; and then the face of the Lord shall not be turned away from thee. \scripture{Tobit iv.}} Omitted in En28

%\emph{He that hath pity upon the poor lendeth unto the Lord: and look, what he layeth out, it shall be paid him again.\scripture{Proverbs xix.}} Omitted in En28

% \pilcrow{1549 : Where there be Clerkes, thei shall syng one, or many of the sentences above written, accordyng to the length and shortenesse of the tyme, that the people be offeryng.}

% \pilcrow{In the meane time, whyles the Clerkes do syng the Offertory, so many as are disposed, shall offer unto the poore mennes boxe every one accordynge to his habilitie and charitable mynde. And at the offeryng daies appoynted, every manne and woman shall paie to the Curate, the due and accustomed offerynges.}


% \pilcrow{Then so manye as shalbe partakers of the holy Communion, shall tary still in the quire, or in some convenient place nigh the quire, the men on the one side, and the women on the other syde. All other (that mynde not to receive the said holy Communion) shall departe out of the quire, except the ministers and Clerkes.}

% \pilcrow{Than shall the minister take so muche Bread and Wine, as shalt suffice for the persons appoynted to receive the holy Communion, laiyng the breade upon the corporas, or els in the paten, or in some other comely thyng, prepared for that purpose. And puttyng ye wyne into the Chalice, or els in some faire or convenient cup, prepared for that use (if the Chalice will not serve), puttyng thereto a litle pure and cleane water: And settyng both the breade and wyne upon the Alter: then the Priest shall saye.}

\pilcrow{Whilst these Sentences are said or sung, the Deacons, Church-wardens, or other fit person appointed for that purpose, shall receive the alms for the poor, or other devotions of the people,
% in a decent basin to be provided by the Parish for that purpose; 
and reverently bring them to the Priest, who shall humbly present and place them upon the Holy Table in a decent bason to be provided for that purpose.}

\bigskip
% \pilcrow{1912: While the Presbyter distinctly pronounceth one or more of these sentences for the Offertory, the Deacon, or (if no such be present) some other fit person, shall receive the devotions of the people there present, in a bason provided for that purpose. And when all have offered, her shall reverently bring the said bason, with the offerings therein, and deliver it to the Presbyter; who shall humbly present it before the Lord, and set it upon the Holy Table.}

% \pilcrow{And when there is a Communion, the Priest shall then place upon the Table so much Bread and Wine, as he shall think sufficient. After which done, the Priest shall say,}
%1912
\pilcrow{{And when there is a Communion,} the Priest shall then offer up, and place the bread and wine prepared for the Sacrament upon the Lord’s Table; and shall say,}

\drop{Blessed be thou, O {\scshape Lord} God, for ever and ever. Thine, O {\scshape Lord}, is the greatness, and the glory, and the victory, and the majesty: for all that is in the heaven and in the earth, is thine: thine is the kingdom, O {\scshape Lord}, and thou art exalted as head above all: both riches and honour come of thee, and of thine own do we give unto thee. \R Amen.}
\scripture{1 Chronicles xxix.~10}

\bigskip
\centerline{Let us pray for the whole state of Christ's Church.} % militant here in earth.

% \begin{wrapfigure}[lineheight]{position}[overhang]{width}

\drop{Almighty and everliving God, who by thy holy Apostle hast taught us to make prayers, and supplications, and to give thanks for all men;}
% \begin{wrapfigure}[5]{r}{0.40\textwidth}{\footnotesize
% \rubric{* If there be no alms or oblations, then the words \emph{[to accept our alms and oblations]} be left out unsaid.}\par
% }\end{wrapfigure}\noindent
We humbly beseech thee most mercifully [\footnote{\rubric{If there be no alms or oblations, then the words \emph{[to accept our alms and oblations]} be left out unsaid.}}\emph{to accept our alms and \grealtcross\ oblations, and}] to receive these our prayers, which we offer unto thy Divine Majesty; beseeching thee to inspire continually the Universal Church with the spirit of truth, unity, and concord: And grant, that all they that do confess thy holy Name may agree in the truth of thy holy Word, and live in unity, and godly love.



% "M & SA" We beseech thee also to lead all nations in the way of righteousness and peace, so directing all Kings, Presidents, and ruling Authorities, that under them the world may bc godly and quietly governed. And grant unto thy servant George, our King, his Ministers and Parliaments, and all who are set in authority over us, that they may truly and impartially minister justice, to the removing of all wickedness and vice, and the maintenance of thy true religion and virtue.


% We beseech thee also to save and defend all Christian Kings, Princes, and Governours; and specially thy Servant ELIZABETH our Queen; that under her we may be godly and quietly governed: And grant unto her whole Council, and to all that are put in authority under her, that they may truly and impartially minister justice, to the punishment of wickedness and vice, and to the maintenance of thy true religion, and virtue.

%orange book
% We beseech thee also to lead all nations in the way of peace and righteousness, so directing all Kings and ruling authorities that under them the world may be godly and quietly governed.
%SA1954
We beseech thee also to lead all nations into the way of righteousness and peace, and so to direct all 
% Kings, Presidents and Rulers
ruling authorities,
that under them the world may be godly and quietly governed. 
And grant unto all that are put in authority,
% under them, 
% American
% [We beseech thee also, so to direct and dispose the hearts of all Christian Rulers,] 
that they may truly and indifferently minister justice, to the punishment of wickedness and vice, and to the maintenance of thy true religion, and virtue.

Give grace, O heavenly Father, to all Bishops, Priests, and Deacons, 
% 1912 Scot
% and Curates (1662)
% and other Ministers, (1928 Amer)
especially to thy servant \emph{N.} our bishop, %En28
that they may both by their life and doctrine set forth thy true and lively Word, and rightly and duly administer thy holy Sacraments.

And to all thy people give thy heavenly grace,
%; and specially to this congregation here present; 
that with meek heart and due reverence, they may hear and receive thy holy Word; truly serving thee in holiness and righteousness all the days of their life.
And especially we commend unto thy merciful goodness this congregation which is here assembled in thy Name, to celebrate the Commemoration of the most glorious death of thy Son.

And we most humbly beseech thee, of thy goodness, O Lord, to comfort and succour all them, who in this transitory life are in trouble, sorrow, need, sickness, or any other adversity.

%And we also bless thy holy Name for all thy servants departed this life in thy faith and fear; beseeching thee to give us grace so to follow their good examples, that with them we may be partakers of thy heavenly kingFdom.

% {\begin{multicols}{2}
% And here we do give unto thee most high praise and hearty thanks for the wonderful grace and virtue declared in all thy saints from the beginning of the world: And chiefly in the glorious and most blessed Virgin Mary, mother of thy Son Jesus Christ our Lord and God, and in the holy Patriarchs, Prophets, Apostles, and Martyrs, whose examples (O Lord) and stedfastness in the faith, and keeping thy holy commandments, grant us to follow.

% We commend unto thy mercy all other thy servants which are departed from us with the sign of faith, and now to rest in the sleep of peace: Grant unto them, we beseech thee, thy mercy and everlasting peace, and that at the day of the general resurrection, we and all they which be of the mystical body of thy Son, may all together be set on his right hand, and hear that his most joyful voice: Come ye blessed of my Father, and possess the kingdom which is prepared for you from the beginning of the world.



And here we do give unto thee most high praise and hearty thanks for the wonderful grace and virtue declared in all thy Saints, from the beginning of the world: And chiefly in the glorious and most blessed Virgin Mary, Mother of thy Son Jesus Christ our Lord and God, [in \emph{N.},] and in the holy Patriarchs, Prophets, Apostles and Martyrs, whose examples, O Lord, and stedfastness in thy faith, and keeping thy holy commandments, grant us to follow.

We commend unto thy mercy, O Lord, all other thy servants which are departed hence from us, with the sign of faith, and now do rest in the sleep of peace.  Grant unto them, we beseech thee, thy mercy, and everlasting peace, and that, at the day of the general resurrection, we and all thy servants which be of the mystical body of thy Son, may altogether be set on his right hand, and hear that his most joyful voice: Come unto me, O ye that be blessed of my Father, and possess the kingdom, which is prepared for you from the beginning of the world. 


% \end{multicols}}

Grant this, O Father, for Jesus Christ's sake, our only Mediator and Advocate. \R Amen.

\section{The Consecration}

\pilcrow{Turning himself to the people the Priest shall say,}
\V The Lord be with you. \R And with thy spirit.

\V Lift up your hearts.  \R We lift them up unto the Lord.

\V Let us give thanks unto our Lord God. \R It is meet and right so to do.

\centerline{\rubric{Then shall the Priest turn to the Lord's Table, and say,}}

\drop{It is very meet, right, and our bounden duty, that we should at all times, and in all places, give thanks unto thee, O Lord, Holy Father, Almighty, Everlasting God.}
% \rubric{* These words} [Holy Father] \rubric{must be omitted on Trinity Sunday.}

{\footnotesize\centering\rubric{Here shall follow the proper Preface, according to the time, if there be any specially appointed: or else immediately shall follow, \emph{Therefore with Angels \etc}}\par}


% \drop{Therefore with Angels and Archangels, and with all the company of heaven, we laud and magnify thy glorious Name; evermore praising thee, and saying,}
% \drop{Holy, holy, holy, Lord God of hosts, heaven and earth are full of thy glory: Glory be to thee, O Lord most High.}
% Blessed is he that cometh in the Name of the Lord;
% Hosanna in the highest.

\section*{Proper Prefaces}
\prefaceCaption{Upon}{Christmas Day,}{and seven days after.}
\drop{Because thou didst give Jesus Christ thine only Son to be born as at this time for us; who, by the operation of the Holy Ghost, was made very man of the substance of the Virgin Mary his mother; and that without spot of sin, to make us clean from all sin. Therefore with Angels, \etc}

\prefaceCaption{Upon the}{Epiphany,}{and seven days after.}
\drop{Through Jesus Christ our Lord: Who in substance of our mortal flesh manifested forth his glory: That he might bring us out of darkness into his own glorious light.  Therefore with Angels, \etc}

% (sE) Maundy Thursday

\prefaceCaption{Upon}{Easter Day,}{and seven days after.}
\drop{But chiefly are we bound to praise thee for the glorious Resurrection of thy Son Jesus Christ our Lord: for he is the very Paschal Lamb, which was offered for us, and hath taken away the sin of the world; who by his death hath destroyed death, and by his rising to life again hath restored to us everlasting life. Therefore with Angels, \etc}


\prefaceCaption{Upon}{Ascension Day,}{and seven days after.}
\drop{Through thy most dearly beloved Son Jesus Christ our Lord; who after his most glorious Resurrection manifestly appeared to all his Apostles, and in their sight ascended up into heaven to prepare a place for us; that where he is, thither we might also ascend, and reign with him in glory. Therefore with Angels, \etc}


\prefaceCaption{Upon}{Whitsunday,}{and six days after.}
\drop{Through Jesus Christ our Lord; according to whose most true promise, the Holy Ghost came down as at this time from heaven with a sudden great sound, as it had been a mighty wind in the likeness of fiery tongues, lighting upon the Apostles, to teach them, and to lead them to all truth; giving them both the gift of divers languages, and also boldness with fervent zeal constantly to preach the Gospel unto all nations; whereby we have been brought out of darkness and error into the clear light and true knowledge of thee, and of thy Son Jesus Christ. Therefore with Angels, \etc}

\prefaceCaption{Upon the Feast of}{Trinity}{only.}
\drop{Who 
with thine only-begotton Son and the Holy Ghost %AmEn1928
art one God, one Lord,
in Trinity of Persons and Unity of Substance. %AmEn1928
% ; not one only Person, but three Persons in one Substance.
For that which we believe of 
% the glory of the Father, 
thy glory, O Father, %AmEn1928
the same we believe of the Son, and of the Holy Ghost, without any difference or inequality. Therefore with Angels, \etc}


\prefaceCaption{Upon the}{Purification, Annunciation, \rubric{and} Transfiguration}{.}
\drop{Because in the Mystery of the Word made flesh, thou hast caused a new light to shine in our hearts, to give the knowledge of thy glory in the face of thy Son, Jesus Christ our Lord. Therefore with Angels, \etc} %am1928

\prefaceCaption{Upon}{All Saints’ Day}{and the Feasts of Apostles, Evangelists, and \emph{St.\ John Baptist’s Nativity}, except when the Proper Preface of any Principal Feast is appointed.} 
\drop{Who, in the multitude of thy saints, hast compassed us about with so great a cloud of witnesses that we, rejoicing in their fellowship, may run with patience the race that is set before us, and together with them may receive the crown of glory that fadeth not away. Therefore with Angels, \etc }


\subsubsection{After each of which Prefaces shall immediately be sung or said,}

\drop{Therefore with Angels and Archangels, and with all the company of heaven, we laud and magnify thy glorious Name; evermore praising thee, and saying,}
\smallskip
\drop{Holy, holy, holy, Lord God of hosts, heaven and earth are full of thy glory: Glory be to thee, O Lord most High.
\grecross\ Blessed is he that cometh in the Name of the Lord;
Hosanna in the highest.}

\bigskip

\pilcrow{When the Priest, standing before the Table, hath so ordered the Bread and Wine, that he may with the more readiness and decency break the Bread before the people, and take the Cup into his hands, he shall say the Prayer of Consecration, as followeth.}

\drop{All glory be to thee, Almighty God, our heavenly Father, for that thou of thy tender mercy didst give thine only Son Jesus Christ to suffer death upon the Cross for our redemption; who made there (by his one oblation of himself once offered) a full, perfect, and sufficient sacrifice, oblation, and satisfaction, for the sins of the whole world; and did institute, and in his holy Gospel command us to continue, a perpetual memory of that his precious death 
and sacrifice, %scottish
until his coming again;}

% 1549: Hear us, O most merciful Father, and with thy Holy Spirit and word vousafe to bless \grecross and sanctify \grecross these thy gifts and creatures of bread and wine that they may be unto us the body and blood of thy most dearly beloved Son Jesus Christ.

%1637 Heare us, O mercifull Father, we most humbly beseech thee, and of thy almighty goodnesse vouchsafe so to blesse and sanctifie with thy word and holy Spirit these thy gifts and creatures of bread and wine, that they may bee unto us the body and bloud of thy most dearly beloved Son; so that wee receiving them according to thy Sonne our Saviour Jesus Christs holy institution, in remembrance of his death and passion, may be partakers of the same his most precious body and bloud
%English: Hear us, O merciful Father, we most humbly beseech thee; and grant that we receiving these thy creatures of bread and wine, according to thy Son our Saviour Jesus Christ's holy institution, in remembrance of his death and passion, may be partakers of his most blessed Body and Blood: 


% [For, in the night (S)  (a) he took]
\drop{Who, in the same night that he was betrayed, \footnote{\rubric{Here the Priest is to take the Bread into his hands:}}took Bread; and, when he had given thanks, \footnote{\rubric{And here to touch or break the Bread:}}he brake it, and gave it to his disciples, saying, Take, eat, \footnote{\rubric{And here to lay his hand upon all the Bread.}}}

{\centering{\scshape this is my Body which is given for you:} 
Do this in remembrance of me.\par}

\medskip

Likewise after supper he \footnote{\rubric{Here he is to take the Cup into his hand:}}took the Cup; and, when he had given thanks, he gave it to them, saying, Drink ye all of this; \footnote{\rubric{And here to lay his hand upon every vessel (be it Chalice or Flagon) in which there is any Wine to be consecrated.}}

{\centering{\scshape for this is my Blood of the New Testament, which is shed for you and for many for the remission of sins:}

Do this, as oft as ye shall drink it, in remembrance of me.\par}


\centerline{\rubric{The Oblation}}
\drop{Wherefore, O Lord, and heavenly Father, according to the institution of thy dearly beloved Son our Saviour Jesus Christ, we thy humble servants do celebrate and make here before thy divine Majesty, with these thy holy gifts, {\scshape which we now offer unto thee,} the memorial thy Son hath commanded us to make; having in remembrance his blessed passion, and precious death, his mighty resurrection and glorious ascension; rendering unto thee most hearty thanks for the innumerable benefits which he hath procured unto us by the same, 
and looking for his coming again with power and great glory.} %late scottish

% ["invocation" - 1928, 1928 scottish (all) ]
% 1549: Hear us, O most merciful Father, and with thy Holy Spirit and word vousafe to bless \grecross and sanctify \grecross these thy gifts and creatures of bread and wine that they may be unto us the body and blood of thy most dearly beloved Son Jesus Christ.

%1637 Heare us, O mercifull Father, we most humbly beseech thee, and of thy almighty goodnesse vouchsafe so to blesse and sanctifie with thy word and holy Spirit these thy gifts and creatures of bread and wine, that they may bee unto us the body and bloud of thy most dearly beloved Son; so that wee receiving them according to thy Sonne our Saviour Jesus Christs holy institution, in remembrance of his death and passion, may be partakers of the same his most precious body and bloud
%English 1928: Hear us, O Merciful Father, we most humbly beseech thee, and with thy Holy and Life-giving Spirit vouchsafe to bless and sanctify both us and these thy gifts of Bread and Wine, that they may be unto us the Body and Blood of thy Son, our Saviour, Jesus Christ, to the end that, receiving the same, we may be strengthened and refreshed both in body and soul.
%English: Hear us, O merciful Father, we most humbly beseech thee; and grant that we receiving these thy creatures of bread and wine, according to thy Son our Saviour Jesus Christ's holy institution, in remembrance of his death and passion, may be partakers of his most blessed Body and Blood: 

% "Scottish office" - newer than 1637; older than 1912.  The 1637 is actually similar but goes on with the english ending.
\centerline{\rubric{The Invocation}}
\drop{Hear us, O Merciful Father, we most humbly beseech thee, % ENglish
% And we most humbly beseech thee, O merciful Father, to hear us, Scottish
and of thy almighty goodness vouchsafe to \grealtcross\  bless and \grealtcross\  sanctify, with thy Word and Holy Spirit, these thy gifts and creatures of bread and wine, that they may become the Body and Blood of thy most dearly beloved Son, }
to the end that all who shall receive the same may be sanctified both in body and soul, and preserved unto everlasting life. % (late Scottish)]


% (scottish)

\drop{And we entirely %english
% [earnestly]  %scottish
desire thy fatherly goodness mercifully to accept this our sacrifice of praise and thanksgiving; most humbly beseeching thee to grant, that by the merits and death of thy Son Jesus Christ, and through faith in his blood, we and all thy whole Church may obtain remission of our sins, and all other benefits of his passion.}

And here we %[humbly] scottish
offer and present unto thee, O Lord, our selves, our souls and bodies, to be a reasonable, holy, and lively sacrifice unto thee; 
% humbly beseeching thee, that whosoever shall be partakers 
% humbly beseeching thee, that all we, who are partakers 
humbly beseeching thee, that all we, who shall be partakers
% that we, and all others who shall be partakers
of this holy Communion, may
worthily receive the most precious Body and Blood of thy Son Jesus Christ, and
be fulfilled with thy grace and heavenly \grecross\  benediction,
and made one body with him, that he may dwell in us, and we in him.

And although we be unworthy, through our manifold sins, to offer unto thee any sacrifice, yet we beseech thee to accept this our bounden duty and service, and command these our prayers and supplications by the ministry of thy holy Angels to be brought up into thy holy Tabernacle before the sight of thy divine Majesty; not weighing our merits, but pardoning our offences,

Through Jesus Christ our Lord; by whom, and with whom, in the unity of the Holy Ghost, all honour and glory be unto thee, O Father Almighty, world without end. \R Amen.

% \centerline{Let us pray.}
% [And now,]
{\centering\rubric{Here shall the people join with the Priest in the Lord’s Prayer, the Priest first saying,}\par}
As our Saviour Christ hath commanded and taught us, we are bold to say,

\smallskip
% \pilcrow{Then shall the Priest say the Lord's Prayer, the people repeating after him every Petition.}
\drop{Our Father, which art in heaven, Hallowed be thy Name. Thy kingdom come. Thy will be done, in earth as it is in heaven. Give us this day our daily bread. And forgive us our trespasses, As we forgive them that trespass against us. And lead us not into temptation; But deliver us from evil.}

For thine is the kingdom, the power, and the glory, For ever and ever. Amen.

\centerline{\rubric{Here the Priest is to break the consecrated bread.}}

\medskip

\pilcrow{Then shall the Priest say or sing,}
\drop{The Peace of the Lord be always with you.  \R And with thy spirit.}

\medskip

\drop{Christ, our Paschal Lamb, is offered up for us, once for all, when he bare our sins on his body upon the Cross; for he is the very Lamb of God that taketh away the sins of the world, wherefore let us keep a joyful and holy feast unto the Lord.}


\section{The Communion}

\pilcrow{Then shall the Minister say to them that come to receive the Holy Communion,}

\drop{Ye that do truly and earnestly repent you of your sins, and are in love and charity with your neighbours, and intend to lead a new life, following the commandments of God, and walking from henceforth in his holy ways; Draw near with faith, and take this holy Sacrament to your comfort; and make your humble confession to Almighty God, meekly kneeling upon your knees.}

\medskip
%1662
% \pilcrow{Then shall this general Confession be made, in the name of all those that are minded to receive the holy Communion, by one of the Ministers; both he and all the people kneeling humbly upon their knees, and saying,}
%Scottish 1912
% Then shall this general confession be made by the people, along with the Presbyter; he first kneeling down.
%american 1928
\pilcrow{Then shall this \emph{General Confession be made}, by the Priest and all those who are minded to receive the Holy Communion, humbly kneeling.}

\drop{Almighty God, Father of our Lord Jesus Christ, Maker of all things, judge of all men; We acknowledge and bewail our manifold sins and wickedness, Which we, from time to time, most grievously have committed, By thought, word, and deed, Against thy Divine Majesty, Provoking most justly thy wrath and indignation against us. We do earnestly repent, And are heartily sorry for these our misdoings; The remembrance of them is grievous unto us; The burden of them is intolerable. Have mercy upon us, Have mercy upon us, most merciful Father; For thy Son our Lord Jesus Christ's sake, Forgive us all that is past; And grant that we may ever hereafter Serve and please thee In newness of life, To the honour and glory of thy Name; Through Jesus Christ our Lord. Amen.}


\medskip


\pilcrow{Then shall the Priest (or the Bishop, being present,) standing up, and turning himself to the people, pronounce this Absolution.}
\drop{Almighty God, our heavenly Father, who of his great mercy hath promised forgiveness of sins to all them that with hearty repentance and true faith turn unto him; Have mercy upon you; pardon  \grealtcross\  and deliver you from all your sins; confirm and strengthen you in all goodness; and bring you to everlasting life; through Jesus Christ our Lord. \R Amen.}

\medskip

\centerline{\pilcrow{Then shall the Priest say,}}
Hear what comfortable words our Saviour Christ saith unto all that truly turn to him.
\drop{Come unto me all that travail and are heavy laden, and I will refresh you.}\scripture{St.~Matthew xj.~28}

So God loved the world, that he gave his only-begotten Son, to the end that all that believe in him should not perish, but have everlasting life.\scripture{St.~John iij.~16}

\centerline{Hear also what Saint Paul saith.}

This is a true saying, and worthy of all men to be received, That Christ Jesus came into the world to save sinners.\scripture{1 Timothy i.~15.}

\centerline{Hear also what Saint John saith.}

If any man sin, we have an Advocate with the Father, Jesus Christ the righteous; and he is the propitiation for our sins.\scripture{1 St.~John ij.~1.}

\medskip


{\centering\footnotesize\rubric{Then shall the Priest, turning him to the Altar, kneel down, and say, in the name of all them that shall communicate, this Collect of humble access to the Holy Communion, as followeth:}\par} % Scottish rubric

\drop{We do not presume to come to this thy Table, O merciful Lord, trusting in our own righteousness, but in thy manifold and great mercies. We are not worthy so much as to gather up the crumbs under thy Table. But thou art the same Lord, whose property is always to have mercy: Grant us therefore, gracious Lord, so to eat the flesh of thy dear Son Jesus Christ, and to drink his blood, that our sinful bodies may be made clean by his body, and our souls washed through his most precious blood, and that we may evermore dwell in him, and he in us. \R Amen.}


\medskip
% Scottish Position:
\pilcrow{Here may be sung or said:} %scottish rubric
% \pilcrow{In the Communion time the choir and people may say or sing the following, beginning so soon as the Priest doth receive the holy Communion.}

\drop{O Lamb of God, that takest away the sins of the world: have mercy upon us.

O Lamb of God that, takest away the sins of the world: have mercy upon us.

O Lamb of God, that takest away the sins of the world: grant us thy peace.}
\bigskip

\pilcrow{Then shall the Minister first receive the Communion in both kinds himself, and then proceed to deliver the same to the Bishops, Priests, and Deacons, in like manner, (if any be present,) and after that to the people also in order, 
into their hands, % Percy: This has always been the custom with us.
all meekly kneeling. And, when he delivereth the Bread to any one, he shall say,}
\drop{The Body of our Lord Jesus Christ, which was given for thee, preserve thy body and soul unto everlasting life. \R Amen.}

\smallskip

{\centering\footnotesize\rubric{Take and eat this in remembrance that Christ died for thee, and feed on him in thy heart by faith with thanksgiving.}\par}

\bigskip

{\centering\footnotesize\rubric{And the Minister that delivereth the Cup to any one shall say,}\par}

\drop{The Blood of our Lord Jesus Christ, which was shed for thee, preserve thy body and soul unto everlasting life. \R Amen.}

\smallskip

{\centering\footnotesize\rubric{Drink this in remembrance that Christ's Blood was shed for thee, and be thankful.}\par}

\bigskip


% \pilcrow{If the consecrated Bread or Wine be all spent before all have communicated, the Priest is to consecrate more cording to the Form before prescribed: Beginning at [Our Saviour Christ in the same night, etc.] for the blessing of the Bread ; and at [Likewise after Supper, etc.] for the blessing of the Cup.}

\pilcrow{When all have communicated, the Minister shall return to the Lord's Table, and reverently place upon it what remaineth of the consecrated Elements, covering the same with a fair linen cloth.}

\medskip
% Green book.
\pilcrow{Here may follow an Anthem or Hymn.}

\section{The Thanksgiving}

\pilcrow{Then shall the Priest give thanks to God in the name of all them that have communicated, turning him first to the people, and saying,}

%1923 (1929 option, along with 1549)
\V O give thanks unto the Lord, for he is gracious:  \R And his mercy endureth for ever.

%1549
\V The Lord be with you. \R And with thy spirit.

\centerline{Let us pray.}

%proposed 1928, shortened from Scottish.
% Having now 
% by faith <- added! 
% received the precious Body and Blood of Christ, let us give thanks unto our Lord God,

\drop{Almighty and everliving God, we most heartily thank thee, for that thou dost vouchsafe to feed us, who have duly received these holy mysteries, with the spiritual food of the most precious Body and Blood of thy Son our Saviour Jesus Christ; and dost assure us thereby of thy favour and goodness towards us; and that we are very members incorporate in the mystical body of thy Son, which is the blessed company of all faithful people; and are also heirs through hope of thy everlasting kingdom, by the merits of the most precious death and passion of thy dear Son. And we most humbly beseech thee, O heavenly Father, so to assist us with thy grace, that we may continue in that holy fellowship, and do all such good works as thou hast prepared for us to walk in; through Jesus Christ our Lord, to whom, with thee and the Holy Ghost, be all honour and glory, world without end. Amen.}

\bigskip
% Rubric from "The Orange Book" - 1923
\pilcrow{The \emph{Gloria in excelsis} may be omitted after the \emph{Kyrie eleison}, and sung here instead: provided that it be always said or sung in one or other position on Holy-days and on all Sundays except those in Advent and from Septuagesima to Palm Sunday inclusive.}
\bigskip

\centerline{\rubric{For the Post-Communions see pages X-X}} %scottish
\bigskip

\V The Lord be with you. \R And with thy spirit.

\V Depart in peace, \rubric{or} Let us bless the Lord.

\R Thanks be to God.

\medskip
\pilcrow{Then the Priest (or Bishop if he be present) shall let them depart with this Blessing.}
\drop{The peace of God, which passeth all understanding, keep your hearts and minds in the knowledge and love of God, and of his son Jesus Christ our Lord: and the blessing of God Almighty, the Father, \grealtcross\ the Son, and the Holy Ghost, be amongst you and remain with you always. \R Amen.}

\medskip

\fleuron

\bigskip

\section{Collects}
\pilcrow{To be said after the Offertory, when there is no Communion, every such day one or more; and the same may be said also, as often as occasion shall serve, after the Collects either of Morning or Evening Prayer, or the Litany, or immediately before the Blessing at Holy Communion, by the discretion of the Minister.}

% Scottish:
\drop{O Almighty Father, wellspring of life to all things that have being, from amid the unwearied praises of Cherubim and Seraphim who stand about thy throne of light which no man can approach unto, give ear, we humbly beseech thee, to the supplications of thy people who put their sure trust in thy mercy, through Jesus Christ our Lord.  \R Amen.}
% [From the Book of Deer.]}

\smallskip


\drop{O Lord Jesus Christ, before whose judgement-seat we must all appear and give account of the things done in the body: Grant, we beseech thee, that when the books are opened in that day, the faces of thy servants may not be ashamed, through thy merits, O blessed Saviour, who livest and reignest with the Father and the Holy Spirit, one God, world without end.  \R Amen.}
% [From the Altus of St Columba]}

\smallskip

\drop{Assist us mercifully, O Lord, in these our supplications and prayers, and dispose the way of thy servants towards the attainment of everlasting salvation; that, among all the changes and chances of this mortal life, they may ever be defended by thy most gracious and ready help; through Jesus Christ our Lord. \R Amen.}

\smallskip

\drop{O Almighty Lord, and everlasting God, vouchsafe, we beseech thee, to direct, sanctify, and govern, both our hearts and bodies, in the ways of thy laws, and in the works of thy commandments; that through thy most mighty protection, both here and ever, we may be pre- served in body and soul; through our Lord and Saviour Jesus Christ. \R Amen.}

\smallskip


\drop{Grant, we beseech thee, Almighty God, that the words, which we have heard this day with our outward ears, may through thy grace be so grafted inwardly in our hearts, that they may bring forth in us the fruit of good living, to the honour and praise of thy Name; through Jesus Christ our Lord. \R Amen.}

\smallskip

\drop{Prevent us O Lord, in all our doings with thy most gracious favour, and further us with thy continual help; that in all our works begun, continued, and ended in thee, we may glorify thy holy Name, and finally by thy mercy obtain everlasting life; through Jesus Christ our Lord. \R Amen.}

\smallskip

\drop{Almighty God, the fountain of all wisdom, who knowest our necessities before we ask, and our ignorance in asking; We beseech thee to have compassion upon our infirmities; and those things, which for our unworthiness we dare not, and for our blindness we cannot ask, vouchsafe to give us, for the worthiness of thy Son Jesus Christ our Lord. \R Amen.}

\smallskip

% The two following collects may be said before the Blessing.

\drop{O Lord, our God, thou Saviour of the world, through whom we have celebrated these sacred mysteries: Receive our humble thanksgiving, and of thy great mercy vouchsafe to sanctify us evermore in body and soul, who livest and reignest with the Father and the Holy Spirit, one God, world without end.  \R Amen.}

\smallskip

\drop{Almighty God, who hast promised to hear the petitions of them that ask in thy Son's Name; We beseech thee mercifully to incline thine ears to us that have made now our prayers and supplications unto thee; and grant, that those things, which we have faithfully asked according to thy will, may effectually be obtained, to the relief of our necessity, and to the setting forth of thy glory; through Jesus Christ our Lord. \R Amen.}

    %Scottish 1912
\subsection{Prayers for Certain Festivals and Seasons}

\prefaceCaption{}{Advent.}{} % Proofed vs 1929 on 11-29-21
\drop{Grant, O Almighty God, that as thy blessed Son Jesus Christ at his first advent came to seek and to save that which was lost, so at his second and glorious appearing he may find in us the fruits of the redemption which he wrought; who liveth and reigneth, with thee and the Holy Spirit, one God, world without end. Amen.}

\prefaceCaption{}{Christmas Day,}{and seven days after.}
\drop{O God, who hast given us grace at this time to celebrate the birth of our Saviour, Jesus Christ: We laud and magnify thy glorious Name for the countless blessings which he hath brought unto us; and we beseech thee to grant that We may ever set forth thy praise in joyful obedience to thy will; through the same Jesus Christ our Lord. \R Amen.}

%1928 prop
\prefaceCaption{}{New Year’s Day}{}
\drop{O eternal Lord God, who hast brought thy servants to the beginning of another year: Pardon, we humbly beseech thee, our transgressions in the past, and graciously abide with us all the days of our life; through Jesus Christ our Lord.  \R Amen}

\prefaceCaption{}{Epiphany,}{and seven days after.}
\drop{Almighty God, who at the baptism of thy blessed Son Jesus Christ in the river Jordan didst manifest his glorious Godhead: Grant, we beseech thee, that the brightness of his presence may shine in our hearts, and his glory be set forth in our lives; through the same Jesus Christ our Lord. \R Amen.}

\prefaceCaption{}{Easter Day,}{and seven days after.}
\drop{O Lord God Almighty, whose blessed Son, our Saviour, Jesus Christ, did on the third day rise triumphant over death: Raise us, we beseech thee, from the death of sin unto the life of righteousness, that we may seek those things which are above, where he sitteth on thy right hand in glory; and this we beg for the sake of the same, thy Son, Jesus Christ our Lord. \R Amen.}

\prefaceCaption{}{Ascension Day,}{and seven days after.}
\drop{Almighty God, whose blessed Son, our Saviour, Jesus Christ, ascended far above all heavens that he might fill all things: Mercifully give us faith to perceive that according to his promise he abideth with his Church on earth, even unto the end of the world; through the same Jesus Christ our Lord. \R Amen.}

\prefaceCaption{}{Whitsunday,}{and six days after.}
\drop{O Almighty God, who on the day of Pentecost didst send the Holy Ghost the Comforter to abide in thy Church unto the end: Bestow upon us and upon all thy faithful people his manifold gifts of grace, that with minds enlightened by his truth and hearts purified by his presence, we may day by day be strengthened with power in the inward man; through Jesus Christ our Lord, who with thee and the same Spirit liveth and reigneth, one God, world without end. \R Amen.}


\prefaceCaption{}{Trinity Sunday.}{}
\drop{O Lord God Almighty, Eternal, Immortal, Invisible, the mysteries of whose being are unsearchable: Accept, we beseech thee, our praises for the revelation which thou hast made of thyself, Father, Son, and Holy Ghost, three Persons, and one God; and mercifully grant, that ever holding fast this faith, we may magnify thy glorious Name; who livest and reignest, one God, world without end. \R Amen.}




\section{Exhortations}

\pilcrow{When the Minister giveth warning for the celebration of the holy Communion, (which he shall always do upon the Sunday, or some Holy-day, immediately preceding,) after the Sermon or Homily ended, he shall read this Exhortation following.}
\drop{Dearly beloved, on — I purpose, through God’s assistance, to administer to all such as shall be religiously and devoutly disposed the most comfortable Sacrament of the Body and Blood of Christ; to be by them received in remembrance of his meritorious Cross and Passion; whereby alone we obtain remission of our sins, and are make partakers of the Kingdom of heaven.} 

Wherefore it is our duty to render most humble and hearty thanks to Almighty God our heavenly Father, for that he hath given his Son our Saviour Jesus Christ, not only to die for us, but also to be our spiritual food and sustenance in that holy Sacrament.

Which being so divine and comfortable a thing to them who receive it worthily, and so dangerous to them that will presume to receive it unworthily; my duty is to exhort you in the mean season to consider the dignity of that holy mystery, and the great peril of the unworthy receiving thereof; and so to search and examine your own consciences, (and that nor lightly, and after the manner of dissemblers with God; but so) that ye may come holy and clean to such a heavenly Feast, in the marriage-garment required by God in holy Scripture, and be received as worthy partakers of that holy Table.

The way and means thereto is; First, to examine your lives and conversations by the rule of God’s commandments; and wherein soever ye shall perceive yourselves to have offended, either by will, word, or deed, there to bewail your own sinfulness, and to confess yourselves to Almighty God, with full purpose of amendment of life. 

And if ye shall perceive your offences to be such as are not only against God, but also against your neighbours; then ye shall reconcile yourselves unto them; being ready to make restitution and satisfaction, according to the uttermost of your powers, for all injuries and wrongs done by you to any other; and being likewise ready to forgive others that have offended you, as ye would have forgiveness of your offences at God’s hand: \emph{[for otherwise the receiving of the holy Communion doth nothing else but increase your \emph{damnation} [guilt, condemnation].]}

Therefore if any of you be a blasphemer of God, an hinderer or slanderer of his Word, an adulterer, or be in malice, or envy, or in any other grievous crime, repent you of your sins, or else come not to that holy Table\rubric{; lest, after the taking of that holy Sacrament, the devil enter into you, as he entered into Judas, and fill you full of all iniquities, and bring you to destruction both of body and soul}.

And because it is requisite, that no man should come to the holy Communion, but with a full trust in God's mercy, and with a quiet conscience; therefore if there be any of you, who by this means cannot quiet his own conscience herein, but requireth further comfort or counsel, let him come to me, or to some other discreet and learned Minister of God's Word, and open his grief; that by the ministry of God's holy Word he may receive the benefit of absolution, together with ghostly counsel and advice, to the quieting of his conscience, and avoiding of all scruple and doubtfulness.

% Rubric moved to Confession.

\medskip
%a new form, composed apparently by Peter Martyr at the instance of Bucer. (i)

{\centering\rubric{Or, in case he shall see the people negligent to come to the Holy Communion, instead of the former, he may use this Exhortation.}\par}


\drop{Dearly beloved brethren, on — I intend, by God’s grace, to celebrate the Lord’s Supper: unto which, in God’s behalf, I bid you all that are here present; and beseech you, for the Lord Jesus Christ’s sake, that ye will not refuse to come thereto, being so lovingly called and bidden by God himself.}

Ye know how grievous and unkind a thing it is, when a man hath prepared a rich feast, decked his table with all kind of provision, so that there lacketh nothing but the guests to sit down; and yet they who are called (without any cause) most unthankfully refuse to come. Which of you in such a case would not be moved? Who would not think a great injury and wrong done unto him? Wherefore, most dearly beloved in Christ, take ye good heed, lest ye, withdrawing yourselves from this holy Supper, provoke God’s indignation against you. It is an easy matter for a man to say, I will not communicate, because I am otherwise hindered with worldly business. But such excuses are not so easily accepted and allowed before God. If any man say, I am a grievous sinner, and therefore am afraid to come: wherefore then do ye not repent and amend? When God calleth you, are ye not ashamed to say ye will not come? When ye should return to God, will ye excuse yourselves, and say ye are not ready? Consider earnestly with yourselves how little such feigned excuses will avail before God. They that refused the feast in the Gospel, because they had bought a farm, or would try their yokes of oxen, or because they were married, were not so excused, but counted unworthy of the heavenly feast.

I, for my part, shall be ready; and, according to mine office, I bid you in the Name of God, I call you in Christ’s behalf, I exhort you, as ye love your own salvation, that ye will be partakers of this Holy Communion. And as the Son of God did vouchsafe to yield up his soul by death upon the Cross for your salvation; so it is your duty to receive the Communion in remembrance of the sacrifice of his death, as he himself hath commanded: which if ye shall neglect to do, consider with yourselves how great injury ye do unto God, and how sore punishment hangeth over your heads for the same; when ye wilfully abstain from the Lord’s Table, and separate from your brethren, who come to feed on the banquet of that most heavenly food.

These things if ye earnestly consider, ye will by God’s grace return to a better mind: for the obtaining whereof we shall not cease to make our humble petitions unto Almighty God our heavenly Father.


\section{General Rubricks}

\pilcrow{Upon the Sundays and other Holy-days (if there be no Communion) shall be said all that is appointed at the Communion, until the end of the general Prayer \emph{For the whole state of Christ’s Church} together with one or more of these Collects last before rehearsed, concluding with the Blessing.}


\pilcrow{And whenever this Service is used, Collects, contained in this Book, or sanctioned by the Bishop, may be said after \emph{The Intercession}, or before the Blessing.}

\medskip

% 1928 Rubrics:
\pilcrow{If any of the consecrated Bread and Wine remain, apart from that which may be reserved,
% for the Communion of the sick,
it shall not be carried out of the church; but the Priest, and such other of the communicants as he shall call unto him, shall, immediately after the Blessing, reverently eat and drink the same.}

% \pilcrow{And there shall be no celebration of the Lord's Supper, except there be a convenient number to communicate with the Priest, according to his discretion.}

% \pilcrow{And if there be not above twenty persons in the Parish of discretion to receive the Communion: yet there shall be no Communion, except four (or three at the least) communicate with the Priest.}

% \pilcrow{And in Cathedral and Collegiate Churches, and Colleges, where there are many Priests and Deacons, they shall all receive the Communion with the Priest every Sunday at the least, except they have a reasonable cause to the contrary.}

% \pilcrow{And to take away all occasion of dissension, and superstition, which any person hath or might have concerning the Bread and Wine, it shall suffice that the Bread be such as is usual to be eaten; but the best and purest Wheat Bread that conveniently may be gotten.}

% \pilcrow{And if any of the Bread and Wine remain unconsecrated, the Curate shall have it to his own use: but if any remain of that which was consecrated, it shall not be carried out of the Church, but the Priest, and such other of the Communicants as he shall then call unto him, shall, immediately after the Blessing, reverently eat and drink the same.}

\pilcrow{The Bread and Wine for the Communion shall be provided by the Curate and the Church-wardens at the charges of the Parish.}


\pilcrow{After the Divine Service ended, the money given at the Offertory shall be disposed of to such pious and charitable uses, as the Minister and Church-wardens shall think fit. Wherein if they disagree, it shall be disposed of as the Ordinary shall appoint.}

\medskip

% all following from 1928Eng
\pilcrow{And note, that every confirmed member of the Church shall communicate at the least three times in the year, of which Easter to be one.} 


\pilcrow{If any be an open and notorious evil liver, or have done any wrong to his neighbours by word or deed, so that the Congregation be thereby offended; the Curate, having knowledge thereof, shall call him and advertise him, that in any wise he presume not to come to the Lord’s Table, until he hath openly declared himself to have truly repented and amended his former naughty life, that the Congregation may thereby be satisfied, which before were offended; and that he hath recompensed the parties, to whom he hath done wrong; or at least declare himself to be in full purpose so to do, as soon as he conveniently may.}


\pilcrow{The same order shall the Curate use with those betwixt whom he perceiveth malice and hatred to reign; not suffering them to be partakers of the Lord’s Table, until he known them to be reconciled. And if one of the parties so at variance be content to forgive from the bottom of his heart all that the other hath trespassed against him, and to make amends for that he himself hath offended, and the other party will not be persuaded to a godly unity, but remain still in his frowardness and malice: the Minister in that case ought to admit the penitent person to the holy Communion, and not him that is obstinate.}


\pilcrow{Provided that every Minister so advertising repelling any, as is specified in the two precedent paragraphs, shall be obliged forthwith to give an account of the same to the Bishop, and therein to obey his order and direction.}

\fleuron

%All the following from the Indian Supplement
\chapter[Order of Communion]{\stylechapter{The}{Order of Communion}{with the Reserved Sacrament.}}
\label{reservedSacrament}
\pilcrow{In the absence of a Priest, a Deacon may administer from the Reserved Sacrament, but he shall use the Collect for the Twenty-first Sunday after Trinity instead of the Absolution, and \emph{The Grace of our Lord Jesus Christ  \etc} instead of the Blessing.}

\pilcrow{When this Order is used for those unable to be present at a celebration of the Lord’s Supper but who are not sick, it shall be used with the Collect, Epistle and Gospel of the Day. At the discretion of the Minister other parts of the Order for the Lord’s Supper (except the setting apart of the Bread and Wine and the Prayer of Consecration) may be added.}

\pilcrow{When this Order is used for the sick, it is fitting that a table be prepared at a convenient place in the sick person’s house. The table shall be covered with a clean white cloth, whereon is to be placed a vessel containing a little water. The Priest shall come at the appointed time, bearing the Reserved Sacrament, and shall place it upon the table.}

% If the sick person is not too weak, the Priest may begin with the Collect, Epistle and Gospel of the Day, or else with the following:

% The Collett

% ALMIGHTY Father, giver of life and health: Look mercifully, we beseech thee, on this thy servant, that by thy blessing upon him and upon those who minister to him, he may speedily be restored to health, if it be thy gracious will, and live to the glory of thy holy name; through Jesus Christ our Lord. Amen.

% Or this,

% ASSIST us mercifully, O Lord, in these our supplications and prayers, and dispose the way of thy servant towards the attainment of everlasting salvation; that among all the changes and chances of this mortal life, he may ever be defended by thy most gracious and ready help; through Jesus Christ our Lord. Amen.

% The Epistle. 2 Corinthians 1. 3

% BLESSED be God, even the Father of our Lord Jesus Christ, the Father of mercies, and the God of all comfort; who comforteth us in all our tribulation, that we may be able to comfort them which are in any trouble, by the comfort wherewith we ourselves are comforted of God. For as the sufferings of Christ abound in us, so our consolation also aboundeth by Christ.

% The Gospel. St John 10. 15, 27-30

% IAM the good shepherd; and I know mine own, and mine own know me, even as the Father knoweth me, and I know the Father; and I lay down my life for the sheep. My sheep hear my voice, and I know them, and they follow me: and I give unto them eternal life; and they shall never perish, and no one shall pluck them out of my hand. My Father, which hath given them unto me, is greater than all; and no one is able to pluck them out of the Father's hand. I and the Father are one.
 

 
\pilcrow{Then the Priest and those present shall say the \emph{General Confession} from the Order for the Lord’s Supper, the Priest adding the \emph{Absolution} and the \emph{Comfortable Words}, or else the following shortened Confession and Absolution may be used:}
\drop{Draw near with faith, and take this Holy Sacrament to your comfort; and make your humble confession to Almighty God.}

\medskip
\drop{We confess to God Almighty, the Father, the Son, and the Holy Ghost, that we have sinned in thought, word, and deed, through our own grievous fault. Wherefore we pray God to have mercy upon us.}

\medskip
{\centering\footnotesize\rubric{After which the Priest shall pronounce this Absolution,}\par} 
\drop{Almighty God have mercy upon you, forgive \grealtcross\ you all your sins, and deliver you from all evil, confirm and strengthen you in all goodness, and bring you to life everlasting; through Jesus Christ our Lord. \R Amen.}

\medskip
{\centering\footnotesize\rubric{Then shall the Priest and those who are to receive the Sacrament say together:}\par} 
\drop{We do not presume to come to this thy Table, O merciful Lord, trusting in our own righteousness, but in thy manifold and great mercies. We are not worthy so much as to gather up the crumbs under thy Table. But thou art the same Lord, whose property is always to have mercy: Grant us therefore, gracious Lord, so to eat the flesh of thy dear Son Jesus Christ, and to drink his blood, that our sinful bodies may be made clean by his body, and our souls washed through his most precious blood, and that we may evermore dwell in him, and he in us. \R Amen.}


\medskip
% Scottish Position:
\pilcrow{Then may be sung or said:} %scottish rubric

\drop{O Lamb of God, that takest away the sins of the world: have mercy upon us.

O Lamb of God that, takest away the sins of the world: have mercy upon us.

O Lamb of God, that takest away the sins of the world: grant us thy peace.}
\bigskip

\medskip
\rubric{After a short space of silence the Priest shall deliver the Sacrament with the customary words of administration.}


\medskip

{\centering\footnotesize\rubric{After a short silence the Priest and all present shall repeat together the Lord's Prayer,}\par} 
\ourFather


\medskip
{\centering\footnotesize\rubric{Then the Priest shall say,}\par} 
\drop{Almighty and everliving God, we most heartily thank thee, for that thou dost vouchsafe to feed us, who have duly received these holy mysteries, with the spiritual food of the most precious Body and Blood of thy Son our Saviour Jesus Christ; and dost assure us thereby of thy favour and goodness towards us; and that we are very members incorporate in the mystical body of thy Son, which is the blessed company of all faithful people; and are also heirs through hope of thy everlasting kingdom, by the merits of the most precious death and passion of thy dear Son. And we most humbly beseech thee, O heavenly Father, so to assist us with thy grace, that we may continue in that holy fellowship, and do all such good works as thou hast prepared for us to walk in; through Jesus Christ our Lord, to whom, with thee and the Holy Ghost, be all honour and glory, world without end. Amen.}

\medskip
{\centering\footnotesize\rubric{And the Blessing,}\par}
\drop{The peace of God, which passeth all understanding, keep your hearts and minds in the knowledge and love of God, and of his son Jesus Christ our Lord: and the blessing of God Almighty, the Father, \grealtcross\ the Son, and the Holy Ghost, be amongst you and remain with you always. \R Amen.}


\fleuron


% \chapter{ An Alternative Order of the Communion}
% \ourFather

% \subsection{\stylesubsec{}{The Collect.}{}}
% \drop{Almighty God, unto whom all hearts be open, all desires known, and from whom no secrets are hid; Cleanse the thoughts of our hearts by the inspiration of thy Holy Spirit, that we may perfectly love thee, and worthily magnify thy holy Name; through Christ our Lord. \R Amen.}

% \medskip
% \centerline{\pilcrow{Here may be sung a Hymn or an Anthem.}}

% \medskip

% \centerline{Lord, have mercy upon us. \rubric{iij.}}
% \centerline{Christ, have mercy upon us. \rubric{iij.}}
% \centerline{Lord, have mercy upon us. \rubric{iij.}}


% \bigskip
% \pilcrow {Then the priest shall turn him to the people and say,}
% \V The Lord be with you. \R And with thy spirit.

% \centerline{Let us pray.}

% \medskip
% \pilcrow{And turning to the Holy Table he shall say the Collect of the Day. Other collects contained in this Book or authorized by the Bishop may follow.}

% \medskip

% \section{The Ministry of the Word}

% \pilcrow{Immediately thereafter he that readeth the Epistle shall say,}
% The Epistle [\rubric{or,} The portion of Scripture appointed for the Epistle] is written in the --- chapter of --- beginning at the --- verse. 
% \rubric{And the Epistle ended, he shall say,} Here endeth the Epistle.

% \medskip
% \centerline{\pilcrow{Here may be sung a Hymn or an Anthem.}}
% \medskip

% \pilcrow{Then the Deacon or Priest that readeth the Gospel (the people all standing up) shall say,}
% \V The Lord be with you. \R And with thy Spirit. 

% \drop{The \grealtcross\ Holy Gospel is written in the — chapter of — beginning at the — verse. \R Glory \grecross\ be to thee, O Lord.}

% \centerline{\rubric{The Gospel ended, there may be said,}}
% \centerline{Praise be to thee, O Christ.}


% \bigskip

% \pilcrow{Then shall be sung or said the Creed following, the people still standing as before: except that at the discretion of the Minister it may be omitted on any day not being a Sunday or a Holy-day.} 

% \drop{I believe in one God the Father Almighty, Maker of heaven and earth, And of all things visible and invisible:}

% And in one Lord Jesus Christ, the only-begotten son of God, Begotten of his Father before all worlds, God of God, Light of Light, Very God of very God, Begotten, not made, Being of one substance with the Father, By whom all things were made: Who for us men, and for our salvation came down from heaven, * And was incarnate by the Holy Ghost of the Virgin Mary, * And was made man, * And was crucified also for us under Pontius Pilate. He suffered and was buried, And the third day he rose again according to the Scriptures, And ascended into heaven, And sitteth on the right hand of the Father. And he shall come again with glory to judge both the quick and the dead: Whose kingdom shall have no end.

% And I believe in the Holy Ghost, The Lord and giver of life, Who proceedeth from the Father and the Son, Who with the Father and the Son together is worshipped and glorified, Who spake by the Prophets. And I believe One {Holy} Catholick and Apostolick Church. I acknowledge one Baptism for the remission of sins. And I look for the Resurrection of the dead, * And the life of the world to come. Amen.

% \bigskip
% \pilcrow{Then the Curate shall declare unto the people what Holy-days, or Fasting-days, are in the week following to be observed. And then also (if occasion be) shall notice be given of the Holy Communion, or of other services; Banns of matrimony may be published, and Briefs, Citations, and Excommunications shall be read, and Bidding of Prayers may be made. And nothing shall be proclaimed or published in the Church during the time of Divine Service, but by the Minister: nor by him any thing, but what is prescribed in the rules of this Book, or enjoined by the %Queen, or by the 
% Ordinary of the place.}% adapted toward En1928

% \smallskip

% \pilcrow{Then may follow the Sermon, or one of the Homilies already set forth, or hereafter to be set forth, by authority.}


% \bigskip

% \section{The Offertory}

% \pilcrow{Then shall the Priest return to the Lord's Table, and begin the \emph{Offertory}. The Priest shall say, or the Clerks shall sing, one of these Sentences following, or some other convenient sentence taken out of Holy Scripture. A Hymn may follow.}

% \drop{To do good, and to distribute, forget not; for with such sacrifices God is pleased.\scripture{Hebrews xiij.~16.}}

% Lay not up for yourselves treasure upon the earth; where the rust and moth doth corrupt, and where thieves break through and steal: but lay up for yourselves treasures in heaven; where neither rust nor moth doth corrupt, and where thieves do not break through and steal.\scripture{St.~Matthew vj.~19.}

% Not every one that saith unto me, Lord, Lord, shall enter into the kingdom of heaven; but he that doeth the will of my Father which is in heaven.\scripture{St.~Matthew vij.~21.}

% If we have sown unto you spiritual things, is it a great matter if we shall reap your worldly things?\scripture{1 Corithians ix.~11.}

% Do ye not know, that they who minister about holy things live of the sacrifice; and they who wait at the altar are partakers with the altar? Even so hath the Lord also ordained, that they who preach the Gospel should live of the Gospel.\scripture{1 Corithians ix.~13.}

% He that soweth little shall reap little; and he that soweth plenteously shall reap plenteously. Let every man do according as he is disposed in his heart, not grudging, or of necessity; for God loveth a cheerful giver.\scripture{2 Corithians ix.~6.}

% Let him that is taught in the Word minister unto him that teacheth, in all good things. Be not deceived, God is not mocked: for whatsoever a man soweth that shall he reap.\scripture{Galatians vj.~6.}

% While we have time, let us do good unto all men; and specially unto them that are of the household of faith.\scripture{Galatians vj.~10.}

% God is not unrighteous, that he will forget your works, and labour that proceedeth of love; which love ye have shewed for his Name’s sake, who have ministered unto the saints, and yet do minister.\scripture{Hebrews vj.~10.}

% Charge them who are rich in this world, that they be ready to give, and glad to distribute; laying up in store for themselves a good foundation against the time to come, that they may attain eternal life.\scripture{1 Timothy vj.~17.}

% Whoso hath this world’s good, and seeth his brother have need, and shutteth up his compassion from him, how dwelleth the love of God in him?\scripture{1 St.~John iij.~17}



% \pilcrow{Whilst these Sentences are said or sung, the Deacons, Church-wardens, or other fit person appointed for that purpose, shall receive the alms for the poor, or other devotions of the people, and reverently bring them to the Priest, who shall humbly present and place them upon the Holy Table in a decent bason to be provided for that purpose.}

% \bigskip

% \pilcrow{{And when there is a Communion,} the Priest shall then offer up, and place the bread and wine prepared for the Sacrament upon the Lord’s Table; and shall say,}

% \drop{Blessed be thou, O {\scshape Lord} God, for ever and ever. Thine, O {\scshape Lord}, is the greatness, and the glory, and the victory, and the majesty: for all that is in the heaven and in the earth, is thine: thine is the kingdom, O {\scshape Lord}, and thou art exalted as head above all: both riches and honour come of thee, and of thine own do we give unto thee. \R Amen.}
% \scripture{1 Chronicles xxix.~10}

% \bigskip
% \centerline{Let us pray for the whole state of Christ's Church.}

% \drop{Almighty and everliving God, who by thy holy Apostle hast taught us to make prayers, and supplications, and to give thanks for all men;}
% We humbly beseech thee most mercifully [\footnote{\rubric{If there be no alms or oblations, then the words \emph{[to accept our alms and oblations]} be left out unsaid.}}\emph{to accept our alms and \grealtcross\ oblations, and}] to receive these our prayers, which we offer unto thy Divine Majesty; beseeching thee to inspire continually the Universal Church with the spirit of truth, unity, and concord: And grant, that all they that do confess thy holy Name may agree in the truth of thy holy Word, and live in unity, and godly love.

% We beseech thee also to lead all nations into the way of righteousness and peace, and so to direct all 
% ruling authorities, that under them the world may be godly and quietly governed. 
% And grant unto all that are put in authority, 
% that they may truly and indifferently minister justice, to the punishment of wickedness and vice, and to the maintenance of thy true religion, and virtue.

% Give grace, O heavenly Father, to all Bishops, Priests, and Deacons, especially to thy servant \emph{N.} our bishop, that they may both by their life and doctrine set forth thy true and lively Word, and rightly and duly administer thy holy Sacraments.

% And to all thy people give thy heavenly grace; and specially to this congregation here present; that with meek heart and due reverence, they may hear and receive thy holy Word; truly serving thee in holiness and righteousness all the days of their life.

% And we most humbly beseech thee, of thy goodness, O Lord, to comfort and succour all them, who in this transitory life are in trouble, sorrow, need, sickness, or any other adversity.

% And we commend to thy gracious keeeping, O Lord, all thy servants departed this life in thy faith and fear, beseeching them to grant them everlasting light and peace.

% And here we give thee most high praise and hearty thanks for all thy Saints, who have been the chosen vessels of thy grace, and lights of the world in their several generations; and we pray, that rejoicing in their fellowship, and following their good examples, we may be partakers with them of thy heavenly kingdom.

% Grant this, O Father, for Jesus Christ's sake, our only Mediator and Advocate. \R Amen.

% \section{The Consecration}

% \pilcrow{Turning himself to the people the Priest shall say,}

% \V The Lord be with you. \R And with thy spirit.

% \V Lift up your hearts.  \R We lift them up unto the Lord.

% \V Let us give thanks unto our Lord God. \R It is meet and right so to do.


% \centerline{\rubric{Then shall the Priest turn to the Lord’s Table, and say,}}

% \drop{It is very meet, right, and our bounden duty, that we should at all times, and in all places, give thanks unto thee, O Lord, Holy Father, Almighty, Everlasting God.


% {\centering\rubric{Here shall follow the Proper Preface, according to the time, if there be any specially appointed, or else immediately shall follow,}\par}


% Therefore with Angels and Archangels, and with all the company of heaven, we laud and magnify thy glorious Name; evermore praising thee, and saying,}
% \smallskip

% \drop{Holy, holy, holy, Lord God of hosts, heaven and earth are full of thy glory: Glory be to thee, O Lord most High.}
% \begin{leftbar}
%     \grecross\ Blessed is he that cometh in the Name of the Lord; Hosanna in the highest.
% \end{leftbar}

% \centerline{\rubric{Then shall the priest continue thus.}}
% \drop{All glory be to thee, Almighty God, our heavenly Father, for that thou of thy tender mercy didst give thine only Son Jesus Christ to suffer death upon the Cross for our redemption; who made there (by his one oblation of himself once offered) a full, perfect, and sufficient sacrifice, oblation, and satisfaction, for the sins of the whole world; and did institute, and in his holy Gospel command us to continue, a perpetual memory of that his precious death, until his coming again;

% Hear us, O merciful Father, we most humbly beseech thee; and grant that we receiving these thy creatures of bread and wine, according to thy Son our Saviour Jesus Christ’s holy institution, in remembrance of his death and passion, may be partakers of his most blessed \grealtcross\ Body and \grealtcross\ Blood: 

% Who, in the same night that he was betrayed, \footnote{\rubric{Here the Priest is to take the Paten unto his hands:}}took Bread; and, when he had given thanks, \footnote{\rubric{And here to break the Bread:}}he brake it, and gave it to his disciples, saying, Take, eat, \footnote{\rubric{And here to lay his hand upon all the Bread.}}{\scshape this is my Body which is given for you}: Do this in remembrance of me. Likewise after supper he \footnote{\rubric{Here he is to take the Cup into his hand:}}took the Cup; and, when he had given thanks, he gave it to them, saying, Drink ye all of this; \footnote{\rubric{And here to lay his hand upon every vessel (be it Chalice or Flagon) in which there is any Wine to be consecrated.}}{\scshape for this is my Blood of the New Testament, which is shed for you and for many for the remission of sins}: Do this, as oft as ye shall drink it, in remembrance of me.}

% Wherefore, O Lord and heavenly Father, we thy humble servants, having in remembrance the precious death of thy dear Son, his mighty resurrection and glorious ascension, looking also for his coming again, do render unto thee most hearty thanks for the innumerable benefits which he hath procured unto us; and we entirely desire thy fatherly goodness mercifully to accept this our sacrifice of praise and thanksgiving; most humbly beseeching thee to grant, that by the merits and death of thy Son Jesus Christ, and through faith in his blood, we and all thy whole Church may obtain remission of our sins, and all other benefits of his passion.

% And here we offer and present unto thee, O Lord, ourselves, our souls and bodies, to be a reasonable, holy, and lively sacrifice unto thee; 
% and we pray thee of thine almighty goodness to send upon us, and upon these thy gifts, thy holy and blessed Spirit, who is the Sanctifier and the Giver of life; humbly beseeching thee, that all we, who are partakers of this holy Communion, may be fulfilled with thy grace and heavenly \grecross\ benediction. 

% And although we be unworthy, through our manifold sins, to offer unto thee any sacrifice, yet we beseech thee to accept this our bounden duty and service; not weighing our merits, but pardoning our offences;

% Through Jesus Christ our Lord; by whom, and with whom, in the unity of the Holy Ghost, all honour and glory be unto thee, O Father Almighty, world without end. \R Amen.

% \smallskip
% {\centering\footnotesize\rubric{Here shall the people join with the Priest in the Lord’s Prayer, the Priest first saying,}\par}
% As our Saviour Christ hath commanded and taught us we are bold to say,
% \drop{Our Father, which art in heaven, Hallowed be thy Name. Thy kingdom come. Thy will be done, in earth as it is in heaven. Give us this day our daily bread. And forgive us our trespasses, As we forgive them that trespass against us. And lead us not into temptation; But deliver us from evil.}
% For thine is the kingdom, the power, and the glory, For ever and ever. Amen.

% \medskip
% \begin{leftbar}
%     \pilcrow{Then may the Priest say,}
%     \drop{The Peace of the Lord be always with you.  \R And with thy spirit.}
% \end{leftbar}

% \bigskip
% \pilcrow{Then shall the Priest say to them that come to receive the holy Communion,}
% \drop{Draw near with faith, and take this Holy Sacrament to your comfort; and make your humble confession to Almighty God, meekly kneeling upon your knees.}

% \smallskip
% \rubric{Then shall be said by the Minister and people together,}
% \drop{We confess to God Almighty, the Father, the Son, and the Holy Ghost, that we have sinned in thought, word, and deed, through our own grievous fault.  Wherefore we pray God to have mercy upon us.}

% \medskip
% {\centering\footnotesize\rubric{Then shall the Priest standing up, and turning himself to the people, pronounce this Absolution.}\par}
% \drop{Almighty God have mercy upon you, \grealtcross\ forgive you all your sins, and deliver you from all evil, confirm and strengthen you in all goodness, and bring you to everlasting life; through Jesus Christ our Lord. \R Amen.}

% \centerline{\pilcrow{Then shall the Priest say,}}
% Hear what comfortable words our Saviour Christ saith unto all that truly turn to him.
% \drop{Come unto me all that travail and are heavy laden, and I will refresh you.}\scripture{St.~Matthew xj.~28}

% So God loved the world, that he gave his only-begotten Son, to the end that all that believe in him should not perish, but have everlasting life.\scripture{St.~John iij.~16}

% \centerline{Hear also what Saint Paul saith.}

% This is a true saying, and worthy of all men to be received, That Christ Jesus came into the world to save sinners.\scripture{1 Timothy i.~15.}

% \centerline{Hear also what Saint John saith.}

% If any man sin, we have an Advocate with the Father, Jesus Christ the righteous; and he is the propitiation for our sins.\scripture{1 St.~John ij.~1.}

% \medskip

% {\centering\footnotesize\rubric{Then shall the Priest, kneeling down at the Lord’s Table, say in the name of all them that shall receive the Communion this Prayer following.}\par}
% \drop{We do not presume to come to this thy Table, O merciful Lord, trusting in our own righteousness, but in thy manifold and great mercies. We are not worthy so much as to gather up the crumbs under thy Table. But thou art the same Lord, whose property is always to have mercy: Grant us therefore, gracious Lord, so to eat the flesh of thy dear Son Jesus Christ, and to drink his blood, that our sinful bodies may be made clean by his body, and our souls washed through his most precious blood, and that we may evermore dwell in him, and he in us. Amen.}

% \medskip

% \pilcrow{Then shall the Minister first receive the Communion in both kinds himself, and then proceed to deliver the same to the Bishops, Priests, and Deacons, in like manner, (if any be present,) and after that to the people also in order, into their hands, all meekly kneeling. And, when he delivereth the Bread to any one, he shall say,}
% \drop{The Body of our Lord Jesus Christ, which was given for thee, preserve thy body and soul unto everlasting life. \R Amen.}

% \smallskip
% {\centering\footnotesize\rubric{And the Minister that delivereth the Cup to any one shall say,}\par}
% \drop{The Blood of our Lord Jesus Christ, which was shed for thee, preserve thy body and soul unto everlasting life.}

% \medskip
% \centerline{\pilcrow{Here may be sung a Hymn or an Anthem.}}
% \medskip

% {\footnotesize\rubric{When all have communicated, the Minister shall return to the Lord’s Table, and reverently place upon it what remaineth of the consecrated Elements, covering the same with a fair linen cloth.}\par}

% \medskip
% {\centering\footnotesize\rubric{Then the Priest shall say,}\par} 
% \drop{Almighty and everliving God, we most heartily thank thee, for that thou dost vouchsafe to feed us, who have duly received these holy mysteries, with the spiritual food of the most precious Body and Blood of thy Son our Saviour Jesus Christ; and dost assure us thereby of thy favour and goodness towards us; and that we are very members incorporate in the mystical body of thy Son, which is the blessed company of all faithful people; and are also heirs through hope of thy everlasting kingdom, by the merits of the most precious death and passion of thy dear Son. And we most humbly beseech thee, O heavenly Father, so to assist us with thy grace, that we may continue in that holy fellowship, and do all such good works as thou hast prepared for us to walk in; through Jesus Christ our Lord, to whom, with thee and the Holy Ghost, be all honour and glory, world without end. Amen.}

% \medskip
% \pilcrow{Then, on Sundays and on Feast days (except in Advent and from Septuagesima to Palm Sunday inclusive), shall be said or sung as follows:}

% \drop{Glory be to God on high, and in earth peace, good will towards men. We praise thee, we bless thee, \footnote{Bow} we worship thee, we glorify thee, we give thanks to thee for thy great glory, O Lord God, heavenly King, God the Father Almighty.}

% O Lord, the only-begotten Son Jesu Christ; O Lord God, Lamb of God, Son of the Father, that takest away the sins of the world, have mercy upon us. 
% Thou that takest away the sins of the world, * receive our prayer. Thou that sittest at the right hand of God the Father, have mercy upon us.

% For thou only art holy; thou only art the Lord; thou only, * O {Jesu} Christ, with the Holy Ghost, art most high \grecross\ in the glory of God the Father. Amen.


% {\centering\rubric{This hymn may be omitted here, and sung instead at the beginning of this Order after the \emph{Kyrie eleison}.}\par}


% \bigskip
% {\centering\rubric{Then the Priest shall let them depart with this Blessing.}\par}
% \drop{The peace of God, which passeth all understanding, keep your hearts and minds in the knowledge and love of God, and of his son Jesus Christ our Lord: and the blessing of God Almighty, the Father, \grealtcross\ the Son, and the Holy Ghost, be amongst you and remain with you always. \R Amen.}

% {\centering\rubric{Or, if there be no congregation,}\par}
% \drop{In the Name of the Father, \grealtcross\ and of the Son, and of the Holy Ghost. \R Amen.}

