
\chapter{The Form of Solemnization of Matrimony}

\pilcrow{The laws respecting Matrimony
%, whether by publishing the Banns in Churches, or by Licence, 
being different in the several States, every Minister is left to the direction of those laws, in every thing that regards the civil contract between the parties.} %1928 American

\medskip
% And when the Banns are published, it shall be in the following form: I publish the Banns of Marriage between N. of —, and N. of —. If any of you know cause, or just impediment, why these two persons should not be joined together in holy Matrimony, ye are to declare it. This is the first [second or third] time of asking. %1928 American

\pilcrow{First, the Banns of all that are to be married together, 
% must %1662, etc
should % ICCC
be published in the Church three several Sundays, or Holy-days in the time of Divine Service, immediately before the the sentences for the Offertory: the 
% Curate 1662
Priest saying after the accustomed manner.} % Parson's Handbook

I publish the Banns of Marriage between \emph{N.} of \emph{this Parish} and \emph{N.} of —. If any of you know cause, or just impediment, why these two persons should not be joined together in holy Matrimony, ye are to declare it.  This is the first [second \rubric{or} third] time of asking.% Parson's Handbook

% I publish the Banns of Marriage between M. of ----- and N. of -----. If any of you know cause, or just impediment, why these two persons should not be joined together in holy Matrimony, ye are to declare it. This is the first [second, or third] time of asking.


% \pilcrow{And if the persons that are to be married dwell in divers Parishes, the Banns must be asked in both Parishes; and the Curate of the one Parish shall not solemnize Matrimony betwixt them, without a Certificate of the Banns being thrice asked, from the Curate of the other Parish.}

\section{The Introduction}

\pilcrow{At the day and time appointed for solemnization of Matrimony, the persons to be married shall come 
into the body of the Church %1662 (Piercy - chancel gate)
% to the Door of the Church, or some other convenient location, %ICCC
with their friends and neighbours: and there standing together, [the Man on the right hand, and the Woman on the left], the Priest shall say,}

\drop{Dearly beloved, we are gathered together here in the sight of God, and in the face of this congregation, to join together \emph{this Man and this Woman} in holy Matrimony; which is an honourable estate, instituted of God}
% in the time of man’s innocency, %1662, scottish
himself, %1928 proposed
signifying unto us the mystical union that is betwixt Christ and his Church; which holy estate Christ adorned and beautified with his presence, and first miracle that he wrought, in Cana of Galilee; and is commended of Saint Paul to be honourable among all men: and therefore is not by any to be enterprised, nor taken in hand, unadvisedly, lightly, or wantonly,
%to satisfy men's carnal lusts and appetites, like brute beasts that have no understanding; 1662; removed in 1928 proposed
but reverently, discreetly, advisedly, soberly, and in the fear of God;
\begin{leftbar}
duly considering the 
chief %scottish
causes for which Matrimony was ordained.

% First, 
It was ordained for the 
%procreation of children, to %1662
increase of mankind according to the will of God, and that children might %1912
be brought up in the fear and nurture of the Lord, and to the praise of his holy Name.

% Secondly, It was ordained for a remedy against sin, and to avoid fornication; that such persons as have not the gift of continency might marry, and keep themselves undefiled members of Christ’s body.

% Thirdly, 
It was 
also %1912
ordained for the mutual society, help, and comfort, that the one ought to have of the other, both in prosperity and adversity.

Canada 1918 has a one-sentence combination.:
Matrimony was ordained for the hallowing of the union betwixt man and woman; for the procreationof children to be brought up in the fear and nurtureof the Lord; and for the mutual society, help, and comfort, that the one ought to have of the other, in
both prosperity and adversity. 

1928 removes all this.
\end{leftbar}

Into which holy estate these two persons present come now to be joined.

Therefore if any man can shew any just cause, why they may not lawfully be joined together, let him now speak, or else hereafter for ever hold his peace.

\smallskip
\pilcrow{And also, speaking unto the persons that shall be married, he shall say,}
\drop{I require and charge you both, as ye will answer at the dreadful day of judgement when the secrets of all hearts shall be disclosed, that if either of you know any impediment, why ye may not be lawfully joined together in Matrimony, ye do now confess it.  For be ye well assured, that so many as are coupled together otherwise than God’s Word doth allow
[other:] are not joined together by God; neither is their Matrimony lawful.
 
[US1928]:  their marriage is not lawful.

}


% \pilcrow{At which day of Marriage, if any man do allege and declare any impediment, why they may not be coupled together in Matrimony, by God’s law, or the laws of this Realm; and will be bound, and sufficient sureties with him, to the parties; or else put in a caution (to the full value of such charges as the persons to be married do thereby sustain) to prove his allegation: then the solemnization must be deferred, until such time as the truth be tried.} %1662

{\footnotesize\rubric{The Minister, if he shall have reason to doubt of the lawfulness of the proposed Marriage, may demand sufficient surety for his indemnification.}\par} % 1928

\section{The Marriage}

\pilcrow{If no impediment be alleged, then shall the Curate say unto the \emph{Man},}
\lettrine{\emph{N.}}{ wilt} thou have this \emph{woman} to thy wedded \emph{wife}, to live together after God’s ordinance in the holy estate of Matrimony? Wilt thou love \emph{her}, comfort \emph{her}, honour, and keep \emph{her} in sickness and in health; and, forsaking all other, keep thee only unto \emph{her}, so long as ye both shall live?

\centerline{\rubric{The \emph{Man} shall answer,} I will.}

\medskip

\centerline{\rubric{Then shall the Priest say unto the \emph{Woman},}}
\lettrine{\emph{N.}}{ wilt} thou have this \emph{man} to thy wedded \emph{husband}, to live together after God’s ordinance in the holy estate of Matrimony? Wilt thou [obey \emph{him}, and serve \emph{him}], love [him, comfort him], honour, and keep \emph{him} in sickness and in health; and, forsaking all other, keep thee only unto \emph{him}, so long as ye both shall live?
\centerline{\rubric{The \emph{Woman} shall answer,} I will.}


\begin{leftbar}
\centerline{\rubric{Then shall the Priest say,}} %1928 proposed
\centerline{Who giveth this woman to be married to this man?}
\end{leftbar}


\pilcrow{Then shall they give their troth to each other in this manner.}

\smallskip

\pilcrow{The Minister, [receiving the Woman at her father’s or friend’s hands,] shall cause the \emph{Man} with \emph{his} right hand to take the \emph{Woman} by \emph{her} right hand, and to say after him as followeth.}
\drop{I, \emph{N.} take thee \emph{N.} to my wedded \emph{wife}, to have and to hold from this day forward, for better for worse, for richer for poorer, in sickness and in health, to love and to cherish, till death us do part, according to God’s holy ordinance; and thereto I plight thee my troth.}

\pilcrow{Then shall they loose their hands; and the \emph{Woman}, with \emph{her} right hand taking the \emph{Man} by \emph{his} right hand, shall likewise say after the Minister,}
\drop{I, \emph{N.} take thee \emph{N.} to my wedded \emph{husband}, to have and to hold from this day forward, for better for worse, for richer for poorer, in sickness and in health, to love, [and to] cherish,[ and to obey], till death us do part, according to God’s holy ordinance; and thereto I give thee my troth.}

\bigskip
\pilcrow{Then shall they again loose their hands; and the Man shall give unto the Woman a Ring, [and other tokens of spousage, as gold or silver,] laying the same upon the book.  And the Priest, taking the ring, shall
%  with the accustomed duty to the Priest and Clerk %1662
say,} %Scottish, 1929
\drop{Bless this ring, \grealtcross\ O merciful Lord, that he who giveth and she who weareth it may ever be faithful one to another; through Jesus Christ our Lord. \R Amen} %1929

\pilcrow{Then shall he %scottish 1929
deliver the ring unto the Man, to put it upon the fourth finger of the Woman’s left hand. And the Man holding the Ring there, and taught by the Priest, shall say,}
\drop{With this ring I thee wed, [this gold and silver I thee give,] with my body I thee worship, and with all my worldly goods I thee endow: In the Name of the Father, and of the Son, and of the Holy Ghost. Amen.}



\bigskip
\pilcrow{Then the Man leaving the Ring upon the fourth finger of the Woman’s left hand, they shall both kneel down; and the Priest shall say,} %1928 proposed
\centerline{Let us pray.}
\drop{O eternal God, Creator and Preserver of all mankind, Giver of all spiritual grace, the Author of everlasting life: Send thy blessing upon these thy servants, \emph{this man and this woman}, whom we bless in thy Name; that, 

as Isaac and Rebecca (after bracelets and jewels of gold given of the one to the other for tokens of their matrimony) lived faithfully together %1549 
% as Isaac and Rebecca lived faithfully together, so these persons %1662
living faithfully together, they %1929


may surely perform and keep the vow and covenant betwixt them made, (whereof this Ring given and received is a token and pledge,) and may ever remain in perfect love and peace together, and live according to thy laws; through Jesus Christ our Lord. \R Amen.}

\pilcrow{Then shall the Priest join their right hands together, and say,}
{\centering Those whom God hath joined together let no man put asunder.\par}

\centerline{\rubric{Then shall the Minister speak unto the people.}}
\drop{Forasmuch as \emph{N.} and \emph{N.} have consented together in holy wedlock, and have witnessed the same before God and this company, and thereto have given and pledged their troth either to other, and have declared the same by giving and receiving of a Ring, and by joining of hands; I pronounce that they be \emph{Man and Wife} together, In the Name of the Father, and of the Son, and of the Holy Ghost. Amen.}

\centerline{\pilcrow{And the priest shall add this Blessing.}}
\drop{God the Father, \grealtcross\ God the Son, God the Holy Ghost, bless, preserve, and keep you; the Lord mercifully with his favour look upon you; and so fill you with all spiritual benediction and grace, that ye may so live together in this life, that in the world to come ye may have life everlasting. \R Amen.}

\section{The Benediction}
\pilcrow{Then the Minister or Clerks, going to the Lord’s Table, shall say or sing this Psalm following.}
\subsection[{Psalm 128}]{\stylesubsec{Psalm 128.}{Beati omnes.}{}}
\drop{Blessed are all they that fear the {\scshape Lord}, * and walk in his ways.}

2\enspace For thou shalt eat the labours of thine hands: * O well is thee, and happy shalt thou be.

3\enspace Thy wife shall be as the fruitful vine * upon the walls of thine house.

4\enspace Thy children like the olive-branches * round about thy table.

5\enspace Lo, thus shall the man be blessed * that feareth the {\scshape Lord}.

6\enspace The {\scshape Lord} from out of Sion shall so bless thee, * that thou shalt see Jerusalem in prosperity all thy life long;

7\enspace Yea, that thou shalt see thy children’s children, * and peace upon Israel.

Glory be to the Father, and to the Son, \star\  and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be, \star\  world without end. Amen.

\medskip
\centerline{\rubric{Or this Psalm,}}
\subsection[{Psalm 67}]{\stylesubsec{Psalm 67.}{Deus misereatur.}{}}
\drop{God be merciful unto us, and bless us, \star\ and shew us the light of his countenance, and be merciful unto us:}

2\enspace That thy way may be known upon earth, \star\ thy saving health among all nations.

3\enspace Let the people praise thee, O God; \star\ yea, let all the people praise thee.

4\enspace O let the nations rejoice and be glad; \star\ for thou shalt judge the folk righteously, and govern the nations upon earth.

5\enspace Let the people praise thee, O God; \star\ let all the people práise thee.

6\enspace Then shall the earth bring forth her íncrease; \star\ and God, even our own God, shall gíve us his blessing.

7\enspace God shall bless us; \star\ and all the ends of the world shall fear him.

Glory be to the Father, and to the Son, \star\  and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be, \star\  world without end. Amen.


\medskip
\centerline{\rubric{Or this Psalm,}}
\subsection[{Psalm 133}]{\stylesubsec{Psalm 133.}{Ecce, quam bonum!}{}}
\drop{Behold, how good and joyful a thing it is, \star\ brethren, to dwell together in unity!}

2\enspace It is like the precious ointment upon the head, that ran down unto the beard, \star\ even unto Aaron’s beard, and went down to the skirts of his clothing.

3\enspace Like as the dew of Hermon, \star\ which fell upon the hill of Sion.

4\enspace For there the {\scshape Lord} promised his blessing, \star\ and life for evermore.

Glory be to the Father, and to the Son, \star\  and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be, \star\  world without end. Amen.

\medskip
\centerline{\rubric{Or this Canticle,}}

\subsection[{The Song of Ruth}]{\stylesubsec{The Song of Ruth.}{Ne adverseris mihi.}{Ruth i.}}
\drop{Intreat me not to leave thee, \star\  or to return from following after thee:}

For whither thou goest, I will go; \star\ and where thou lodgest, I will lodge:

Thy people shall be my people, \star\ and thy God my God:

Where thou diest, will I die, \star\ and there will I be buried:

The {\scshape Lord} do so to me, and more also,  \star\  if ought but death part thee and me.

Glory be to the Father, and to the Son, \star\  and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be, \star\  world without end. Amen.

\bigskip

\pilcrow{The Psalm ended, and the \emph{Man and the Woman} kneeling before the Lord’s Table, the Priest standing at the Table, and turning his face towards them, shall say,}
\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\medskip

\ourFather

\smallskip

\V O Lord, save thy \emph{servant, and thy handmaid}; \R Who put their trust in thee.

\V O Lord, send them help from thy holy place; \R And evermore defend them.

\V Be unto them a tower of strength, \R From the face of their enemy.

\V O Lord, hear our prayer. \R And let our cry come unto thee.

\centerline{Let us pray.}
\drop{O God of Abraham, God of Isaac, God of Jacob, bless these thy servants, and sow the seed of eternal life in their hearts; that whatsoever in thy holy Word they shall profitably learn, they may in deed fulfil the same.}


Look, O Lord, mercifully upon them from heaven, and bless them. 
And as thou didst send thy Angel Raphael to Toby and Sarah, the daughter of Raguel, to their great comfort, so vouchsafe... %1549


And as thou didst send thy blessing upon Abraham and Sarah, to their great comfort, so vouchsafe to send thy blessing upon these thy servants; %1662, omitted in 1928 proposed


that they obeying thy Will, and alway being in safety under thy protection, may abide in thy love unto their lives’ end; through Jesus Christ our Lord. \R Amen.

Or, offer alternatives to Abraham and Sarah?
Ruth and Naomi
David and Jonathan
Mary and Martha
Peter and Paul
Sergius and Bacchus
Perpetua and Felicity

Polyeuct and Nearchus
Symeon of Emessa and John
Vivaldo and Bartolo
Avertanus and Romeo
Brigid and Darlughdach
Protus and Hyacinth, martyrs


\begin{leftbar}
\pilcrow{This Prayer next following shall be omitted, where the woman is past child-bearing.}
% \drop{O merciful Lord, and heavenly Father, by whose gracious gift mankind is increased: We beseech thee, assist with thy blessing these two persons, that they may both be fruitful in procreation of children, and also live together so long in godly love and honesty, that they may see their children Christianly and virtuously brought up, to thy praise and honour; through Jesus Christ our Lord. \R Amen.} %1928 proposed alters toward the below
% 1928 american, 1912, 1929 scottish
\drop{O Almighty God, Creator of mankind, who only art the well-spring of life; Bestow upon these thy servants, if it be thy will, the gift and heritage of children; and grant that they may see their children brought up in thy faith and fear, to the honour and glory of thy Name; through Jesus Christ our Lord. \R Amen.}
\end{leftbar}

\centerline{\pilcrow{This Prayer shall follow.}}
% 1928 american
\drop{O God, who hast so consecrated the state of Matrimony that in it is represented the spiritual marriage and unity betwixt Christ and his Church; Look mercifully upon these thy servants, that they may love, honour, and cherish each other, and so live together in faithfulness and patience, in wisdom and true godliness, that their home may be a haven of blessing and of peace; through the same Jesus Christ our Lord, who liveth and reigneth with thee and the Holy Spirit ever, one God, world without end. \R Amen.}

% \drop{O God, who by thy mighty power hast made all things of nothing; who also (after other things set in order) didst appoint, that out of man (created after thine own image and similitude) woman should take her beginning; and, knitting them together, didst teach that it should never be lawful to put asunder those whom thou by Matrimony hadst made one: O God, who hast consecrated the state of Matrimony to such an excellent mystery, that in it is signified and represented the spiritual marriage and unity betwixt Christ and his Church: Look mercifully upon these thy servants, that both this man may love his wife, according to thy Word, (as Christ did love his spouse the Church, who gave himself for it, loving and cherishing it even as his own flesh,) and also that this woman may be loving and amiable,[ faithful and obedient to her husband]; or [to her husband as Rachel, wise as Rebecca, faithful and obedient as Sara;] and in all quietness, sobriety, and peace, be a follower of holy and godly matrons. O Lord, bless them both, and grant them to inherit thy everlasting kingdom; through Jesus Christ our Lord. \R Amen.}  %1662. Cut down and slightly altered in 1928 proposed and scottish; completely replaced in the american 1928 with the above.


\centerline{\pilcrow{Then shall the Priest say
this Blessing,}} %1928 proposed
\drop{Almighty God, 
% who at the beginning did create our first parents, Adam and Eve, and did sanctify and join them together in marriage; %1662
the Father of our Lord Jesus Christ, %1928 proposed
Pour upon you the riches of his grace, sanctify and \grealtcross\ bless you, that ye may please him both in body and soul, and live together in holy love unto your lives’ end. \R Amen.}

\pilcrow{If there be a Communion, the foregoing Prayer and Blessing shall be said over the \emph{man and woman} immediately before the final Blessing of the congregation at the Communion.}

\bigskip

\pilcrow{If there be no Communion, nor sermon declaring the duties of man and wife, there shall be read some portion of Scripture, or the Priest shall say the Exhortation appended to this Liturgy.  And then the Priest shall dismiss those that are gathered together, saying,}

\centerline{Let us pray.}
\drop{O almighty Lord, and everlasting God, vouchsafe, we beseech thee, to direct, sanctify, and govern, both our hearts and bodies, in the ways of thy laws, and in the works of thy commandments; that through thy most mighty protection, both here and ever, we may be preserved in body and soul; through our Lord and Saviour Jesus Christ. \R Amen.}
\medskip
\drop{The Blessing of God Almighty, \grealtcross\ the Father, the Son, and the Holy Ghost, be amongst you and remain with you always. \R Amen.}


\section{The Communion}

\pilcrow{If there be a Communion, the following Collect, Epistle, and Gospel may be used, immediately after the prayer for child-bearing; or after the Collect preceding it, if that be not said.}

\subsection{\stylesubsec{}{The Collect.}{}}
\drop{O God our Father, who by thy holy apostle hast taught us that love is the fulfilling of the law: Grant to these thy \emph{servants} that, loving one another, they may continue in thy love unto their lives’ end; through Jesus Christ our Lord, who liveth and reigneth with thee in the unity of the Holy Ghost, one God world without end. \R Amen.}

\subsection{\stylesubsec{}{The Epistle.}{Ephesians 3.~14.}}
\drop{For this cause I bow my knees unto the Father of our Lord Jesus Christ, of whom the whole family in heaven and earth is named, that he would grant you, according to the riches of his glory, to be strengthened with might by his Spirit in the inner man; that Christ may dwell in your hearts by faith; that ye, being rooted and grounded in love, May be able to comprehend with all saints what is the breadth, and length, and depth, and height; and to know the love of Christ, which passeth knowledge, that ye might be filled with all the fulness of God.}

\subsection{\stylesubsec{}{The Gospel.}{St.~John 15.~9.}}
\drop{As the Father hath loved me, so have I loved you: continue ye in my love.  If ye keep my commandments, ye shall abide in my love; even as I have kept my Father’s commandments, and abide in his love. These things have I spoken unto you, that my joy might remain in you, and that your joy might be full.  This is my commandment, that ye love one another, as I have loved you.}


{\footnotesize\rubric{The Priest shall then continue the Order of Holy Communion, the new-married persons remaining before the Holy Table until the end of the service. The last prayer and blessing, from \emph{The Benediction}, shall be said over them immediately before \emph{The peace of God,} \etc}\par}



% \section{An Exhortation}

% \drop{All ye that are married, or that intend to take the holy estate of Matrimony upon you, hear what the holy Scripture doth say as touching the duty of husbands towards their wives, and wives towards their husbands.}

% Saint Paul, in his Epistle to the Ephesians, the fifth Chapter, doth give this commandment to all married men; Husbands, love your wives, even as Christ also loved the Church, and gave himself for it, that he might sanctify and cleanse it with the washing of water, by the Word; that he might present it to himself a glorious Church, not having spot, or wrinkle, or any such thing; but that it should be holy, and without blemish. So ought men to love their wives as their own bodies. He that loveth his wife loveth himself: for no man ever yet hated his own flesh, but nourisheth and cherisheth it, even as the Lord the Church: for we are members of his body, of his flesh, and of his bones. For this cause shall a man leave his father and mother, and shall be joined unto his wife; and they two shall be one flesh. This is a great mystery; but I speak concerning Christ and the Church. Nevertheless, let every one of you in particular so love his wife, even as himself.


% Likewise the same Saint Paul, writing to the Colossians, speaketh thus to all men that are married; Husbands, love your wives, and be not bitter against them.

% Hear also what Saint Peter, the Apostle of Christ, who was himself a married man, saith unto them that are married; Ye husbands, dwell with your wives according to knowledge; giving honour unto the wife, as unto the weaker vessel, and as being heirs together of the grace of life, that your prayers be not hindered.

% Hitherto ye have heard the duty of the husband toward the wife. Now likewise, ye wives, hear and learn your duties toward your husbands, even as it is plainly set forth in holy Scripture.

% Saint Paul, in the aforenamed Epistle to the Ephesians, teacheth you thus; Wives, submit yourselves unto your own husbands, as unto the Lord. For the husband is the head of the wife, even as Christ is the head of the Church: and he is the Saviour of the body. Therefore as the Church is subject unto Christ, so let the wives be to their own husbands in every thing. And again he saith, Let the wife see that she reverence her husband.

% And in his Epistle to the Colossians, Saint Paul giveth you this short lesson; Wives, submit yourselves unto your own husbands, as it is fit in the Lord.

% Saint Peter also doth instruct you very well, thus saying; Ye wives, be in subjection to your own husbands; that, if any obey not the Word, they also may without the Word be won by the conversation of the wives; while they behold your chaste conversation coupled with fear. Whose adorning, let it not be that outward adorning of plaiting the hair, and of wearing of gold, or of putting on of apparel; but let it be the hidden man of the heart, in that which is not corruptible; even the ornament of a meek and quiet spirit, which is in the sight of God of great price. For after this manner in the old time the holy women also, who trusted in God, adorned themselves, being in subjection unto their own husbands; even as Sarah obeyed Abraham, calling him lord; whose daughters ye are as long as ye do well, and are not afraid with any amazement.


\bigskip
\section{The Blessing of a Civil Marriage}

\pilcrow{If persons, having been married by civil authorities, desire to receive the blessing of the Church upon their union, the Priest shall demand the certificate of the Civil Marriage; and if he is satisfied that they are truly married, and that their marriage is not contrary to the Canons and Rules of the Church, he shall proceed in the following form.}

\rubric{The man and the woman standing before the altar, the Priest shall say to them,}

\drop{Forasmuch as ye have been already joined together in marriage, and have now come to ask the Church’s blessing thereon: I shall now ask you both to promise by the help of God to fulfil the obligations which Christian marriage demands.}


\smallskip
\centerline{\rubric{Then shall he say to the man,}}

\lettrine{\emph{N.}}{ doest} thou acknowledge this \emph{woman} as thy wedded \emph{wife}, to live together after God’s ordinance in the holy estate of Matrimony? 

\centerline{\R I do.}

\drop{Wilt thou love \emph{her}, comfort \emph{her}, honour, and keep \emph{her} in sickness and in health; and, forsaking all other, keep thee only unto \emph{her}, so long as ye both shall live?}

\centerline{\R I will.}

\smallskip

\centerline{\rubric{Then shall the Priest say unto the woman,}}

\lettrine{\emph{N.}}{ doest} thou acknowledge this \emph{man} as thy wedded \emph{husband}, to live together after God’s ordinance in the holy estate of Matrimony? 

\centerline{\R I do.}

\drop{Wilt thou love \emph{him}, comfort \emph{him}, honour, and keep \emph{him} in sickness and in health; and, forsaking all other, keep thee only unto \emph{him}, so long as ye both shall live?}

\centerline{\R I will.}


\medskip
\pilcrow{Then shall the \emph{man} and the \emph{woman} kneel down, and the Priest shall say this Blessing:}
\drop{God the Father, \grealtcross\ God the Son, God the Holy Ghost, bless, preserve, and keep you; the Lord mercifully with his favour look upon you; and so fill you with all spiritual benediction and grace, that ye may so live together in this life, that in the world to come ye may have life everlasting. \R Amen.}


\medskip
\centerline{Let us pray.}
\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\medskip

\ourFather

\smallskip

\V O Lord, save thy \emph{servant, and thy handmaid}; \R Who put their trust in thee

\V O Lord, send them help from thy holy place; \R And evermore defend them.

\V Be unto them a tower of strength, \R From the face of their enemy.

\V O Lord, hear our prayer. \R And let our cry come unto thee.

\centerline{Let us pray.}

\drop{O God of Abraham, God of Isaac, God of Jacob, bless these thy servants, and sow the seed of eternal life in their hearts; that whatsoever in thy holy Word they shall profitably learn, they may in deed fulfil the same; that they obeying thy Will, and alway being in safety under thy protection, may abide in thy love unto their lives’ end; through Jesus Christ our Lord. \R Amen.}


\begin{leftbar}
\pilcrow{The following Prayer shall be added, if it be deemed fitting:}
\drop{O Almighty God, Creator of mankind, who only art the well-spring of life; Bestow upon these thy servants, if it be thy will, the gift and heritage of children; and grant that they may see their children brought up in thy faith and fear, to the honour and glory of thy Name; through Jesus Christ our Lord. \R Amen.}
\end{leftbar}

% The following in the Roman use is said after the fraction: where the blessing, or pontifical blessing, or comfortable words are otherwise said.
\centerline{\pilcrow{This Prayer shall follow.}}
\drop{O God, who hast so consecrated the state of Matrimony that in it is represented the spiritual marriage and unity betwixt Christ and his Church; Look mercifully upon these thy servants, that they may love, honour, and cherish each other, and so live together in faithfulness and patience, in wisdom and true godliness, that their home may be a haven of blessing and of peace; through the same Jesus Christ our Lord, who liveth and reigneth with thee and the Holy Spirit ever, one God, world without end. \R Amen.}
\smallskip

\centerline{\pilcrow{Then shall the Priest say this Blessing,}}
\drop{Almighty God, the Father of our Lord Jesus Christ, Pour upon you the riches of his grace, sanctify and \grealtcross\ bless you, that ye may please him both in body and soul, and live together in holy love unto your lives’ end. \R Amen.}

\pilcrow{If there be a Communion, the foregoing Prayer and Blessing shall be said over the \emph{man and woman} immediately before the final Blessing of the congregation at the Communion.}

\bigskip

\pilcrow{lt is very desirable that the persons whose marriage is thus blessed should receive the Holy Communion immediately after the ceremony.}

\smallskip
\pilcrow{If there is a special celebration of the Lord's Supper at the time of the Blessing, the Priest shall use the Collect, Epistle, and Gospel appointed at the end of the Marriage Service; and the foregoing Prayer and Blessing shall be said over the man and woman immediately before the final blessing of the congregation at the Communion.}

\smallskip
\pilcrow{If there be no Communion, the Priest shall dismiss those that are gathered together, saying,}

\centerline{Let us pray.}
\drop{O almighty Lord, and everlasting God, vouchsafe, we beseech thee, to direct, sanctify, and govern, both our hearts and bodies, in the ways of thy laws, and in the works of thy commandments; that through thy most mighty protection, both here and ever, we may be preserved in body and soul; through our Lord and Saviour Jesus Christ. \R Amen.}
\medskip
\drop{The Blessing of God Almighty, \grealtcross\ the Father, the Son, and the Holy Ghost, be amongst you and remain with you always. \R Amen.}


[Note.—No entry of such a Blessing of a Civil Marriage shall be made in a Marriage Register. A record of every such Blessing shall be sent forthwith to the Bishop in the form prescribed.]


\fleuron


