\chapter[An Introduction]{An Introduction to Morning or Evening Prayer}

% This rite is made optional/occasional; and so needs none of the shortenings
% provided in the various books.

\bigskip

% Only penitential sentences are provided, since this is a penitential opening,
% not the "general" beginning of mattins.
\pilcrow{The Minister shall read with a loud voice some one or more of these Sentences of the Scriptures that follow.}

\drop{When the wicked man turneth away from his wickedness that he hath committed, and doeth that which is lawful and right, he shall save his soul alive.}
\scripture{Ezekiel xviii.~27.}

I acknowledge my transgressions, and my sin is ever before me.\scripture{Psalm lj.~3.}

Hide thy face from my sins, and blot out all mine iniquities.\scripture{Psalm lj.~9.}

The sacrifices of God are a broken spirit: a broken and a contrite heart, O God, thou wilt not despise.\scripture{Psalm lj.~17.}

Rend your heart, and not your garments, and turn unto the Lord your God: for he is gracious and merciful, slow to anger, and of great kindness, and repenteth him of the evil.\scripture{Joel ij.~13.}

To the Lord our God belong mercies and forgivenesses, though we have rebelled against him; neither have we obeyed the voice of the Lord our God, to walk in his laws which he set before us.\scripture{Daniel ix.~9, 10.}

O Lord, correct me, but with judgement; not in thine anger, lest thou bring me to nothing.\par\scripture{Jeremiah x.~24, Psalm vj.~1.}

Repent ye; for the Kingdom of Heaven is at hand.\scripture{St.~Matthew iij.~2.}

I will arise and go to my father, and will say unto him, Father, I have sinned against heaven, and before thee, and am no more worthy to be called thy son.\scripture{St.~Luke xv.~18, 19.}

Enter not into judgement with thy servant, O Lord; for in thy sight shall no man living be justified.\scripture{Psalm cxliij.~2.}

If we say that we have no sin, we deceive ourselves, and the truth is not in us; but if we confess our sins, God is faithful and just to forgive us our sins, and to cleanse us from all unrighteousness.\scripture{1 St.~John i.~8, 9.}

\bigskip


\centerline{\pilcrow{Then the Minister shall say,}}

\drop{Dearly beloved brethren, the Scripture moveth us in sundry places to acknowledge and confess our manifold sins and wickedness; and that we should not dissemble nor cloak them before the face of Almighty God our heavenly Father; but confess them with an humble, lowly, penitent, and obedient heart; to the end that we may obtain forgiveness of the same, by his infinite goodness and mercy. And although we ought at all times humbly to acknowledge our sins before God; yet ought we most chiefly so to do, when we assemble and meet together to render thanks for the great benefits that we have received at his hands, to set forth his most worthy praise, to hear his most holy Word, and to ask those things which are requisite and necessary, as well for the body as the soul. Wherefore I pray and beseech you, as many as are here present, to accompany me with a pure heart, and humble voice, unto the throne of the heavenly grace, saying after me;} %proofread vs 1928 proposed


\bigskip

{\centering\pilcrow{A general Confession to be said of the whole Congregation after the Minister, all kneeling.}}

\drop{Almighty and most merciful Father; \ We have erred, and strayed from thy ways like lost sheep. We have followed too much the devices and desires of our own hearts. We have offended against thy holy laws. We have left undone those things which we ought to have done; \ And we have done those things which we ought not to have done; \ And there is no health in us. But thou, O Lord, have mercy upon us, miserable offenders. Spare thou them, O God, which confess their faults. Restore thou them that are penitent; \ According to thy promises declared unto mankind in Christ Jesu our Lord. And grant, \ O most merciful Father, for his sake; \ That we may hereafter live a godly, righteous, and sober life, \ To the glory of thy holy Name. Amen.} %proofread vs 1928 proposed

\medskip

{\centering\pilcrow{The Absolution, or Remission of sins, to be pronounced by the Priest alone, standing; the people still kneeling.}}

\drop{Almighty God, the Father of our Lord Jesus Christ, who desireth not the death of a sinner, but rather that he may turn from his wickedness, and live; and hath given power, and commandment, to his Ministers, to declare and pronounce to his people, being penitent, the Absolution and Remission of their sins: He pardoneth \grealtcross\ and absolveth all them that truly repent, and unfeignedly believe his holy Gospel. Wherefore let us beseech him to grant us true repentance, and his Holy Spirit, that those things may please him, which we do at this present; and that the rest of our life hereafter may be pure, and holy; so that at the last we may come to his eternal joy; through Jesus Christ our Lord.} %proofread vs 1928 proposed
\centerline{\rubric{The people shall answer here, and at the end of all other prayers,}}
\centerline{\R Amen.}

\medskip

\pilcrow{If no priest be present the person saying the service shall read the Collect for the Twenty-First Sunday after Trinity, that person and the people still kneeling.}


\fleuron

\chapter[The Order for Mattins]{\stylechapter{The Order for}{Mattins}{Daily Throughout the Year.}}

\bigskip

\pilcrow{The Minister shall kneel, and say the Lord's Prayer with an audible voice; the people also kneeling, and repeating it with him, both here, and wheresoever else it is used in Divine Service.}
\newcommand\ourFather{\drop{Our Father, which art in heaven, \ Hallowed be thy Name; \ Thy kingdom come; \ Thy will be done; \ In earth as it is in heaven. Give us this day our daily bread. And forgive us our trespasses, \ As we forgive them that trespass against us. And lead us not into temptation; \ But deliver us from evil. Amen.}}
\ourFather


\medskip
\pilcrow{Here all standing up, the Priest shall say,}

\V O Lord, open thou my lips.  \R And my mouth shall shew forth thy praise.

\drop{O God, \cross make speed to save me.  \R O Lord, make haste to help me.}

\V Glory be to the Father, and to the Son, and to the Holy Ghost;  \R As it was in the beginning, is now, and ever shall be, world without end. Amen.

\centerline{Alleluya.}

\centeredrubric{From First Evensong of Septuagesima until Easter, instead of \emph{\black Alleluya} is said:}

\V Praise ye the Lord.  \R The Lord’s Name be praised.

\bigskip

\pilcrow{Then shall be said or sung this Psalm following; Except on Easter Day, upon which another Anthem is appointed; and on the nineteenth day of every month it is not to be read here, but in the ordinary course of the Psalms.}

\pilcrow{On the days hereafter named, before this Psalm and after the \emph{Gloria Patri} which follows it, may be sung or said the Invitatory:}

\rubric{On the Sundays in Advent.—}Our King and Saviour draweth nigh \star\ O come, let us adore him.

\rubric{On Christmas Day and until the Epiphany.—}Alleluya.  Unto us a child is born \star\ O come, let us adore him.  Alleluya.

\rubric{On the Epiphany and seven days after, and on the Feast of the Transfiguration.—}The Lord hath manifested forth his glory \star\ O come, let us adore him.

\rubric{On the Purification and the Annunciation.—}The Word was made flesh \star\ O come, let us adore him.

\rubric{On the Monday following the first Sunday after Easter, and until Ascension Day.—}Alleluya. The Lord is risen indeed \star\ O come, let us adore him.  Alleluya.

\rubric{On Ascension Day and until Whitsunday.—}Alleluya.  Christ the Lord ascendeth into heaven \star\ O come, let us adore him.  Alleluya.

\rubric{On Whitsunday and six days after.—}Alleluya. The Spirit of the Lord filleth the world \star\ O come, let us adore him.  Alleluya.

\rubric{On Trinity Sunday.—}One God in Trinity, and Trinity in Unity \star\ O come, let us adore him.

\rubric{On other Festivals for which a proper Epistle and Gospel are appointed.—}The Lord is glorious in his saints \star\ O come, let us adore him.



\subsection{\stylesubsec{}{Venite, exultemus Domino.}{Psalm xcv.}}
\drop{O come, let us sing unto the {\scshape Lord};\ \star\ let us heartily rejoice in the strength of our salvation.}

2\enspace Let us come before his presence with thanksgiving;\ \star\ and shew ourselves glad in him with psalms.

3\enspace For the {\scshape Lord} is a great God;\ \star\ and a great King above all gods.

4\enspace In his hand are all the corners of the earth;\ \star\ and the strength of the hills is his also.

5\enspace  The sea is his, and he made it;\ \star\ and his hands prepared the dry land.

6\enspace  {\red ‡} O come, let us worship and fall down,\ \star\ and kneel before the {\scshape Lord} our Maker.

7\enspace  For he is the Lord our God;\ \star\ and we are the people of his pasture, and the sheep of his hand.

8\enspace  To day if ye will hear his voice, harden not your hearts\ \star\ as in the provocation, and as in the day of temptation in the wilderness;

9\enspace  When your fathers tempted me,\ \star\ proved me, and saw my works.

10\enspace  Forty years long was I grieved with this generation, and said,\ \star\ It is a people that do err in their hearts, for they have not known my ways.

11\enspace  Unto whom I sware in my wrath,\ \star\ that they should not enter into my rest.

Glory be to the Father, and to the Son,\ \star\ and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be,\ \star\ world without end. Amen.

\bigskip
\pilcrow{Then shall follow the Psalms in order as they be appointed. And at the end of every Psalm throughout the year, and likewise at the end of \emph{Benedicite, Benedictus, Magnificat,} and \emph{Nunc dimittis}, shall be repeated,}
Glory be to the Father, and to the Son,\ \star\ and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be,\ \star\ world without end. Amen.

\bigskip
\pilcrow{Then shall be read distinctly with an audible voice the First Lesson, taken out of the Old Testament, as is appointed in the Table of Lessons, except there be proper Lessons assigned for that day: He that readeth so standing and turning himself, as he may best be heard of all such as are present.}

\pilcrow{Note, that before every Lesson the Minister shall say,}
Here beginneth such a Chapter, \rubric{or} Verse of such a Chapter, of such a Book: \rubric{And after every Lesson,} Here endeth the First, \rubric{or} the Second Lesson.

\medskip

\pilcrow{And after that, shall be said or sung, in English, the Hymn called \emph{Te Deum Laudamus}, except from Septuagesima until Easter: during which time shall be used instead \emph{Benedicite, omnia opera}, in English, as followeth:}

\subsection{\stylesubsec{}{Te Deum Laudamus.}{}}
\drop{We praise thee, O God;\ \star\ we acknowledge thee to be the Lord.}

2\enspace All the earth doth worship thee,\ \star\ the Father everlasting.

3\enspace To thee all Angels cry aloud;\ \star\ the Heavens, and all the Powers therein.

4\enspace To thee Cherubin and Seraphin\ \star\ continually do cry,

5\enspace {\red †} Holy, Holy, Holy,\ \star\ Lord God of Sabaoth;

6\enspace Heaven and earth are full of the Majesty\ \star\ of thy glory.

7\enspace The glorious company of the Apostles\ \star\ praise thee.

8\enspace The goodly fellowship of the Prophets\ \star\ praise thee.

9\enspace The noble army of Martyrs\ \star\ praise thee.

10\enspace The holy Church throughout all the world\ \star\ doth acknowledge thee;

11\enspace The Father,\ \star\ of an infinite Majesty;

12\enspace Thine honourable, true,\ \star\ and only Son;

13\enspace Also the Holy Ghost,\ \star\ the Comforter.


\drop{Thou art the King of Glory,\ \star\ O Christ.}

15\enspace Thou art the everlasting Son\ \star\ of the Father.

16\enspace {\red †} When thou tookest upon thee to deliver man,\ \star\ thou didst not abhor the Virgin’s womb.

17\enspace When thou hadst overcome the sharpness of death,\ \star\ thou didst open the Kingdom of Heaven to all believers.

18\enspace Thou sittest at the right hand of God,\ \star\ in the glory of the Father.

19\enspace We believe that thou shalt come\ \star\ to be our Judge.

20\enspace {\red †} We therefore pray thee, help thy servants,\ \star\ whom thou hast redeemed with thy precious blood.

21\enspace Make them to be numbered with thy Saints,\ \star\ in glory everlasting.


\drop{O Lord, save thy people,\ \star\ and bless thine heritage.}\\
23\enspace Govern them,\ \star\ and lift them up for ever.

24\enspace Day by day\ \star\ we magnify thee;

25\enspace And we worship thy Name\ \star\ ever, world without end.

26\enspace Vouchsafe, O Lord,\ \star\ to keep us this day without sin.

27\enspace O Lord, have mercy upon us,\ \star\ have mercy upon us.

28\enspace O Lord, let thy mercy lighten upon us,\ \star\ as our trust is in thee.

29\enspace O Lord, in thee have I trusted;\ \star\ let me never be confounded.

\subsubsection{Or this Canticle,}

\subsection{\stylesubsec{The Song of the Three Children}{Benedicite, omnia opera.}{vv.~35-66.}}

\drop{O all ye Works of the Lord, bless ye the Lord; * praise him, and magnify him for ever.}

2 O ye Angels of the Lord, bless ye the Lord; * O ye Heavens, bless ye the Lord.

3 O ye Waters that be above the firmament, bless ye the Lord; * O all ye Powers of the Lord, bless ye the Lord.

4 O ye Sun and Moon, bless ye the Lord; * O ye Stars of heaven, bless ye the Lord.

5 O ye Showers and Dew, bless ye the Lord; * O ye Winds of God, bless ye the Lord.

6 O ye Fire and Heat, bless ye the Lord; * O ye Winter and Summer, bless ye the Lord.

7 O ye Dews and Frosts, bless ye the Lord; * O ye Frost and Cold, bless ye the Lord.

8 O ye Ice and Snow, bless ye the Lord; * O ye Nights and Days, bless ye the Lord.

9 O ye Light and Darkness, bless ye the Lord; * O ye Lightnings and Clouds, bless ye the Lord.

\drop{O let the Earth bless the Lord; * yea, let it praise him, and magnify him for ever.}

11 O ye Mountains and Hills, bless ye the Lord; * O all ye Green Things upon the earth, bless ye the Lord.

12 O ye Wells, bless ye the Lord; * O ye Seas and Floods, bless ye the Lord.

13 O ye Whales, and all that move in the waters, bless ye the Lord; * O all ye Fowls of the air, bless ye the Lord.

14 O all ye Beasts and Cattle, bless ye the Lord; * O ye Children of Men, bless ye the Lord.

\drop{O let Israel bless the Lord; * praise him, and magnify him for ever.}

16 O ye Priests of the Lord, bless ye the Lord; * O ye Servants of the Lord, bless ye the Lord.

17 O ye Spirits and Souls of the Righteous, bless ye the Lord; * O ye holy and humble Men of heart, bless ye the Lord.

18 O Ananias, Azarias, and Misael, bless ye the Lord; * praise him, and magnify him for ever.

% Composite: 1928 (America), 1929 (scottish), Anglican Breviary, Monastic Diurnal, Sarum Diurnal.
Let us bless the Father, and the Son, and the Holy Ghost; * let us praise him and magnify him for ever.

Blessed art thou, O Lord, in the firmament of heaven; * and to be praised and exalted above all for ever.


\medskip


{\centering\footnotesize{\rubric{On week days, the following Canticle \emph{Benedictus es Domine} may be said or sung instead; or, from Septuagesima until Easter, \emph{Psalm 51}, pg.~\pageref{psalm51}}\par}}


% \subsubsection{Or during Lent, \emph{Psalm 51, pg.~\pageref{psalm51}},}

% \medskip

% \subsubsection{Or this Canticle,}

\subsection{\stylesubsec{The Song of the Three Children}{Benedictus es Domine.}{vv.~29-34}}
\drop{Blessed art thou, O Lord God of our fathers;\ \star\ praised and exalted above all for ever.}

2\enspace Blessed art thou for the Name of thy Majesty;\ \star\ praised and exalted above all for ever.

3\enspace Blessed art thou in the temple of thy holiness;\ \star\ praised and exalted above all for ever.

4\enspace Blessed art thou that beholdest the depths, and dwellest between the Cherubim:\ \star\ praised and exalted above all for ever.

5\enspace Blessed art thou on the glorious throne of thy Kingdom:\ \star\ praised and exalted above all for ever.

6\enspace Blessed art thou in the firmament of heaven:\ \star\ praised and exalted above all for ever.

% American 1928, Scottish 1929
Glory be to the Father, and to the Son,\ \star\ and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be,\ \star\ world without end. Amen.

\subsubsection{}

\pilcrow{Then shall be read in like manner the Second Lesson, taken out of the New Testament. And after that, the Hymn following.}

\subsection{\stylesubsec{The Song of Zachary}{Benedictus.}{St.~Luke i.~68.}}

\drop{Blessed \cross be the Lord God of Israel;\ \star\ for he hath visited and redeemed his people;}

2\enspace And hath raised up a mighty salvation for us,\ \star\ in the house of his servant David;

3\enspace As he spake by the mouth of his holy Prophets,\ \star\ which have been since the world began;

4\enspace That we should be saved from our enemies,\ \star\ and from the hand of all that hate us.

5\enspace To perform the mercy promised to our forefathers,\ \star\ and to remember his holy Covenant;

6\enspace To perform the oath which he sware to our forefather Abraham,\ \star\ that he would give us;

7\enspace That we being delivered out of the hand of our enemies\ \star\ might serve him without fear;

8\enspace In holiness and righteousness before him,\ \star\ all the days of our life.

9\enspace And thou, Child, shalt be called the Prophet of the Highest:\ \star\ for thou shalt go before the face of the Lord to prepare his ways;

10\enspace To give knowledge of salvation unto his people\ \star\ for the remission of their sins,

11\enspace Through the tender mercy of our God;\ \star\ whereby the day-spring from on high hath visited us;

12\enspace To give light to them that sit in darkness, and in the shadow of death,\ \star\ and to guide our feet into the way of peace.

Glory be to the Father, and to the Son,\ \star\ and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be,\ \star\ world without end. Amen.

\bigskip
\pilcrow{Then shall be sung or said the Apostle's Creed, by the Minister and the people standing: Except only such days as the Creed of Saint Athanasius is appointed to be read.}

\drop{I believe in God the Father Almighty, \ Maker of heaven and earth:}

And in Jesus Christ his only Son our Lord: \ Who was conceived by the Holy Ghost, \ Born of the Virgin Mary: \ Suffered under Pontius Pilate, \ Was crucified, dead, and buried: \ He descended into hell; \ The third day he rose again from the dead: \ He ascended into heaven, \ And sitteth on the right hand of God the Father Almighty: \ From thence he shall come to judge the quick and the dead.

I believe in the Holy Ghost: \ The holy Catholick Church; \ The Communion of Saints: \ The Forgiveness of sins: \ The Resurrection of the body, \ And the Life everlasting. Amen.


\bigskip
\pilcrow{And after that, these Prayers following, all devoutly kneeling,}

\centerline{Lord, have mercy upon us.}
\centerline{\emph{Christ, have mercy upon us.}}
\centerline{Lord, have mercy upon us.}

\medskip
{\centering\footnotesize\rubric{Then the Minister, Clerks, and people shall say the Lord's Prayer with a loud voice.}\par}
\ourFather

\smallskip

\centerline{\rubric{Then the Minister standing up shall say,}}

\V O Lord, shew thy mercy upon us.  \R And grant us thy salvation.

\V O Lord, save the \emph{State}.  \R And mercifully hear us when we call upon thee.

\V Endue thy Ministers with righteousness.  \R And make thy chosen people joyful.

\V  O Lord, save thy people.  \R And bless thine inheritance.

\V Give peace in our time, O Lord.  \R Because there is none other that fighteth for us, but only thou, O God.

\V O God, make clean our hearts within us.  \R And take not thy Holy Spirit from us.

\V The Lord be with you.  \R And with thy spirit.

\centerline{Let us pray.}


\pilcrow{Then shall follow three Collects; the first of the day, which shall be the same that is appointed at the Communion; The second for Peace; The third for Grace to live well. And the two last Collects shall never alter, but daily be said at Morning Prayer throughout all the year, as followeth, all kneeling.}

\subsection{\stylesubsec{The second Collect, }{for Peace.}{}}
\drop{O God, who art the author of peace and lover of concord, in knowledge of whom standeth our eternal life, whose service is perfect freedom; Defend us thy humble servants in all assaults of our enemies; that we, surely trusting in thy defence, may not fear the power of any adversaries, through the might of Jesus Christ our Lord. \R Amen.}

\subsection{\stylesubsec{The third Collect, }{for Grace.}{}}
\drop{O Lord, our heavenly Father, Almighty and everlasting God, who hast safely brought us to the beginning of this day; Defend us in the same with thy mighty power; and grant that this day we fall into no sin, neither run into any kind of danger; but that all our doings may be ordered by thy governance, to do always that is righteous in thy sight; through Jesus Christ our Lord. \R Amen.}
\bigskip

\V The Lord be with you.  \R And with thy spirit.

\V Let us bless the Lord.  \R Thanks be to God.
\bigskip

\pilcrow{Here may follow any of the Occasional Prayers and Thanksgivings, as need may require, ending with one of the Conclusions.}

\subsubsection{Here endeth the Order of Morning Prayer throughout the Year.}

\fleuron