\chapter{Quicunque vult} %proofed 12-26-21 vs anglican breviary
\pilcrow{Upon these Feasts; Christmas Day, the Epiphany, Easter Day, Ascension Day, Whitsunday, and upon Trinity Sunday, shall be sung or said at Morning Prayer, instead of the Apostles' Creed, this Confession of our Christian Faith, commonly called the Creed of Athanasius, by the Minister and people standing.}
\pilcrow{\emph{Quicunque vult} may also be sung or said at Mattins upon these Feasts; Saint Matthias, Saint John Baptist, Saint James, Saint Bartholomew, Saint Matthew, Saint Simon and Saint Jude, and Saint Andrew}

\subsubsection{Quicunque vult.}
\drop{Whosoever will be saved,\ \star\ before all things it is necessary that he hold the Catholick Faith.}

2\enspace Which Faith except every one do keep whole and undefiled,\ \star\ without doubt he shall perish everlastingly.

\drop{And the Catholick Faith is this:\ \star\  That we worship one God in Trinity, and Trinity in Unity;}

4\enspace Neither confounding the Persons,\ \star\ nor dividing the Substance.

5\enspace For there is one Person of the Father, another of the Son,\ \star\ and another of the Holy Ghost.

6\enspace But the Godhead of the Father, of the Son, and of the Holy Ghost, is all one;\ \star\ the Glory equal, the Majesty co-eternal.

7\enspace Such as the Father is, such is the Son,\ \star\ and such is the Holy Ghost.

8\enspace The Father uncreate, the Son uncreate,\ \star\ and the Holy Ghost uncreate.

9\enspace The Father incomprehensible, the Son incomprehensible,\ \star\ and the Holy Ghost incomprehensible.

10\enspace The Father eternal, the Son eternal,\ \star\ and the Holy Ghost eternal.

11\enspace And yet they are not three eternals,\ \star\ but one eternal.

12\enspace As also there are not three incomprehensibles, nor three uncreated;\ \star\ but one uncreated, and one incomprehensible.

13\enspace So likewise the Father is Almighty, the Son Almighty,\ \star\ and the Holy Ghost Almighty.

14\enspace And yet they are not three Almighties,\ \star\ but one Almighty.

15\enspace So the Father is God, the Son is God,\ \star\ and the Holy Ghost is God.

16\enspace And yet they are not three Gods,\ \star\ but one God.

17\enspace So likewise the Father is Lord, the Son Lord,\ \star\ and the Holy Ghost Lord.

18\enspace And yet not three Lords,\ \star\ but one Lord.

19\enspace For like as we are compelled by the Christian verity\ \star\ to acknowledge every Person by himself to be both God and Lord;

20\enspace So are we forbidden by the Catholick Religion\ \star\ to say, There be three Gods, or three Lords.

21\enspace The Father is made of none;\ \star\ neither created, nor begotten.

22\enspace The Son is of the Father alone;\ \star\ not made, nor created, but begotten.

23\enspace The Holy Ghost is of the Father and of the Son;\ \star\ neither made, nor created, nor begotten, but proceeding.

24\enspace So there is one Father, not three Fathers; one Son, not three Sons;\ \star\ one Holy Ghost, not three Holy Ghosts.

25\enspace And in this Trinity none is afore, or after other;\ \star\ none is greater, or less than another;

26\enspace But the whole three Persons are co-eternal together,\ \star\ and co-equal.

27\enspace So that in all things, as is aforesaid,\ \star\ the Unity in Trinity and the Trinity in Unity is to be worshipped.

28\enspace He therefore that will be saved\ \star\ must think thus of the Trinity.

\drop{Furthermore, it is necessary to everlasting salvation\ \star\ that he also believe rightly the Incarnation of our Lord Jesus Christ.}

30\enspace For the right Faith is, that we believe and confess\ \star\ that our Lord Jesus Christ, the Son of God, is God and Man;

31\enspace God, of the substance of the Father, begotten before the worlds;\ \star\ and Man of the substance of his Mother, born in the world;

32\enspace Perfect God and perfect Man:\ \star\ of a reasonable soul and human flesh subsisting.

33\enspace Equal to the Father, as touching his Godhead,\ \star\ and inferior to the Father, as touching his manhood;

34\enspace Who, although he be God and Man,\ \star\ yet he is not two, but one Christ;

35\enspace One, not by conversion of the Godhead into flesh,\ \star\ but by taking of the Manhood into God;

36\enspace One altogether; not by confusion of Substance,\ \star\ but by unity of Person.

37\enspace For as the reasonable soul and flesh is one man,\ \star\ so God and Man is one Christ;

38\enspace Who suffered for our salvation:\ \star\ descended into hell, rose again the third day from the dead.

39\enspace He ascended into heaven, he sitteth at the right hand of the Father, God Almighty:\ \star\ from whence he will come to judge the quick and the dead.

40\enspace At whose coming all men will rise again with their bodies,\ \star\ and shall give account for their own works.

41\enspace And they that have done good shall go into life everlasting;\ \star\ and they that have done evil into everlasting fire.

\drop{This is the Catholick Faith,\ \star\ which except a man believe faithfully, he cannot be saved.}

Glory be to the Father, and to the Son,\ \star\ and to the Holy Ghost;

As it was in the beginning, is now, and ever shall be,\ \star\ world without end. Amen.

\fleuron